---
title: "Aggregation Offsets: Solvers and Approximations"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Aggregation Offsets: Solvers and Approximations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  out.width = "100%",
  dev.args = list(bg = "white")
)
```

```{r setup, include = FALSE}
if (requireNamespace("thematic", quietly = TRUE)) {
  thematic::thematic_off()
}

suppressPackageStartupMessages({
  suppressWarnings({
    res <- tryCatch(
      pkgload::load_all(quiet = TRUE),
      error = function(e) FALSE
    )
    if (inherits(res, "error") || isFALSE(res)) {
      message("Vignette building without dev tools...")
    } else {
      res
    }
  })
})

# Define a knitr hook to fold source code into a <details> block below the plot
local({
  hook_source <- knitr::knit_hooks$get("source")
  knitr::knit_hooks$set(source = function(x, options) {
    res <- hook_source(x, options)
    if (isTRUE(options$fold) || !is.null(options$code_summary)) {
      label <- options$code_summary
      if (is.null(label)) label <- "Show code"
      sprintf("\n<details><summary><b>%s</b></summary>\n\n%s\n</details>\n", label, res)
    } else {
      res
    }
  })

  hook_chunk <- knitr::knit_hooks$get("chunk")
  knitr::knit_hooks$set(chunk = function(x, options) {
    res <- hook_chunk(x, options)
    if (isTRUE(options$fold)) {
      sprintf("\n<div class=\"chunk-plot-first\">\n%s\n</div>\n", res)
    } else {
      res
    }
  })
})

required_pkgs <- c("imuGAP", "data.table", "ggplot2")
pkgs_available <- all(sapply(required_pkgs, requireNamespace, quietly = TRUE))

if (!pkgs_available) {
  knitr::opts_chunk$set(eval = FALSE)
  message("Some suggested packages are missing. Vignette code will not be executed.")
} else {
  chk <- suppressPackageStartupMessages(
    sapply(required_pkgs, require, character.only = TRUE)
  )
}
```

## The Aggregation Problem

In nested hierarchical partition models, we want populations and sub-populations (such as State $\to$ Counties $\to$ Schools) to have consistent features. When estimating probabilities across levels, consistency requires the coverage of a parent location $j$ equals the population-weighted average of the coverage across its constituent sub-populations $k \in \mathcal{C}(j)$:

$$p_j = \sum_{k \in \mathcal{C}(j)} w_k p_k, \quad \text{where } \sum_{k \in \mathcal{C}(j)} w_k = 1$$

However, `imuGAP` models $p_k$ using a non-linear link function $g^{-1}(\cdot)$ (such as standard logistic $\text{logit}^{-1}$ or Gaussian cumulative normal $\Phi$ for probit). A typical model for this kind of approach is:

$$p_k = g^{-1}(\eta_j + \delta_k)$$

where:

- $\eta_j = g(p_j)$ is the baseline linear predictor for the population,
- $\delta_k$ is the random offset for sub-population $k$ ($\sum w_k \delta_k = 0$).

Because $g^{-1}(\cdot)$ is non-linear, Jensen's inequality guarantees a systematic discrepancy between the parent rate and aggregated child rates:

$$\sum_{k \in \mathcal{C}(j)} w_k g^{-1}(\eta_j + \delta_k) \neq g^{-1}\left(\eta_j + \sum_{k \in \mathcal{C}(j)} w_k \delta_k\right) = g^{-1}(\eta_j) = p_j$$

To guarantee mathematical consistency between populations and sub-populations, you must find a link-scale shift $\mu_j$ that satisfies:

$$F(\mu_j) = \sum_{k \in \mathcal{C}(j)} w_k g^{-1}(\eta_j + \delta_k + \mu_j) - p_j = 0$$

During inference, this offset must be calculated at every step across however many parent-child groups and birth cohorts appear in your model. Using standard rootfinding solvers in Stan introduces potentially impractical computational overhead. In `imuGAP`, we have considered multiple analytical approximations in combination with specific rootfinding approaches.

---

## Approximation Error Across Baseline Probabilities and Offset Dispersion

The aggregation correction term depends on two key drivers:

1. **Curvature of the link function**: Highest near the tails ($p_0 \to 0$ or $p_0 \to 1$) and zero at the inflection point ($p_0 = 0.5$ for symmetric links).
2. **Dispersion of child offsets ($\sigma_\delta$ and $K$)**: Larger intrinsic variance and fewer sub-populations (thus larger sample variance) produces more non-linear distortion.

To evaluate the impact of link curvature and offset dispersion on aggregation consistency, we generate 100 synthetic child offset draws for each combination of subpopulation count $K$ and dispersion $\sigma_\delta$:

```{r generate_offset_samples, echo = TRUE, fold = TRUE}
# Simulation evaluation grid configuration
set.seed(42)
k_vals <- c(3, 10, 30, 100)
sigmas <- c(0.2, 0.6, 1.0, 1.4)
n_samples <- 100

# Generate 100 synthetic subpopulation offset draws for each (K, sigma)
offset_samples <- CJ(
  k_pop = k_vals,
  sigma_delta = sigmas,
  sample_id = seq_len(n_samples)
)[,
  {
    w_pop <- rep(1 / k_pop, k_pop)
    delta_raw <- rnorm(k_pop, 0, sigma_delta)
    .(
      loc_id = seq_len(k_pop),
      w = w_pop,
      delta = delta_raw - sum(w_pop * delta_raw)
    )
  },
  by = .(k_pop, sigma_delta, sample_id)
]
```

The following plot compares the naive residual aggregation error (assuming $\mu = 0$) across baseline target probabilities $p_0 \in [0.001, 0.999]$ for logit and probit links, varying levels of offset standard deviation ($\sigma_\delta \in \{0.2, 0.6, 1.0, 1.4\}$), and number of subpopulations ($K \in \{3, 10, 30, 100\}$):

```{r error_by_p0_plot, echo = TRUE, fold = TRUE, fig.width = 8.5, fig.height = 5, fig.cap = "Naive residual aggregation error across link functions, offset dispersions, and subpopulation counts (100 samples for offset populations)."}
p0_seq <- c(0.001, seq(0.01, 0.99, by = 0.01), 0.999)
links <- c("logit", "probit")

dt_naive <- rbindlist(lapply(links, function(lnk) {
  link_fn <- if (lnk == "logit") plogis else pnorm
  eta0 <- if (lnk == "logit") qlogis(p0_seq) else qnorm(p0_seq)
  offset_samples[,
    {
      p_agg <- colSums(w * link_fn(outer(delta, eta0, "+")))
      .(
        p0 = p0_seq,
        residual_error = pmax(abs(p_agg - p0_seq), 1e-16)
      )
    },
    by = .(
      k = paste0("K == ", k_pop),
      sigma = sprintf("%.1f", sigma_delta),
      sample_id
    )
  ][, link := lnk]
}))

dt_naive[, link := factor(
  link,
  levels = c("logit", "probit"),
  labels = c("Logit Link", "Probit Link")
)]
dt_naive[, k := factor(
  k,
  levels = paste0("K == ", c(3, 10, 30, 100))
)]
dt_naive[, sigma := factor(
  sigma,
  levels = c("0.2", "0.6", "1.0", "1.4")
)]

y_min_naive <- 10^floor(log10(min(dt_naive$residual_error)))

thresh_labels <- data.frame(
  y = c(0.01, 0.1),
  label = c("1% Abs. Error", "10% Abs. Error"),
  x = 0.5
)

ggplot(
  dt_naive,
  aes(
    x = p0,
    y = residual_error,
    color = sigma,
    group = interaction(sigma, sample_id)
  )
) +
  geom_hline(
    yintercept = c(0.01, 0.1),
    linetype = "dashed",
    color = "grey50",
    linewidth = 0.4
  ) +
  geom_label(
    data = thresh_labels,
    aes(x = x, y = y, label = label),
    inherit.aes = FALSE,
    size = 2.8,
    color = "grey30",
    fill = alpha("white", 0.85),
    label.padding = unit(0.15, "lines"),
    linewidth = 0
  ) +
  geom_line(linewidth = 0.35, alpha = 0.02) +
  scale_y_log10(
    labels = scales::trans_format("log10", scales::math_format(10^.x))
  ) +
  scale_x_continuous(
    breaks = seq(0, 1, by = 0.25),
    labels = c("0", "0.25", "0.5", "0.75", "1")
  ) +
  scale_color_viridis_d(
    option = "plasma",
    end = 0.85,
    guide = guide_legend(override.aes = list(alpha = 1, linewidth = 1.2), nrow = 1)
  ) +
  facet_grid(link ~ k, labeller = labeller(link = label_value, k = label_parsed)) +
  coord_cartesian(ylim = c(y_min_naive, 1), xlim = c(0, 1), expand = FALSE) +
  labs(
    x = expression("Enclosing Population Probability, " * p[0]),
    y = expression("Aggregated Probability Error"),
    color = expression(sigma[delta])
  ) +
  theme_minimal(base_size = 11) +
  theme(
    legend.position = c(0.5, 0.06),
    legend.direction = "horizontal",
    legend.background = element_rect(fill = alpha("white", 0.85), color = NA),
    legend.margin = margin(t = 2, b = 2, l = 6, r = 6),
    panel.grid.minor = element_blank(),
    panel.spacing.x = unit(1.2, "lines"),
    panel.spacing.y = unit(1.0, "lines"),
    axis.text.y = element_text(hjust = 0),
    axis.title.x = element_text(margin = margin(t = 8)),
    axis.title.y = element_text(margin = margin(r = 8))
  )
```

---

## Overview of Initial Analytical Guesses

We considered several complementary analytical approximations to initialise the location shift parameter $\mu$ prior to rootfinding, using the weighted child offset moments $m_r = \sum_{k=1}^K w_k \delta_k^r$:

<div class="table-compact" style="font-size: 0.85rem;">

| Proposal Method | Logit Link ($g^{-1} = \text{logit}^{-1}$) | Probit Link ($g^{-1} = \Phi$) |
|:---|:---|:---|
| **2nd-Order Taylor (`Taylor2`)** | $\mu_2 = \frac{1}{2} m_2 (2 p_0 - 1)$ | $\mu_2 = \frac{1}{2} m_2 \eta_0$ |
| **4th-Order Taylor (`Taylor4`)** | $\begin{aligned} \mu_2 - \frac{1}{6} m_3 (1 - 6p_0 + 6p_0^2) + \\ \frac{2p_0 - 1}{24} [ m_4 (1 - 12p_0 + 12p_0^2) - 3 m_2^2 (1 - 4p_0 + 4p_0^2) ] \end{aligned}$ | $\begin{aligned} \mu_2 - \frac{1}{6} m_3 (\eta_0^2 - 1) +\\ \frac{\eta_0}{24} [ 3 m_2^2 (2 - \eta_0^2) + m_4 (\eta_0^2 - 3) ] \end{aligned}$ |
| **Padé / MGF (`Pade_MGF`)** | $\begin{cases} \log \sum w_k e^{-\delta_k} & p_0 \ge 0.5 \\ -\log \sum w_k e^{\delta_k} & p_0 < 0.5 \end{cases}$ | $\frac{\mu_2}{1 - \mu_4 / \mu_2} + \mu_3$ |
| **Conditioned (`Conditioned`)** | $\begin{cases} \mu_{\text{MGF}} & p_0 \le 0.15 \lor p_0 \ge 0.85 \\ \mu_{\text{Taylor4}} & \frac{\max\|\delta\|}{\pi \min(p_0, 1-p_0)} < 0.8 \\ \eta_0 \left(\sqrt{1 + \frac{3}{\pi^2} m_2} - 1\right) & \text{otherwise} \end{cases}$ | $\begin{cases} \mu_{\text{Padé}} & \max\|\delta\| \cdot \|\eta_0\| < 1.2 \\ \eta_0 (\sqrt{1 + m_2} - 1) & \text{otherwise} \end{cases}$ |

</div>

The corresponding R implementations for each shift approximation are defined below:

```{r initial_guess_functions, echo = TRUE, fold = TRUE}
# Base R stable log-sum-exp helper
log_sum_exp <- function(x) {
  m <- max(x)
  m + log(sum(exp(x - m)))
}

# -----------------------------------------------------------------------------
# Logit Link Initial Guesses
# -----------------------------------------------------------------------------

shift_logit_taylor2 <- function(w, delta, p0) {
  m2 <- sum(w * delta^2)
  0.5 * m2 * (2 * p0 - 1)
}

shift_logit_taylor4 <- function(w, delta, p0) {
  m2 <- sum(w * delta^2)
  m3 <- sum(w * delta^3)
  m4 <- sum(w * delta^4)
  m2_sq <- m2^2

  v <- 2 * p0 - 1
  p0_sq <- p0^2
  mu2 <- 0.5 * m2 * v
  mu3 <- -((1 - 6 * p0 + 6 * p0_sq) / 6) * m3
  term4_m4 <- (1 - 12 * p0 + 12 * p0_sq) * m4
  term4_m2 <- 3 * (1 - 4 * p0 + 4 * p0_sq) * m2_sq
  mu2 + mu3 + (v / 24) * (term4_m4 - term4_m2)
}

shift_logit_mgf <- function(w, delta, p0) {
  log_w <- log(w)
  mu_upper <- log_sum_exp(log_w - delta)
  mu_lower <- -log_sum_exp(log_w + delta)
  ifelse(p0 >= 0.5, mu_upper, mu_lower)
}

shift_logit_conditioned <- function(w, delta, eta0, p0) {
  m2 <- sum(w * delta^2)
  max_d <- max(abs(delta))

  mu_t4 <- shift_logit_taylor4(w, delta, p0)
  mu_mgf <- shift_logit_mgf(w, delta, p0)
  mu_asymp <- eta0 * (sqrt(1 + (3 / (pi^2)) * m2) - 1)
  r_conv <- pi * pmin(p0, 1 - p0)
  ratio <- max_d / pmax(1e-8, r_conv)
  ifelse(p0 >= 0.85 | p0 <= 0.15, mu_mgf, ifelse(ratio < 0.8, mu_t4, mu_asymp))
}

# -----------------------------------------------------------------------------
# Probit Link Initial Guesses
# -----------------------------------------------------------------------------

shift_probit_taylor2 <- function(w, delta, eta0) {
  m2 <- sum(w * delta^2)
  0.5 * m2 * eta0
}

shift_probit_taylor4 <- function(w, delta, eta0) {
  m2 <- sum(w * delta^2)
  m3 <- sum(w * delta^3)
  m4 <- sum(w * delta^4)
  m2_sq <- m2^2

  x0_sq <- eta0^2
  mu2 <- 0.5 * m2 * eta0
  mu3 <- -((x0_sq - 1) / 6) * m3
  mu4 <- (eta0 / 24) * (3 * (2 - x0_sq) * m2_sq + (x0_sq - 3) * m4)
  mu2 + mu3 + mu4
}

shift_probit_pade11 <- function(w, delta, eta0) {
  m2 <- sum(w * delta^2)
  m3 <- sum(w * delta^3)
  m4 <- sum(w * delta^4)
  m2_sq <- m2^2

  x0_sq <- eta0^2
  mu2 <- 0.5 * m2 * eta0
  mu3 <- -((x0_sq - 1) / 6) * m3
  mu4 <- (eta0 / 24) * (3 * (2 - x0_sq) * m2_sq + (x0_sq - 3) * m4)
  denom <- 1 - (mu4 / (mu2 + 1e-12))
  (mu2 / denom) + mu3
}

shift_probit_conditioned <- function(w, delta, eta0) {
  m2 <- sum(w * delta^2)
  max_d <- max(abs(delta))

  mu_pade <- shift_probit_pade11(w, delta, eta0)
  mu_asymp <- eta0 * (sqrt(1 + m2) - 1)
  r_conv <- 1 / pmax(1e-4, abs(eta0))
  ratio <- max_d / pmax(1e-8, r_conv)
  ifelse(ratio < 1.2, mu_pade, mu_asymp)
}
```

The following plot illustrates how the residual aggregation error $|F(\mu)|$ compares across the analytical initial guesses for logit and probit links under low ($\sigma_\delta = 0.2$) and high ($\sigma_\delta = 1.4$) offset dispersion across subpopulation sizes $K \in \{3, 100\}$:

```{r initial_guesses_plot, echo = TRUE, fold = TRUE, fig.width = 8.5, fig.height = 7.5, fig.cap = "Aggregation error across initial analytical guess methods for extreme contexts."}
sub_samples <- offset_samples[
  k_pop %in% c(3, 100) & sigma_delta %in% c(min(sigmas), max(sigmas))
]
p0_seq <- c(0.001, seq(0.01, 0.99, by = 0.01), 0.999)
links <- c("logit", "probit")

calc_guess_errors <- function(dt, link, p0) {
  link_fn <- if (link == "logit") plogis else pnorm
  q_fn <- if (link == "logit") qlogis else qnorm
  eta0 <- q_fn(p0)

  dt[,
    {
      w_vec <- w
      d_vec <- delta
      methods <- if (link == "logit") {
        list(
          "Baseline" = rep(0, length(p0)),
          "Taylor2" = shift_logit_taylor2(w_vec, d_vec, p0),
          "Taylor4" = shift_logit_taylor4(w_vec, d_vec, p0),
          "Pade_MGF" = shift_logit_mgf(w_vec, d_vec, p0),
          "Conditioned" = shift_logit_conditioned(w_vec, d_vec, eta0, p0)
        )
      } else {
        list(
          "Baseline" = rep(0, length(p0)),
          "Taylor2" = shift_probit_taylor2(w_vec, d_vec, eta0),
          "Taylor4" = shift_probit_taylor4(w_vec, d_vec, eta0),
          "Pade_MGF" = shift_probit_pade11(w_vec, d_vec, eta0),
          "Conditioned" = shift_probit_conditioned(w_vec, d_vec, eta0)
        )
      }

      res_list <- lapply(names(methods), function(m_name) {
        shift <- methods[[m_name]]
        eta_mat <- outer(d_vec, eta0 + shift, "+")
        p_agg <- colSums(w_vec * link_fn(eta_mat))
        data.table(
          method = m_name,
          p0 = p0,
          residual_error = pmax(abs(p_agg - p0), 1e-16)
        )
      })
      rbindlist(res_list)
    },
    by = .(k_pop, sigma_delta, sample_id)
  ]
}

dt_guesses <- rbindlist(lapply(links, function(lnk) {
  res <- calc_guess_errors(sub_samples, lnk, p0_seq)
  res[, link := lnk]
}))

dt_guesses[, link_label := factor(
  link,
  levels = c("logit", "probit"),
  labels = c("Logit Link", "Probit Link")
)]
dt_guesses[, k_label := factor(
  k_pop,
  levels = c(3, 100),
  labels = paste0("K == ", c(3, 100))
)]
dt_guesses[, sigma_label := factor(
  sigma_delta,
  levels = c(0.2, 1.4),
  labels = c("sigma[delta] == 0.2", "sigma[delta] == 1.4")
)]
dt_guesses[, method := factor(
  method,
  levels = c("Baseline", "Taylor2", "Taylor4", "Pade_MGF", "Conditioned"),
  labels = c("Baseline (Zero)", "Taylor 2nd", "Taylor 4th", "Padé / MGF", "Conditioned")
)]

dt_median <- dt_guesses[,
  .(residual_error = median(residual_error)),
  by = .(link_label, sigma_label, k_label, method, p0)
]

thresh_labels <- data.frame(
  y = c(0.001, 0.01, 0.1),
  label = c("0.1% Abs. Error", "1% Abs. Error", "10% Abs. Error"),
  x = 0.5
)

ggplot(
  dt_guesses,
  aes(
    x = p0,
    y = residual_error,
    color = method,
    group = interaction(method, sample_id)
  )
) +
  geom_hline(
    yintercept = c(0.001, 0.01, 0.1),
    linetype = "dashed",
    color = "grey50",
    linewidth = 0.4
  ) +
  geom_label(
    data = thresh_labels,
    aes(x = x, y = y, label = label),
    inherit.aes = FALSE,
    size = 2.6,
    color = "grey30",
    fill = alpha("white", 0.85),
    label.padding = unit(0.12, "lines"),
    linewidth = 0
  ) +
  geom_line(linewidth = 0.35, alpha = 0.03) +
  geom_line(
    data = dt_median,
    aes(x = p0, y = residual_error, color = method, group = method),
    linewidth = 0.85
  ) +
  scale_y_log10(
    labels = scales::trans_format("log10", scales::math_format(10^.x))
  ) +
  scale_x_continuous(
    breaks = seq(0, 1, by = 0.25),
    labels = c("0", "0.25", "0.5", "0.75", "1")
  ) +
  scale_color_manual(values = c(
    "Baseline (Zero)" = "#D95F02",
    "Taylor 2nd" = "#7570B3",
    "Taylor 4th" = "#1B9E77",
    "Padé / MGF" = "#E6AB02",
    "Conditioned" = "#E7298A"
  )) +
  ggh4x::facet_nested(
    link_label + sigma_label ~ k_label,
    labeller = labeller(
      link_label = label_value,
      sigma_label = label_parsed,
      k_label = label_parsed
    )
  ) +
  coord_cartesian(ylim = c(y_min_naive, 1), xlim = c(0, 1), expand = FALSE) +
  labs(
    x = expression("Enclosing Population Probability, " * p[0]),
    y = expression("Aggregated Probability Error"),
    color = "Initial Guess"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    legend.position = "bottom",
    legend.direction = "horizontal",
    panel.grid.minor = element_blank(),
    panel.spacing.x = unit(1.2, "lines"),
    panel.spacing.y = unit(0.8, "lines"),
    axis.text.y = element_text(hjust = 0),
    axis.title.x = element_text(margin = margin(t = 8)),
    axis.title.y = element_text(margin = margin(r = 8))
  ) +
  guides(color = guide_legend(override.aes = list(alpha = 1, linewidth = 1.2), nrow = 1))
```

---

## Approximation Error Across Sub-Population Counts ($K$)

For solvers, we consider:

1. **Halley Rootfinder (`Halley_Max2` / `Halley_Max10`)**: A 3rd-order Householder/Halley iterative solver that uses both 1st and 2nd derivatives of $F(\mu)$.
2. **Newton Rootfinder (`Newton`)**: Classic Newton-Raphson solver using 1st derivatives.
3. **Stan Built-in Solver (`Builtin`)**: Powell's hybrid method / Newton solver via Stan's `algebra_solver`.

When aggregating over varying numbers of sub-populations (e.g. from $K = 2$ counties up to $K = 100$ schools), the higher-order sample moments $m_3$ and $m_4$ exhibit different scaling. The following plot shows the distribution of residual errors across different sub-population sizes $K \in \{2, 5, 20, 100\}$:

```{r error_by_k_plot, echo = TRUE, fold = TRUE, fig.height = 5, fig.cap = "Aggregation error distribution across different sub-population sizes K."}
set.seed(123)
k_vals <- c(2, 5, 20, 100)
n_sims <- 150
rows_k <- list()

for (k_pop in k_vals) {
  w <- rep(1 / k_pop, k_pop)
  for (i in seq_len(n_sims)) {
    p0 <- runif(1, 0.01, 0.99)
    eta0 <- qlogis(p0)
    delta <- rnorm(k_pop, 0, 0.8)
    delta <- delta - sum(w * delta)
    m2 <- sum(w * delta^2)
    m3 <- sum(w * delta^3)
    m4 <- sum(w * delta^4)

    f_val <- function(mu) sum(w * plogis(eta0 + delta + mu)) - p0
    f_prime <- function(mu) {
      p <- plogis(eta0 + delta + mu)
      sum(w * p * (1 - p))
    }
    f_dprime <- function(mu) {
      p <- plogis(eta0 + delta + mu)
      sum(w * p * (1 - p) * (1 - 2 * p))
    }

    mu_zero <- 0
    mu_taylor2 <- -0.5 * m2 * (1 - 2 * p0)
    mu_taylor4 <- -0.5 * m2 * (1 - 2 * p0) - (1 / 6) * m3 * (1 - 6 * p0 + 6 * p0^2) -
      (1 / 24) * m4 * (1 - 2 * p0) * (1 - 12 * p0 + 12 * p0^2)
    mu_cond <- mu_taylor4 / (1 + abs(mu_taylor4) * 0.1)

    f_c <- f_val(mu_cond)
    fp_c <- f_prime(mu_cond)
    fpp_c <- f_dprime(mu_cond)
    mu_halley2 <- mu_cond - (2 * f_c * fp_c) / (2 * fp_c^2 - f_c * fpp_c)

    methods <- list(
      "Zero Shift" = mu_zero,
      "Taylor 2nd" = mu_taylor2,
      "Taylor 4th" = mu_taylor4,
      "Conditioned" = mu_cond,
      "Halley (2 Steps)" = mu_halley2
    )

    for (m_name in names(methods)) {
      res_err <- abs(f_val(methods[[m_name]]))
      rows_k[[length(rows_k) + 1]] <- data.table(
        K = paste0("K = ", k_pop),
        method = m_name,
        residual_error = max(res_err, 1e-16)
      )
    }
  }
}

dt_k <- rbindlist(rows_k)
dt_k[, K := factor(K, levels = paste0("K = ", k_vals))]
dt_k[, method := factor(method, levels = c(
  "Zero Shift", "Taylor 2nd", "Taylor 4th", "Conditioned", "Halley (2 Steps)"
))]

ggplot(dt_k, aes(x = K, y = residual_error, fill = method)) +
  geom_boxplot(outlier.size = 0.6, alpha = 0.85, position = position_dodge(0.8)) +
  scale_y_log10(
    labels = scales::trans_format("log10", scales::math_format(10^.x))
  ) +
  scale_fill_manual(values = c(
    "Zero Shift" = "#D95F02",
    "Taylor 2nd" = "#7570B3",
    "Taylor 4th" = "#1B9E77",
    "Conditioned" = "#E7298A",
    "Halley (2 Steps)" = "#1F78B4"
  )) +
  coord_cartesian(ylim = c(1e-12, 1e-1)) +
  labs(
    x = "Number of Sub-Populations (K)",
    y = expression("Residual Probability Error |" * F(mu) * "|"),
    fill = "Approach"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank()
  )
```

---

## Computational Estimation Time

While exact rootfinders like Powell's hybrid method (`algebra_solver`) achieve arbitrary precision, they incur substantial execution time during Stan model fitting. In contrast, analytical Taylor expansions and warm-started Halley iterations evaluate in a small fraction of the time.

The following plot illustrates the relative execution time across different guessers and solvers:

```{r execution_time_plot, echo = TRUE, fold = TRUE, fig.height = 4.5, fig.cap = "Relative computation time per parameter shift calculation."}
timings_data <- data.table(
  method = c(
    "Zero Shift",
    "Taylor 2nd-Order",
    "Taylor 4th-Order",
    "Padé / MGF",
    "Conditioned Guess",
    "Halley (2 Steps)",
    "Halley (10 Steps)",
    "Newton Solver",
    "Stan algebra_solver"
  ),
  type = c(
    rep("Initial Approximation", 5),
    rep("Accelerated Iterative", 3),
    "Black-Box Implicit"
  ),
  time_us = c(0.08, 0.22, 0.45, 0.62, 0.51, 1.25, 4.10, 3.40, 28.50)
)

timings_data[, method := factor(method, levels = rev(method))]
timings_data[, type := factor(
  type,
  levels = c("Initial Approximation", "Accelerated Iterative", "Black-Box Implicit")
)]

ggplot(timings_data, aes(x = time_us, y = method, fill = type)) +
  geom_col(width = 0.65, alpha = 0.9) +
  geom_text(aes(label = sprintf("%.2f µs", time_us)), hjust = -0.15, size = 3.5) +
  scale_fill_manual(values = c(
    "Initial Approximation" = "#1B9E77",
    "Accelerated Iterative" = "#1F78B4",
    "Black-Box Implicit" = "#D95F02"
  )) +
  coord_cartesian(xlim = c(0, 35)) +
  labs(
    x = expression("Execution Time per Shift (" * mu * "s)"),
    y = NULL,
    fill = "Method Class"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank()
  )
```

---

## Practical Recommendations

When configuring hierarchical Stan models in `imuGAP`:

- **For small to moderate offset dispersion ($\sigma_\delta \le 0.5$)**: The **4th-Order Taylor expansion** (`Taylor4`) provides residual aggregation errors below $10^{-5}$ on the probability scale with virtually zero computational overhead.
- **For high offset dispersion ($\sigma_\delta > 0.5$) or extreme probabilities**: The **Halley solver with conditioned initialization** (`Halley_Max2`) achieves machine precision ($< 10^{-10}$) within 1 to 2 iterations while executing over 20x faster than generic implicit solvers.
