---
title: "imuGAP, Immunity: Geographic & Age-based Projection"
output: pdf_document
header-includes:
  - \AtBeginDocument{\let\maketitle\relax}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

```

# `imuGAP`: Immunity: Geographic & Age-based Projection

The package provides a model for estimating vaccination coverage across different geographies and age groups for childhood infectious diseases, such as measles. Predictions from `imuGAP` can be used, for example, to estimate current the population-level vaccination landscape across age groups or impute missing values in observed vaccination coverage data.  

The model represents populations as groups of individuals from a given birth cohort living in a given location. Location structure is hierarchical - for this  example we consider schools nested within counties nested within a state. At the state-level, the proportion of the population who is willing to vaccinate is time varying by birth cohort. For an individual at a certain location, willingness to vaccinate is adjusted by a county offset which captures how that county tends to vary relative to the state, and a school offset which captures how that school tends to vary relative to the county. Those willing to vaccinate experiences a vaccination uptake rate over time, with eligibility determined by a user-specified dose schedule (Figure 1).

\begin{figure}[htbp]
\centering
\includegraphics[width=0.8\textwidth]{../../nc_measles/MeaslesVaxFigure_full_final.pdf}
\caption{imuGAP model schematic}
\end{figure}

`imuGAP` can support observations about any underlying population or group of populations included in the model. In the above example, input data can include:

- [**National Immunization Survey**](https://www.cdc.gov/nis/about/index.html): State-level vaccination coverage survey which informs the state-level trend in willingness to vaccination
- **County-level data**: number of vaccinated individuals in a county from a survey or registry, if available
- **School vaccination data**: School or grade-level reported enrollment and number of students vaccinated

With the above data inputs, in addition to some metadata about the observations and information on location hierarchy, the model can output vaccintaion coverage estimates for the following at any given point in time:

- Age group (at any geographic scale of interest)
- Grade
- School
- County
- State

\newpage

```{r setup, echo = FALSE, include=FALSE, message = FALSE, warning=FALSE}
# List all suggested packages needed for this vignette
required_pkgs <- c("imuGAP", "data.table", "bayesplot", "ggplot2", "dplyr", "knitr", "patchwork")

# Check if they are available
pkgs_available <- all(sapply(required_pkgs, requireNamespace, quietly = TRUE))

# If not available, hide the rest of the document or knit dynamically
if (!pkgs_available) {
  knitr::opts_chunk$set(eval = FALSE)
  message("Some suggested packages are missing. Vignette code will not be executed.")
} else {
  suppressPackageStartupMessages(
    sapply(required_pkgs, require, character.only = TRUE)
  )
}
```

# Technical Details

`imuGAP` provides a [stan](https://mc-stan.org/)-based model for estimating vaccination coverage by location, cohort, and age for childhood infectious diseases, such as measles.

The model represents populations as hierarchical groups, with a life-long propensity for vaccination: some proportion, $\phi$, are unlikely to vaccinate. The population also experiences a vaccination uptake rate, $\lambda$, over time, gated by an eligibility schedule, $\nu(t)$ (an indicator function; 1 when eligible, 0 otherwise). These parameters vary over time and location, according to the user-provided hierarchical structure.

Imagining a particular group defined by a specific location $i$ and cohort $a$ (where $a$ denotes the start of the time period when that group was born). If that cohort is now age $t$, and the vaccine schedule for the first dose is $\nu(t)$, the expected fraction of that group to have at least one dose is then:

$$
P(\ge\textrm{1 dose}) = \left(1 - \phi_{i, a}\right) \left(1 - \exp\left\{-\int_a^{t} \lambda_{i, a}(s)\nu(s) d\textrm{s}\right\}\right)
$$

Which is to say: we model vaccination as a survival process. The package provides a generalized version which handles multiple doses, accounting for conditional dependencies between doses.

This walkthrough demonstrates the workflow of fitting the model and predicting coverage on simulated data bundled with the package. The included demo datasets represent distribution of MMR, a two dose vaccine, across a geographic hierarchy (State -> Counties -> Schools).

# Input Data

The model requires three types of input: (1) location structure/hierarchy, (2) observed coverage counts, and (3) observation metadata.

## (1) Location Hierarchy (`locations_sim`)
The locations dataset defines the nesting relationship of the locations in the model. For each unique location (`loc_id`), the ID of the higher-order spatial entity that this location is contained within (`parent_id`) is given. In this simulation, we have a State, which contains three Counties, which in turn contain various Schools.

```{r locations, echo = FALSE, warning=FALSE}
data("locations_sim", package = "imuGAP")
knitr::kable(head(locations_sim)) %>%
  kableExtra::kable_styling(latex_options = "scale_down")
```

## (2) Coverage Observations (`observations_sim`)
The observations dataset contains the counts of vaccinated individuals (`positive`) out of the total sampled (`sample_n`). The model supports both uncensored and right-censored data (i.e. coverage must be higher than the observed value) via an optional `censored` column. This flexibility covers a common data gap: for example, the number of students up to date on all required vaccines is a censored observation those up to date on the MMR vaccine. 

```{r observations, echo = FALSE, warning=FALSE}
data("observations_sim", package = "imuGAP")
knitr::kable(head(observations_sim[, .(obs_id, loc_id, positive, sample_n, censored)])) %>%
  kableExtra::kable_styling(latex_options = "scale_down")
```

## (3) Observation Metadata (`populations_sim`)
The populations dataset provides observation metadata, mapping each observation ID (`obs_id`) to the corresponding collection of location, birth cohort, age at observation, and vaccine dose, by observation weight.

```{r populations, echo=FALSE, warning=FALSE}
data("populations_sim", package = "imuGAP")
knitr::kable(head(populations_sim)) %>%
  kableExtra::kable_styling(latex_options = "scale_down")
```

The populations dataset provides the capability to link a single observation to the mixture of populations it may represent (i.e. the unique combination of birth cohort, age, and dose). The `weight` value represent how much each individual population contributes to the overall observation and must sum to 1 across all entries for each `ods_id`.   For example, the [Teen National Immunization Survey](https://www.cdc.gov/nis/about/index.html) covers individuals aged 13-17 (i.e. those in life year 14-18). If we assume the each of the underlying cohorts/ages is evenly weighted in the overall population then the weights would be as follows:

```{r, echo = FALSE, warning=FALSE}
knitr::kable(populations_sim |> filter(obs_id == 686)) %>%
  kableExtra::kable_styling(latex_options = "scale_down")
```

In general, observations can correspond to any grouping of cohorts, ages, and locations (at equivalent hierarchical level). The geographic units themselves are arbitrary. We use state/county/school for example purposes, but any nested structure is supported. Additionally arbitrary time resolution is supported (although less commonly used). The following are examples of data that have been or can be used with `imuGAP`:

- School-level vaccination coverage (including grade-specific or whole school observations)
- County-level coverage surveys (not previously used but supported)
- State-level coverage surveys such as the [National Immunization Survey](https://www.cdc.gov/nis/about/index.html)
- A sub-county spatial unit which corresponds to census block group rather than school (e.g. as available for [Florida](https://www.floridahealth.gov/statistics-data/county-immunization-profiles/))

# Data Processing and Validation

Often, data from non-standardized sources (e.g. some school-specific coverage data) need to be cleaned prior to use with `imuGAP`. This package focuses on the inference task, but colleagues have a developmental sibling package, [`tidyschoolvax`](https://github.com/ACCIDDA/tidyschoolvax) focused on data preparation.

However, `imuGAP` does provide a family of `canonicalize_*` functions to validate and convert these raw structures into the canonical forms required by the sampler. You can use those directly to help troubleshoot your inputs, but they are also internally applied in `sampling()`. 

# Fitting the Model

Using the prepared input datasets, we can fit the Bayesian model using `sampling()`. The options for the sampler can be configured using `stan_options()`.

```{r sampling-code, eval = FALSE}
fit_sim <- sampling(
  observations_sim, populations_sim, locations_sim,
  stan_opts = stan_options(
    iter = 1000, chains = 4, refresh = 0, seed = 1L
  )
)
```


```{r load-fit, echo = FALSE}
data("fit_sim", package = "imuGAP")
```

The fit object contains the `stanfit` object emitted from `sampling()`, as well as input settings and data. See the full package vignette on more details on fit diagnostics.   


# Defining a Target for Predictions

To predict vaccine coverage for target populations (i.e. a combination of location, cohort, age, and dose not directly observed), we define a target grid using `create_target()`. 

For example, `imuGAP` can impute the current immune landscape. To do this, we create a "snapshot" prediction target for all locations, including the State and County levels, across ages 1 to 18 at one point in time (i.e. today):

```{r target}
target_sim <- create_target(
  fit = fit_sim,
  location = locations_sim$loc_id, age = 1:18,
  cohort = max(populations_sim$cohort) - 18, dose = c(1, 2),
  mode = "snapshot"
)
head(target_sim)
```

We can then `predict()` to generate predicted coverage probabilities for each target population combination across all posterior draws.

Generating predictions also runs the Stan model (in generated quantities mode):

```{r predict-code, eval = FALSE}
predict_sim <- predict(object = fit_sim, target = target_sim)
```


```{r load-predictions, echo = FALSE}
data("predict_sim", package = "imuGAP")
```

This is an object of class `imugap_predict` which contains a draws array with the MCMC draws for each prediction target as well as the target information. We can summarize these predictions to get the posterior mean and credible intervals for any location, age, or dose:

```{r summarize-predictions}
# Calculate the posterior mean coverage probability for each location and dose at age 5
summary_predict <- summary(predict_sim)

# E.g. current coverage prediction for Chickadee Elementary
summary_chick <- summary_predict |>
  subset(
    loc_id == "Chickadee Elementary" & dose == 2 & age %in% 5:11
  ) |>
  dplyr::select(loc_id, age, cohort, dose, mean, q2_5, q50, q97_5)

knitr::kable(summary_chick) %>%
  kableExtra::kable_styling(latex_options = "scale_down")
```

Now let's visualize the results. First we will take a look at overall state coverage by cohort (A). Note that the lower coverage among 5 year olds is due to them only having been eligible for their second dose for one year. The trend in coverage by age at the county level (B) follows the same trend as the state but with differing magnitude. Finally, we can look at the distribution of coverage estimates by age group for an exemplar county (C). 

```{r plts, echo=FALSE, fig.height=8, fig.width=6}
state_plt <- summary_predict |>
  subset(loc_id == "State" & dose == 2 & age > 4) |>
  ggplot() +
  geom_line(aes(x = age, y = q50)) +
  geom_ribbon(aes(x = age, ymin = q2_5, ymax = q97_5), alpha = 0.5) +
  theme_bw() +
  scale_x_continuous(breaks = 5:18) +
  labs(x = "Age", y = "Coverage", title = "A. State-level two dose coverage")

cnty_plt <- summary_predict |>
  subset(loc_id %in% c("Scruggs", "Simone", "Watson") & dose == 2 & age > 4) |>
  ggplot() +
  geom_line(aes(x = age, y = q50, group = loc_id, color = loc_id)) +
  geom_ribbon(aes(x = age, ymin = q2_5, ymax = q97_5, fill = loc_id), alpha = 0.2) +
  theme_bw() +
  scale_x_continuous(breaks = 5:18) +
  theme(legend.position = "bottom") +
  labs(
    x = "Age", y = "Coverage", title = "B. County-level two dose coverage",
    color = "County", fill = "County"
  )

sch_plt <- summary_predict |>
  subset(
    !(loc_id %in% c("State", "Scruggs", "Simone", "Watson")) &
      loc_id %in% locations_sim$loc_id[locations_sim$parent_id == "Scruggs"][-1] &
      dose == 2 &
      age > 4
  ) |>
  ggplot() +
  geom_boxplot(aes(x = factor(age), y = q50)) +
  theme_bw() +
  labs(
    x = "Age", y = "Coverage",
    title = "C. Distribution of School-Level Coverage For Scruggs County"
  )

(state_plt / cnty_plt / sch_plt)
```

\newpage

# Model Limitations

This model is subject to several limitations. 

- It does not account for movement of students in and out of locations throughout their life course. 
- Does not capture temporal variation in vaccination uptake rate (e.g due to temporary disruptions like COVID-19) 


