---
title: "imuGAP, Immunity: Geographic & Age-based Projection"
output: pdf_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, echo = FALSE, message = FALSE, warning=FALSE}
# List all suggested packages needed for this vignette
required_pkgs <- c("imuGAP", "data.table", "bayesplot", "ggplot2", "dplyr", "knitr", "cowplot")

# Check if they are available
pkgs_available <- all(sapply(required_pkgs, requireNamespace, quietly = TRUE))

# If not available, hide the rest of the document or knit dynamically
if (!pkgs_available) {
  knitr::opts_chunk$set(eval = FALSE)
  message("Some suggested packages are missing. Vignette code will not be executed.")
} else {
  library(imuGAP)
  library(data.table)
  library(bayesplot)
  library(ggplot2)
  library(dplyr)
  library(knitr)
  library(cowplot)
}
```

# Introduction

The name `imuGAP` stands for "Immunity: Geographic & Age-based Projection". The package provides a [stan](https://mc-stan.org/)-based model for estimating vaccination coverage by location, cohort, and age for childhood infectious diseases, such as measles.

The core model represents a target population as having a life-long propensity for vaccination; some proportion, $\phi$, of that population is unlikely to vaccinate and the complementary proportion, $1 - \phi$, is likely to vaccinate. That population then experiences a vaccination uptake rate, $\lambda$, over the model time eras, according to the vaccination eligibility schedule, $\nu$. These core parameters can vary over time and location, in a user-specifiable way.

Focusing just on the core model element, imagine a particular population location $i$ and cohort $a$ (where $a$ denotes the start of the time period when that group was born). If that cohort is now age $t$, and the vaccine schedule for the first dose is $\nu(t)$, the expected fraction of that group to have at least one dose is then:

$$
P(\ge\textrm{1 dose}) = \left(1 - \phi_{i, a}\right) \left(1 - \exp\left\{-\int_a^{t} \lambda_{i, a}(s)\nu(s) d\textrm{s}\right\}\right)
$$

Which is to say, we are representing vaccination coverage via a survival-like model. We have implemented a generalized version of this model in `imuGAP` which allows for a vaccine with more than one dose, accounting for the fact that being eligible to receive any given dose is conditional on having received the prior required doses. In the following example, we apply the model to simulated observations of coverage for a two-dose vaccine such as MMR.  

# Walkthrough of Basic Usage

This walkthrough demonstrates the workflow of fitting the model and predicting coverage on simulated data. The package includes seven bundled datasets for demonstration, representing a nested geographic setup (State -> Counties -> Schools). 

## Input Data

The package requires three types of input data: 1) information of location structure/hierarchy, 2) observed coverage counts, and 3) observations metadata. 

### 1. Location Hierarchy (`locations_sim`)
The locations dataset defines the nesting relationship of the locations in the model. For each unique location (`loc_id`), the ID of the higher-order spatial entity that this location is contained within (`parent_id`) is given. In this simulation, we have a State, which contains three Counties, which in turn contain various Schools.

```{r locations, echo = FALSE, warning=FALSE}
data("locations_sim", package = "imuGAP")
knitr::kable(head(locations_sim)) %>%
  kableExtra::kable_styling(latex_options = "scale_down")

```

### 2. Coverage Observations (`observations_sim`)
The observations dataset contains the counts of individuals who were vaccinated (`positive`) out of the total sampled (`sample_n`) for each observation. It also includes a `censored` column, which is `1` if the observation is right-censored and `NA` otherwise. For example, the number of students up to date on all required vaccines must be less than or equal to the number of students up to date on the MMR vaccine and is therefore a censored observation of MMR coverage. 

```{r observations, echo = FALSE, warning=FALSE}
data("observations_sim", package = "imuGAP")
knitr::kable(head(observations_sim[, .(obs_id, loc_id, positive, sample_n, censored)])) %>%
  kableExtra::kable_styling(latex_options = "scale_down")

```

### 3. Observation Metadata (`populations_sim`)
The populations dataset acts as observation metadata, mapping each observation ID (`obs_id`) to the corresponding location, birth cohort, age at observation, vaccine dose, and observation weight.

```{r populations, echo=FALSE, warning=FALSE}
data("populations_sim", package = "imuGAP")

knitr::kable(head(populations_sim)) %>%
  kableExtra::kable_styling(latex_options = "scale_down")

```

Observations may have multiple entries in the populations dataset as an observation correspond to multiple individual populations (defined as a unique combination of birth cohort, age, and dose). The `weight` value represent how much each individual population contributes to the overall observation and must sum to 1 across all entries for each `ods_id`.   For example, consider an observation such as the [Teen National Immunization Survey](https://www.cdc.gov/nis/about/index.html) which covers teens ages 13-17 (i.e. those in life year 14-18). If we assume the each of the underlying cohorts/ages is evenly weighted in the overall population then the weights would be as follows:

```{r, echo = FALSE, warning=FALSE}

knitr::kable(populations_sim |> filter(obs_id == 686)) %>%
  kableExtra::kable_styling(latex_options = "scale_down")

```

### Supported Data

Observations can correspond to any grouping of cohorts, ages, and locations. Furthermore, the geographic units included are arbitrary. We use state/county/school for example purposes, but any nested spatial structure is supported. Additionally arbitrary time resolution is supported (although less commonly used). The following are examples of data that have been or can be used with `imuGAP`:

- School-level vaccination coverage (including grade-specific or whole school observations)
- County-level coverage surveys (not previously used but supported)
- State-level coverage surveys such as the [National Immunization Survey](https://www.cdc.gov/nis/about/index.html)
- A sub-county spatial unit which corresponds to census block group rather than school (e.g. as available for [Florida](https://www.floridahealth.gov/statistics-data/county-immunization-profiles/))

## Data Processing and Valibation

Often, data from non-standardized sources (e.g. some school-specific coverage data) need to be cleaned prior to use with `imuGAP`. While this functionality is not supported by `imuGAP` directly, colleagues at ACCIDDA are working on a pipeline to standardize and combine school-level vaccination data for the United States (`USImmunityProfiles`, not yet public).

The `imuGAP` package provides a family of `canonicalize_*` functions to validate, clean, and convert these raw structures into the canonical forms required by the sampler. You can use those directly to help troubleshoot your inputs. However, as shown in the next section, the `sampling()` method also automatically canonicalizes the inputs. Please see the full package vignette for examples of canonicalization and common errors. 

## Fitting the Model

Using the prepared input datasets, we can fit the Bayesian model using `sampling()`. The options for the sampler can be configured using `stan_options()`.

```{r sampling-code, eval = FALSE}
fit_sim <- sampling(
  observations_sim,
  populations_sim,
  locations_sim,
  stan_opts = stan_options(
    iter = 1000,
    chains = 4,
    refresh = 0,
    seed = 1L
  )
)
```


```{r load-fit, echo = FALSE}
data("fit_sim", package = "imuGAP")
```

The fit object contains the `stanfit` object emitted from `sampling()`, as well as input settings and data. See the full package vignette on more details on fit diagnostics.   


## Defining a Target for Predictions

To predict vaccine coverage for a target population (which can include populations without direct observations, as long as they exist in the locations hierarchy), we first define a target grid using `create_target()`. 

For example, one use case for imuGAP is to help understand the current immune landscape. To do this, we can obtain a "snapshot" prediction target for all locations, including the State and County levels, across ages 1 to 18 at one point in time (i.e. today):

```{r target}
target_sim <- create_target(
  fit = fit_sim,
  location = locations_sim$loc_id,
  age = 1:18,
  cohort = max(populations_sim$cohort) - 18,
  dose = c(1, 2),
  mode = "snapshot"
)
head(target_sim)
```


## Predicting Coverage

Finally, we run `predict()` to generate predicted coverage probabilities for each target population combination across all posterior draws.

Generating predictions also runs the Stan model (in generated quantities mode):

```{r predict-code, eval = FALSE}
predict_sim <- predict(object = fit_sim, target = target_sim)
```


```{r load-predictions, echo = FALSE}
data("predict_sim", package = "imuGAP")
```

This is an object of class `imugap_predict` which contains a draws array with the MCMC draws for each prediction target as well as the target information. We can summarize these predictions to get the posterior mean and credible intervals for any location, age, or dose:

```{r summarize-predictions}
# Calculate the posterior mean coverage probability for each location and dose at age 5
summary_predict <- summary(predict_sim)

# E.g. current coverage prediction for Chickadee Elementary
summary_chick <- summary_predict |>
  filter(loc_id == "Chickadee Elementary",
         dose == 2,
         age %in% 5:11) |>
  dplyr::select(loc_id, age, cohort, dose, mean, q2_5, q50, q97_5)

knitr::kable(summary_chick) %>%
  kableExtra::kable_styling(latex_options = "scale_down")


```

\newpage

Now let's visualize the results. First we will take a look at overall state coverage by cohort (A). Note that the lower coverage among 5 year olds is due to them only having been eligible for their second dose for one year. The trend in coverage by age at the county level (B) follows the same trend as the state but with differing magnitude. Finally, we can look at the distribution of coverage estimates by age group for an exemplar county (C). 

```{r plts, echo=FALSE, fig.height=8, fig.width=6}
state_plt <- summary_predict |>
  subset(loc_id == "State" & dose == 2 & age > 4) |>
  ggplot() +
  geom_line(aes(x = age, y = q50)) +
  geom_ribbon(aes(x = age, ymin = q2_5, ymax = q97_5), alpha = 0.5) +
  theme_bw() +
  scale_x_continuous(breaks = 5:18) +
  labs(x = "Age", y = "Coverage", title = "A. State-level two dose coverage")

cnty_plt <- summary_predict |>
  subset(loc_id %in% c("Scruggs", "Simone", "Watson") & dose == 2 & age > 4) |>
  ggplot() +
  geom_line(aes(x = age, y = q50, group = loc_id, color = loc_id)) +
  geom_ribbon(aes(x = age, ymin = q2_5, ymax = q97_5, fill = loc_id), alpha = 0.2) +
  theme_bw() +
  scale_x_continuous(breaks = 5:18) +
  theme(legend.position = "bottom") +
  labs(
    x = "Age", y = "Coverage", title = "B. County-level two dose coverage",
    color = "County", fill = "County"
  )

sch_plt <- summary_predict |>
  subset(
    !(loc_id %in% c("State", "Scruggs", "Simone", "Watson")) &
      loc_id %in% locations_sim$loc_id[locations_sim$parent_id == "Scruggs"][-1] &
      dose == 2 &
      age > 4
  ) |>
  ggplot() +
  geom_boxplot(aes(x = factor(age), y = q50)) +
  theme_bw() +
  labs(
    x = "Age", y = "Coverage",
    title = "C. Distribution of School-Level Coverage For Scruggs County"
  )

plot_grid(state_plt, cnty_plt, sch_plt, nrow = 3)
```

# Model Limitations

This model is subject to several limitations. 

- It does not account for movement of students in and out of locations throughout their life course. 
- Does not capture temporal variation in vaccination uptake rate (e.g due to temporary disruptions like COVID-19) 


