---
title: "Fit Inspection and Stan Diagnostics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Fit Inspection and Stan Diagnostics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  out.width = "100%",
  dev.args = list(bg = "white")
)
```

```{r setup, echo = FALSE, message = FALSE, warning = FALSE}
if (requireNamespace("thematic", quietly = TRUE)) {
  thematic::thematic_off()
}

# Define a knitr hook to fold source code into a <details> block below the plot
local({
  hook_source <- knitr::knit_hooks$get("source")
  knitr::knit_hooks$set(source = function(x, options) {
    res <- hook_source(x, options)
    if (isTRUE(options$fold) || !is.null(options$code_summary)) {
      label <- options$code_summary
      if (is.null(label)) label <- "Show code"
      sprintf("\n<details><summary><b>%s</b></summary>\n\n%s\n</details>\n", label, res)
    } else {
      res
    }
  })

  hook_chunk <- knitr::knit_hooks$get("chunk")
  knitr::knit_hooks$set(chunk = function(x, options) {
    res <- hook_chunk(x, options)
    if (isTRUE(options$fold)) {
      sprintf("\n<div class=\"chunk-plot-first\">\n%s\n</div>\n", res)
    } else {
      res
    }
  })
})

required_pkgs <- c("imuGAP", "data.table", "bayesplot", "ggplot2")
pkgs_available <- all(sapply(required_pkgs, requireNamespace, quietly = TRUE))

if (!pkgs_available) {
  knitr::opts_chunk$set(eval = FALSE)
  message("Some suggested packages are missing. Vignette code will not be executed.")
} else {
  chk <- suppressPackageStartupMessages(
    sapply(required_pkgs, require, character.only = TRUE)
  )

  # Always enforce white background and black text on plots
  ggplot2::theme_set(
    ggplot2::theme_bw() +
      ggplot2::theme(
        plot.background = ggplot2::element_rect(fill = "white", colour = NA),
        panel.background = ggplot2::element_rect(fill = "white", colour = NA),
        legend.background = ggplot2::element_rect(fill = "white", colour = NA),
        legend.box.background = ggplot2::element_blank(),
        text = ggplot2::element_text(colour = "black"),
        axis.text = ggplot2::element_text(colour = "black"),
        axis.title = ggplot2::element_text(colour = "black"),
        plot.title = ggplot2::element_text(colour = "black"),
        plot.subtitle = ggplot2::element_text(colour = "black"),
        strip.text = ggplot2::element_text(colour = "black"),
        strip.background = ggplot2::element_rect(fill = "grey90", colour = "grey20")
      )
  )
}
```

# Introduction

Because `imuGAP` estimates an underlying process model for vaccination (including baseline uptake
splines $\beta_{\text{bs}}$, force-of-vaccination rates $\lambda$, and hierarchical location
offsets $\delta$) via Bayesian inference in Stan, evaluating a model fit requires assessing whether
MCMC chains converged without pathologies and successfully recovered these latent process
parameters.

This vignette demonstrates how to:

1. Load and extract posterior parameter draws from an `imugap_fit` object using `extract_imugap()`.
2. Inspect parameter trace plots across multiple chains.
3. Compare posterior parameter estimates against true data-generating simulation values (`latent_params_sim`).
4. Evaluate spatial variance components ($\sigma$) and location-specific random offsets ($\delta$).

---

# 1. Loading the Model Fit and Extracting Parameters

You can load the bundled `fit_sim` object, which wraps the underlying Stan model fit along with dataset and options metadata:

```{r load-fit}
data("fit_sim", package = "imuGAP")
data("latent_params_sim", package = "imuGAP")
data("locations_sim", package = "imuGAP")
data("observations_sim", package = "imuGAP")

fit_sim
```

To extract posterior draws for specific model parameters, use `extract_imugap()`:

```{r extract-params}
# Extract B-spline basis coefficients
beta_draws <- extract_imugap(fit_sim, pars = "beta_bs")
str(beta_draws)

# Extract unconstrained force of vaccination rate parameters
lambda_draws <- extract_imugap(fit_sim, pars = "lambda_raw")
str(lambda_draws)
```

---

# 2. Visual Diagnostics with `bayesplot` and Parameter Recovery

### State-Level Cohort Uptake: Basis Spline Coefficients ($\beta_{\text{bs}}$)

The state-level baseline propensity curve is parameterized as a B-spline over birth cohorts with coefficients $\beta_{\text{bs}}$. The following trace plots show parameter draws across all 4 MCMC chains alongside within-chain medians and the true simulation parameters (dashed red lines):

```{r trace-plot-beta, fold = TRUE, code_summary = "Show plot code"}
beta_pars <- paste0("beta_bs[", seq_along(latent_params_sim$beta_bs), "]")
beta_arr <- as.array(fit_sim$stanfit, pars = beta_pars)
beta_chain_meds <- rbindlist(lapply(beta_pars, function(p) {
  data.table(
    parameter = p,
    Chain = factor(seq_len(dim(beta_arr)[2])),
    med_val = apply(beta_arr[, , p, drop = FALSE], 2, median)
  )
}))

beta_ref <- data.frame(
  parameter = beta_pars,
  true_val = latent_params_sim$beta_bs,
  label = sprintf(
    "True~beta[%d] == %.2f",
    seq_along(latent_params_sim$beta_bs),
    latent_params_sim$beta_bs
  )
)

bayesplot::mcmc_trace(
  beta_arr,
  facet_args = list(labeller = ggplot2::as_labeller(function(x) {
    gsub("beta_bs\\[(\\d+)\\]", "beta[\\1]", x)
  }, default = ggplot2::label_parsed))
) +
  geom_hline(
    data = beta_chain_meds,
    aes(yintercept = med_val, color = Chain),
    linetype = "solid", linewidth = 0.5, alpha = 0.8
  ) +
  geom_hline(
    data = beta_ref,
    aes(yintercept = true_val),
    color = "firebrick", linetype = "dashed", linewidth = 0.8
  ) +
  geom_label(
    data = beta_ref,
    aes(x = 100, y = true_val, label = label),
    parse = TRUE, color = "firebrick", fill = ggplot2::alpha("white", 0.75),
    linewidth = NA, vjust = -0.3, hjust = 0, size = 3.2
  ) +
  theme(legend.position = "none")
```

---

### Hierarchy Layer Variances ($\sigma_{\text{layer}}$)

In multi-layer models, `sigma_layer` captures the standard deviation of location random offsets at each level of the tree (e.g. $\sigma_{\text{county}}$ at layer 1 and $\sigma_{\text{school}}$ at layer 2):

```{r trace-plot-sigmas, fold = TRUE, code_summary = "Show plot code"}
sigma_pars <- c("sigma_layer[1]", "sigma_layer[2]")
sigma_arr <- as.array(fit_sim$stanfit, pars = sigma_pars)
sigma_chain_meds <- rbindlist(lapply(sigma_pars, function(p) {
  data.table(
    parameter = p,
    Chain = factor(seq_len(dim(sigma_arr)[2])),
    med_val = apply(sigma_arr[, , p, drop = FALSE], 2, median)
  )
}))

sigma_ref <- data.frame(
  parameter = sigma_pars,
  true_val = c(latent_params_sim$sigma_cnty, latent_params_sim$sigma_sch),
  label = sprintf(
    "True~sigma == %.2f",
    c(latent_params_sim$sigma_cnty, latent_params_sim$sigma_sch)
  )
)

bayesplot::mcmc_trace(
  sigma_arr,
  facet_args = list(labeller = ggplot2::as_labeller(c(
    "sigma_layer[1]" = "sigma[County]",
    "sigma_layer[2]" = "sigma[School]"
  ), default = ggplot2::label_parsed))
) +
  geom_hline(
    data = sigma_chain_meds,
    aes(yintercept = med_val, color = Chain),
    linetype = "solid", linewidth = 0.5, alpha = 0.8
  ) +
  geom_hline(
    data = sigma_ref,
    aes(yintercept = true_val),
    color = "firebrick", linetype = "dashed", linewidth = 0.8
  ) +
  geom_label(
    data = sigma_ref,
    aes(x = 100, y = true_val, label = label),
    parse = TRUE, color = "firebrick", fill = ggplot2::alpha("white", 0.75),
    linewidth = NA, vjust = -0.3, hjust = 0, size = 3.2
  ) +
  coord_cartesian(ylim = c(0, 2.5)) +
  theme(legend.position = "none")
```

---

### County Location Offsets ($\delta_{\text{county}}$)

Location offsets represent deviation in vaccination propensity from the parent region. The following trace plots show offset trajectories for Scruggs, Simone, and Watson counties.

Indicator arrows show:

- **Red arrows (left edge, iteration 0)**: Expected error direction based on observational sampling noise from finite sample draws ($\Delta_{\delta} = -\overline{\Delta\text{logit}}_{\text{cov}}$).
- **Blue arrows (right edge, iteration 500)**: Realized difference between the posterior median estimate and the true latent offset.

```{r trace-plot-county-offsets, fold = TRUE, code_summary = "Show plot code"}
get_weighted_qr_basis <- function(w) {
  n_w <- length(w)
  v1 <- w / sqrt(sum(w^2))
  mat_m <- matrix(0, nrow = n_w, ncol = n_w)
  mat_m[, 1] <- v1
  for (j in seq_len(n_w - 1L)) {
    for (i in seq_len(n_w)) {
      mat_m[i, j + 1L] <- if (i == j) 1.0 else 0.0
    }
  }
  q_star <- qr.Q(qr(mat_m))[, 2:n_w, drop = FALSE]
  for (j in seq_len(ncol(q_star))) {
    nz <- which(abs(q_star[, j]) > 1e-10)[1]
    if (!is.na(nz) && q_star[nz, j] < 0) {
      q_star[, j] <- -q_star[, j]
    }
  }
  q_star
}

loc_info <- canonicalize_locations(locations_sim)
ld <- imuGAP:::assemble_layer_data(loc_info)

bounds_to_range <- function(starts, total) {
  rbind(starts, c(tail(starts, -1L) - 1L, total))
}

layer_bounds <- bounds_to_range(ld$layer_starts, ld$n_locs)
parent_child_bounds <- bounds_to_range(ld$parent_child_starts, ld$n_locs)

loc_layer_idx <- integer(ld$n_locs - 1L)
for (k in seq_len(ld$n_layers - 1L)) {
  st <- layer_bounds[1, k + 1L] - 1L
  en <- layer_bounds[2, k + 1L] - 1L
  loc_layer_idx[st:en] <- k
}

loc_pop_scale <- numeric(ld$n_locs - 1L)
for (k in seq_len(ld$n_layers - 1L)) {
  st <- layer_bounds[1, k + 1L]
  en <- layer_bounds[2, k + 1L]
  layer_pop <- ld$loc_population[st:en]
  mean_layer_pop <- mean(layer_pop)
  loc_pop_scale[(st - 1L):(en - 1L)] <- sqrt(mean_layer_pop / layer_pop)
}

n_unconstrained <- (ld$n_locs - 1L) - ld$n_parent_locs
qr_basis <- matrix(0, nrow = ld$n_locs - 1L, ncol = n_unconstrained)
col_offset <- 0L
for (p in seq_len(ld$n_parent_locs)) {
  st <- parent_child_bounds[1, p]
  en <- parent_child_bounds[2, p]
  n_child <- en - st + 1L
  pop_slice <- ld$loc_population[st:en]
  w <- pop_slice / sum(pop_slice)
  w_prime <- sqrt(w)
  q_star <- get_weighted_qr_basis(w_prime)
  qr_basis[(st - 1L):(en - 1L), (col_offset + 1L):(col_offset + n_child - 1L)] <- q_star
  col_offset <- col_offset + (n_child - 1L)
}

z_arr <- as.array(fit_sim$stanfit, pars = "z_layer")
sigma_arr <- as.array(fit_sim$stanfit, pars = "sigma_layer")
n_iter <- dim(z_arr)[1]
n_chains <- dim(z_arr)[2]

off_layer_arr <- array(0, dim = c(n_iter, n_chains, ld$n_locs - 1L))
for (iter in seq_len(n_iter)) {
  for (chain in seq_len(n_chains)) {
    z_vec <- z_arr[iter, chain, ]
    sigma_vec <- sigma_arr[iter, chain, ]
    off_vec <- as.vector(((qr_basis %*% z_vec) * loc_pop_scale) * sigma_vec[loc_layer_idx])
    off_layer_arr[iter, chain, ] <- off_vec
  }
}
dimnames(off_layer_arr) <- list(
  iterations = NULL,
  chains = paste0("chain:", seq_len(n_chains)),
  parameters = paste0("off_layer[", seq_len(ld$n_locs - 1L), "]")
)

county_names <- names(latent_params_sim$off_cnty)
non_root_locs <- loc_info$loc_id[-1]
county_pars <- paste0("off_layer[", match(county_names, non_root_locs), "]")
county_arr <- off_layer_arr[, , county_pars, drop = FALSE]
county_chain_meds <- rbindlist(lapply(county_pars, function(p) {
  data.table(
    parameter = p,
    Chain = factor(seq_len(dim(county_arr)[2])),
    med_val = apply(county_arr[, , p, drop = FALSE], 2, median)
  )
}))

obs <- copy(observations_sim)
phi_st <- latent_params_sim$phi_state
cov <- latent_params_sim$uptake
off_cnty <- latent_params_sim$off_cnty
censor_red <- latent_params_sim$censor_reduction

obs_cnty <- obs[loc_id %in% county_names]
obs_cnty[, mu := {
  c_off <- unname(off_cnty[loc_id])
  phi_c <- plogis(qlogis(phi_st[cohort]) + c_off)
  (1 - phi_c) * cov[11, 2] * censor_red
}, by = loc_id]
obs_cnty[, p_adj := (positive + 0.5) / (sample_n + 1.0)]
obs_cnty[, delta_cov := qlogis(p_adj) - qlogis(mu)]

cnty_errs <- obs_cnty[, .(expected_delta_err = -mean(delta_cov)), by = .(loc_id)]
county_post_meds <- apply(county_arr, 3, median)

county_ref <- data.frame(
  parameter = county_pars,
  loc_id = county_names,
  true_val = unname(latent_params_sim$off_cnty[county_names]),
  post_med = unname(county_post_meds[county_pars]),
  label = sprintf(
    "True~delta == %.2f",
    latent_params_sim$off_cnty[county_names]
  )
)
county_ref <- merge(county_ref, cnty_errs, by = "loc_id")

county_ref$red_arrow_x <- 0
county_ref$red_arrow_yend <- county_ref$true_val + county_ref$expected_delta_err
county_ref$blue_arrow_x <- 500
county_ref$blue_arrow_yend <- county_ref$post_med

bayesplot::mcmc_trace(
  county_arr,
  facet_args = list(
    scales = "fixed",
    labeller = ggplot2::as_labeller(setNames(county_names, county_pars))
  )
) +
  geom_hline(
    data = county_chain_meds,
    aes(yintercept = med_val, color = Chain),
    linetype = "solid", linewidth = 0.5, alpha = 0.8
  ) +
  geom_hline(
    data = county_ref,
    aes(yintercept = true_val),
    color = "firebrick", linetype = "dashed", linewidth = 0.8
  ) +
  geom_segment(
    data = county_ref,
    aes(x = red_arrow_x, xend = red_arrow_x, y = true_val, yend = red_arrow_yend),
    arrow = arrow(length = unit(0.04, "inches"), type = "closed"),
    color = "firebrick", linewidth = 0.8
  ) +
  geom_segment(
    data = county_ref,
    aes(x = blue_arrow_x, xend = blue_arrow_x, y = true_val, yend = blue_arrow_yend),
    arrow = arrow(length = unit(0.04, "inches"), type = "closed"),
    color = "royalblue", linewidth = 0.8
  ) +
  geom_label(
    data = county_ref,
    aes(x = 50, y = true_val, label = label),
    parse = TRUE, color = "firebrick", fill = ggplot2::alpha("white", 0.75),
    linewidth = NA, vjust = -0.3, hjust = 0, size = 3.2
  ) +
  theme(legend.position = "none")
```

---

### Force of Vaccination Rate ($\lambda$)

The unconstrained rate parameters $\lambda_{\text{raw}}$ govern the speed at which eligible individuals receive vaccine doses once reaching age eligibility. The following trace plots show posterior draws on the exponentiated rate scale:

```{r trace-plot-lambdas, fold = TRUE, code_summary = "Show plot code"}
lambda_pars <- c("lambda_raw[1]", "lambda_raw[2]")
lambda_arr <- as.array(fit_sim$stanfit, pars = lambda_pars)
lambda_chain_meds <- rbindlist(lapply(lambda_pars, function(p) {
  data.table(
    parameter = p,
    Chain = factor(seq_len(dim(lambda_arr)[2])),
    med_val = apply(lambda_arr[, , p, drop = FALSE], 2, median)
  )
}))

lambda_ref <- data.frame(
  parameter = lambda_pars,
  true_val = log(latent_params_sim$lambda),
  label = sprintf("True~lambda == %.1f", latent_params_sim$lambda)
)

bayesplot::mcmc_trace(
  lambda_arr,
  facet_args = list(labeller = ggplot2::as_labeller(c(
    "lambda_raw[1]" = "lambda[1]~(Dose~1)",
    "lambda_raw[2]" = "lambda[2]~(Dose~2)"
  ), default = ggplot2::label_parsed))
) +
  geom_hline(
    data = lambda_chain_meds,
    aes(yintercept = med_val, color = Chain),
    linetype = "solid", linewidth = 0.5, alpha = 0.8
  ) +
  geom_hline(
    data = lambda_ref,
    aes(yintercept = true_val),
    color = "firebrick", linetype = "dashed", linewidth = 0.8
  ) +
  geom_label(
    data = lambda_ref,
    aes(x = 100, y = true_val, label = label),
    parse = TRUE, color = "firebrick", fill = ggplot2::alpha("white", 0.75),
    linewidth = NA, vjust = -0.3, hjust = 0, size = 3.2
  ) +
  coord_cartesian(ylim = c(0.5, 1.5)) +
  scale_y_continuous(
    transform = "exp",
    labels = function(x) sprintf("%.2f", exp(x))
  ) +
  labs(y = "Uptake Rate (exponential scale)") +
  theme(legend.position = "none")
```

---

# Summary and Related Vignettes

- For an overview of estimating the underlying process model and predicting coverage, see **[Getting Started with imuGAP](imuGAP.html)**.
- To explore input data preparation and surveillance streams, see **[Included Example Datasets and Data Validation](example_data.html)**.
- To learn how `imuGAP` estimates parameters across arbitrary spatial hierarchy resolutions, see **[Flexible Location Layers in imuGAP](user_specified_layers.html)**.
