---
title: "Included Example Datasets and Data Validation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Included Example Datasets and Data Validation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  out.width = "100%",
  dev.args = list(bg = "white")
)
```

```{r setup, include = FALSE}
if (requireNamespace("thematic", quietly = TRUE)) {
  thematic::thematic_off()
}

# Define a knitr hook to fold source code into a <details> block below the plot
local({
  hook_source <- knitr::knit_hooks$get("source")
  knitr::knit_hooks$set(source = function(x, options) {
    res <- hook_source(x, options)
    if (isTRUE(options$fold) || !is.null(options$code_summary)) {
      label <- options$code_summary
      if (is.null(label)) label <- "Show code"
      sprintf("\n<details><summary><b>%s</b></summary>\n\n%s\n</details>\n", label, res)
    } else {
      res
    }
  })

  hook_chunk <- knitr::knit_hooks$get("chunk")
  knitr::knit_hooks$set(chunk = function(x, options) {
    res <- hook_chunk(x, options)
    if (isTRUE(options$fold)) {
      sprintf("\n<div class=\"chunk-plot-first\">\n%s\n</div>\n", res)
    } else {
      res
    }
  })
})

required_pkgs <- c("imuGAP", "data.table", "ggplot2")
pkgs_available <- all(sapply(required_pkgs, requireNamespace, quietly = TRUE))

if (!pkgs_available) {
  knitr::opts_chunk$set(eval = FALSE)
  message("Some suggested packages are missing. Vignette code will not be executed.")
} else {
  chk <- suppressPackageStartupMessages(
    sapply(required_pkgs, require, character.only = TRUE)
  )

  # Always enforce white background and black text on plots
  ggplot2::theme_set(
    ggplot2::theme_bw() +
      ggplot2::theme(
        plot.background = ggplot2::element_rect(fill = "white", colour = NA),
        panel.background = ggplot2::element_rect(fill = "white", colour = NA),
        legend.background = ggplot2::element_rect(fill = "white", colour = NA),
        legend.box.background = ggplot2::element_blank(),
        text = ggplot2::element_text(colour = "black"),
        axis.text = ggplot2::element_text(colour = "black"),
        axis.title = ggplot2::element_text(colour = "black"),
        plot.title = ggplot2::element_text(colour = "black"),
        plot.subtitle = ggplot2::element_text(colour = "black"),
        strip.text = ggplot2::element_text(colour = "black"),
        strip.background = ggplot2::element_rect(fill = "grey90", colour = "grey20")
      )
  )

  data("locations_sim", package = "imuGAP")
  data("observations_sim", package = "imuGAP")
  data("populations_sim", package = "imuGAP")
  data("latent_params_sim", package = "imuGAP")

  county_names <- sort(unique(locations_sim$loc_id[
    locations_sim$parent_id == "State" & !is.na(locations_sim$parent_id)
  ]))
  sch_counts <- table(locations_sim$parent_id[locations_sim$parent_id %in% county_names])
  n_schools <- sum(sch_counts)
  sch_summary <- paste0(sch_counts[county_names], " in ", county_names, collapse = ", ")
  dose_sched <- latent_params_sim$dose_schedule %||% c(1L, 4L)
}
```

# Overview of the Synthetic Simulation Universe

`imuGAP` provides four bundled datasets (`locations_sim`, `observations_sim`, `populations_sim`,
and `target_sim`) alongside latent parameter fixtures (`latent_params_sim`, `fit_sim`,
`predict_sim`) that reflect an underlying vaccination process model and surveillance streams
across a multi-resolution hierarchical population structure.

In this scenario:

- A single **State** is partitioned into `r length(county_names)` **Counties** (`r paste(county_names, collapse = ", ")`), which contain a total of `r n_schools` elementary **Schools** (`r sch_summary`).
- The vaccine requires a `r length(dose_sched)`-dose sequential regimen (e.g. MMR, with dose 1 recommended at age `r dose_sched[1]` and dose 2 eligible at age `r dose_sched[2]`).
- Observations are gathered from diverse empirical surveillance streams:
  - **ChildVaxView / NIS-Child**: State-level single-cohort point estimates at ages 2 and 3 for dose 1.
  - **SchoolVaxView**: State-level kindergarten entry surveys (dose 2 at age 5).
  - **TeenVaxView**: State-level multi-cohort cross-sectional survey snapshots spanning ages 14 to 18 (dose 2).
  - **County Sixth-Grade Surveys**: Right-censored dose 2 coverage at age 11 across all `r length(county_names)` counties.
  - **School Kindergarten Entry Records**: Annual kindergarten entry surveys for dose 2 at age 5 across all `r n_schools` individual schools.

---

# 1. Inspecting the Bundled Datasets

### Location Hierarchy (`locations_sim`)

The `locations_sim` dataset defines the spatial tree structure. Every location node must have either 0 offspring (a leaf node, such as a school) or strictly greater than 1 offspring ($\ge 2$ children).

```{r locations}
data("locations_sim", package = "imuGAP")
head(locations_sim)
nrow(locations_sim)
```

You can validate and canonicalize the location tree using `canonicalize_locations()`:

```{r locations-canon}
canonical_locations <- canonicalize_locations(locations_sim)
head(canonical_locations)
```

`canonicalize_locations()` introduces:

- `loc_c_id` and `loc_cp_id`: 1-based integer continuous indices for the location and its parent.
- `layer`: 1-based integer tree depth (1 for State, 2 for County, 3 for School).
- `layer_bound`: Index slicing offsets used internally by Stan to partition the spatial random effect arrays.

---

### Coverage Observations (`observations_sim`)

The `observations_sim` dataset contains the binomial survey outcomes:

- `obs_id`: Unique identifier for each observation.
- `loc_id`: Location identifier matching a node in `locations_sim`.
- `positive`: Count of vaccinated individuals.
- `sample_n`: Total sample size ($N$).
- `censored`: `1` for right-censored observations, `NA` for uncensored observations.

```{r observations}
data("observations_sim", package = "imuGAP")
head(observations_sim[, .(obs_id, loc_id, positive, sample_n, censored)])
```

Canonicalizing observations with `canonicalize_observations()` assigns continuous 1-based indices `obs_c_id`, sorting uncensored records before censored records for efficient block execution in Stan:

```{r observations-canon}
canonical_observations <- canonicalize_observations(observations_sim)
head(canonical_observations)
```

---

### Observation Metadata & Populations (`populations_sim`)

The `populations` table provides essential metadata for all observations—defining the discrete
location, age, cohort, and dose associated with each survey outcome. Every record in `observations`
receives one or more entries in `populations`:

- **Direct (1:1) observations**: A single row with `weight = 1.0` maps the observation directly
  to a discrete population slice (such as a single-cohort kindergarten survey at age 5).
- **Mixed observations**: When a survey mixes multiple underlying populations (such as
  cross-sectional surveys spanning multiple age groups or cohorts), `populations` captures this
  mixture using multiple rows sharing the same `obs_id`, each assigned a fractional weight
  (e.g. equal weights or demographic population shares summing to 1.0).

```{r populations}
data("populations_sim", package = "imuGAP")
head(populations_sim)
```

For single-cohort kindergarten surveys, `obs_id` has a 1:1 mapping with `weight = 1.0`. For
multi-age surveys (such as TeenVaxView at `obs_id == 761`), multiple rows describe the mixture
across ages:

```{r populations-multi}
# TeenVaxView-style observation spanning ages 14 to 18
observations_sim[
  obs_id == 761,
  .(obs_id, loc_id, positive, sample_n, age_min, age_max, dose)
]

# Corresponding population metadata with distributed weights summing to 1.0
populations_sim[obs_id == 761]
```

You can validate and canonicalize populations using `canonicalize_populations()`:

```{r populations-canon}
canonical_populations <- canonicalize_populations(
  populations_sim, observations_sim, locations_sim
)
head(canonical_populations)
```

---

# 2. Canonicalization Rules and Validation Diagnostics

`imuGAP` enforces strict structural assertions to safeguard model fitting against corrupted or incompatible inputs:

### Observation Binomial Count Consistency

If an observation has `positive > sample_n`, `canonicalize_observations()` raises an informative error:

```{r validation-failure-obs}
invalid_obs <- copy(observations_sim[, .(obs_id, loc_id, positive, sample_n, censored)])
invalid_obs[1, positive := sample_n + 10]

tryCatch(
  canonicalize_observations(invalid_obs),
  error = function(e) message("Caught expected error: ", e$message)
)
```

### Hierarchy Invariant Enforcement

If duplicate location IDs exist, or single-child chains are introduced, `canonicalize_locations()` rejects the tree:

```{r validation-failure-locs}
invalid_locs <- rbind(
  locations_sim,
  data.frame(loc_id = "Scruggs", parent_id = "State"),
  fill = TRUE
)

tryCatch(
  canonicalize_locations(invalid_locs),
  error = function(e) message("Caught expected error: ", e$message)
)
```

---

# 3. Visualizing the Synthetic Observation Streams

The following figures illustrate empirical coverage rates across each observation stream compared against underlying true data-generating parameters (`latent_params_sim`).

### State-Level Observation Streams

```{r synthetic-state-viz, fold = TRUE, code_summary = "Show plot code"}
data("latent_params_sim", package = "imuGAP")

state_obs <- copy(observations_sim[loc_id == "State"])
state_obs[, source := factor(
  fcase(
    dose == 1 & age_min == 2, "ChildVaxView (Dose 1, Age 2)",
    dose == 1 & age_min == 3, "ChildVaxView (Dose 1, Age 3)",
    age_min == 5, "SchoolVaxView (Dose 2, Age 5)",
    default = "TeenVaxView (Dose 2, Ages 14-18)"
  ),
  levels = c(
    "ChildVaxView (Dose 1, Age 2)",
    "ChildVaxView (Dose 1, Age 3)",
    "SchoolVaxView (Dose 2, Age 5)",
    "TeenVaxView (Dose 2, Ages 14-18)"
  )
)]
state_obs[, obs_prop := positive / sample_n]

single_cohort_obs <- state_obs[is.na(age_max) | age_max == age_min + 1L]
multi_cohort_obs <- copy(state_obs[!is.na(age_max) & age_max > age_min + 1L])
multi_cohort_obs[, cohort_max := cohort_min + (age_max - 1L) - age_min]

latent_state <- data.table(
  cohort_min = seq_along(latent_params_sim$phi_state),
  propensity = 1 - latent_params_sim$phi_state
)

n_c <- length(latent_params_sim$phi_state)
skv_max_cohort <- max(state_obs[age_min == 5, cohort_min])
tvv_max_cohort <- max(multi_cohort_obs$cohort_min)

latent_curves <- rbindlist(list(
  data.table(
    cohort_min = seq_len(n_c),
    latent_cov = (1 - latent_params_sim$phi_state) *
      latent_params_sim$uptake[2, 1],
    source = "ChildVaxView (Dose 1, Age 2)"
  ),
  data.table(
    cohort_min = seq_len(n_c),
    latent_cov = (1 - latent_params_sim$phi_state) *
      latent_params_sim$uptake[3, 1],
    source = "ChildVaxView (Dose 1, Age 3)"
  ),
  data.table(
    cohort_min = seq_len(skv_max_cohort),
    latent_cov = (1 - latent_params_sim$phi_state[seq_len(skv_max_cohort)]) *
      latent_params_sim$uptake[5, 2],
    source = "SchoolVaxView (Dose 2, Age 5)"
  ),
  data.table(
    cohort_min = seq_len(tvv_max_cohort),
    latent_cov = (1 - latent_params_sim$phi_state[seq_len(tvv_max_cohort)]) *
      mean(latent_params_sim$uptake[14:18, 2]),
    source = "TeenVaxView (Dose 2, Ages 14-18)"
  )
))
latent_curves[, source := factor(source, levels = levels(state_obs$source))]

ggplot() +
  geom_line(
    data = latent_state,
    aes(x = cohort_min, y = propensity, linetype = "True Lifetime Uptake Propensity (1 - phi)"),
    color = "gray40", linewidth = 0.8, alpha = 0.5
  ) +
  geom_line(
    data = latent_curves,
    aes(x = cohort_min, y = latent_cov, color = source),
    linetype = "dashed", linewidth = 0.7, alpha = 0.4
  ) +
  geom_segment(
    data = multi_cohort_obs,
    aes(x = cohort_min, xend = cohort_max, y = obs_prop, yend = obs_prop, color = source),
    linewidth = 1.1, alpha = 0.95
  ) +
  geom_point(
    data = single_cohort_obs,
    aes(x = cohort_min, y = obs_prop, color = source, shape = source),
    size = 2.4, alpha = 0.95
  ) +
  scale_x_continuous(breaks = seq(0, 30, by = 5), minor_breaks = seq(1, 30, by = 1)) +
  coord_cartesian(xlim = c(0, 30), ylim = c(0.4, 1.0)) +
  scale_linetype_manual(
    name = NULL,
    values = c("True Lifetime Uptake Propensity (1 - phi)" = "dashed")
  ) +
  scale_color_brewer(name = "Data Source", palette = "Dark2") +
  scale_shape_manual(
    name = "Data Source",
    values = c(
      "ChildVaxView (Dose 1, Age 2)" = 16,
      "ChildVaxView (Dose 1, Age 3)" = 17,
      "SchoolVaxView (Dose 2, Age 5)" = 15,
      "TeenVaxView (Dose 2, Ages 14-18)" = 18
    )
  ) +
  guides(
    color = guide_legend(
      override.aes = list(
        shape = c(16, 17, 15, NA),
        linetype = c("blank", "blank", "blank", "solid"),
        linewidth = c(0, 0, 0, 1.1),
        alpha = 1
      )
    ),
    shape = "none"
  ) +
  theme(
    legend.position = "inside",
    legend.position.inside = c(0.98, 0.02),
    legend.justification.inside = c(1, 0)
  ) +
  labs(x = "Birth Cohort Index", y = "Vaccination Proportion")
```

---

### County-Level Right-Censored Surveillance

`imuGAP` supports observations at intermediate spatial resolutions as well as right-censored surveillance streams.

In public health surveillance, records often report composite vaccine series completion (for example, whether a student has received all required middle school entry immunizations) rather than tracking a specific vaccine in isolation. Because an individual who completed the full multi-vaccine series has necessarily received the target vaccine, the observed series completion rate acts as a lower bound on the target vaccine's true coverage (true coverage $\ge$ observed proportion).

`imuGAP` natively models right-censored observations:

- In `observations`, you mark right-censored survey records with `censored = 1` (and uncensored records with `NA`).
- During Bayesian model fitting, the likelihood accounts for the right-censoring mechanism, ensuring that lower-bound surveillance records accurately inform the latent process without biasing coverage trajectories downward.

In this simulation universe, routine county-wide middle school entry tallies (dose 2 at age 11) across all three counties represent right-censored surveillance streams:

```{r synthetic-county-viz, fold = TRUE, code_summary = "Show plot code"}
county_names <- names(latent_params_sim$off_cnty)
county_obs <- copy(observations_sim[loc_id %in% county_names])
county_obs[, obs_prop := positive / sample_n]

max_obs_cohort <- max(county_obs$cohort_min)
n_cohorts <- length(latent_params_sim$phi_state)

county_latent <- rbindlist(lapply(names(latent_params_sim$off_cnty), function(cnty) {
  cohorts <- seq_len(n_cohorts)
  c_idx <- match(cnty, names(latent_params_sim$off_cnty))
  offset <- latent_params_sim$off_cnty[c_idx]
  phi_shifted <- plogis(qlogis(latent_params_sim$phi_state[cohorts]) + offset)
  cov_true <- (1 - phi_shifted) * latent_params_sim$uptake[11, 2]
  cov_censored <- cov_true * latent_params_sim$censor_reduction
  data.table(
    loc_id = cnty,
    cohort_min = cohorts,
    latent_cov = cov_true,
    latent_cov_censored = cov_censored
  )
}))

county_latent_obs <- county_latent[cohort_min <= max_obs_cohort]
county_latent_unobs <- county_latent[cohort_min >= max_obs_cohort]

ggplot() +
  geom_point(
    data = county_obs,
    aes(x = cohort_min, y = obs_prop),
    color = "steelblue", size = 2, alpha = 0.85
  ) +
  geom_line(
    data = county_latent_obs,
    aes(
      x = cohort_min,
      y = latent_cov,
      color = "True Latent Coverage",
      linetype = "Observed Cohorts"
    ),
    linewidth = 0.9
  ) +
  geom_line(
    data = county_latent_obs,
    aes(
      x = cohort_min,
      y = latent_cov_censored,
      color = "Censored Latent (0.95x)",
      linetype = "Observed Cohorts"
    ),
    linewidth = 0.9
  ) +
  geom_line(
    data = county_latent_unobs,
    aes(
      x = cohort_min,
      y = latent_cov,
      color = "True Latent Coverage",
      linetype = "Unobserved Cohorts"
    ),
    linewidth = 0.9
  ) +
  geom_line(
    data = county_latent_unobs,
    aes(
      x = cohort_min,
      y = latent_cov_censored,
      color = "Censored Latent (0.95x)",
      linetype = "Unobserved Cohorts"
    ),
    linewidth = 0.9
  ) +
  facet_wrap(~loc_id) +
  scale_x_continuous(breaks = seq(0, 30, by = 5), minor_breaks = seq(1, 30, by = 1)) +
  scale_y_continuous(expand = c(0, 0)) +
  coord_cartesian(xlim = c(0, 30), ylim = c(0.6, 1.0)) +
  scale_color_manual(
    name = NULL,
    values = c("True Latent Coverage" = "firebrick", "Censored Latent (0.95x)" = "darkorange")
  ) +
  scale_linetype_manual(
    name = NULL,
    values = c("Observed Cohorts" = "dashed", "Unobserved Cohorts" = "dotted")
  ) +
  guides(
    color = guide_legend(order = 1, override.aes = list(linewidth = 0.9, linetype = "solid")),
    linetype = guide_legend(order = 2, override.aes = list(linewidth = 0.9, color = "black"))
  ) +
  theme(
    legend.position = "inside",
    legend.position.inside = c(0.85, 0.12),
    legend.justification.inside = c(1, 0)
  ) +
  labs(x = "Birth Cohort Index", y = "6th Grade Survey Coverage (Dose 2, Age 11)")
```

---

### School Kindergarten Enrollment Records

The following plot illustrates school-level kindergarten entry records (dose 2 at age 5) across
all elementary schools grouped by county, comparing empirical survey counts against the
underlying school-specific latent coverage curves:

```{r synthetic-school-viz, fold = TRUE, code_summary = "Show plot code"}
canonical_locations <- canonicalize_locations(locations_sim)
sch_locs <- canonical_locations[layer == 3]

n_cohorts <- length(latent_params_sim$phi_state)
off_c <- latent_params_sim$off_cnty
off_s <- latent_params_sim$off_sch
uptake_val <- latent_params_sim$uptake[5, 2]

school_latent <- rbindlist(lapply(seq_len(nrow(sch_locs)), function(i) {
  sch <- sch_locs$loc_id[i]
  cnty <- sch_locs$parent_id[i]
  tot_off <- unname(off_c[cnty]) + unname(off_s[sch])
  cohorts <- seq_len(n_cohorts)
  phi_shifted <- plogis(qlogis(latent_params_sim$phi_state[cohorts]) + tot_off)
  cov_true <- (1 - phi_shifted) * uptake_val
  data.table(loc_id = sch, county_id = cnty, cohort_min = cohorts, latent_cov = cov_true)
}))

sch_obs <- copy(observations_sim[loc_id %in% sch_locs$loc_id])
sch_obs[, obs_prop := positive / sample_n]
sch_obs <- merge(sch_obs, sch_locs[, .(loc_id, county_id = parent_id)], by = "loc_id")

max_obs_cohort <- max(sch_obs$cohort_min)
school_latent_obs <- school_latent[cohort_min <= max_obs_cohort]
school_latent_unobs <- school_latent[cohort_min >= max_obs_cohort]

ggplot() +
  geom_point(
    data = sch_obs,
    aes(x = cohort_min, y = obs_prop, color = loc_id),
    size = 1.3, alpha = 0.65
  ) +
  geom_line(
    data = school_latent_obs,
    aes(
      x = cohort_min,
      y = latent_cov,
      group = loc_id,
      color = loc_id,
      linetype = "Observed Cohorts"
    ),
    linewidth = 0.75, alpha = 0.85
  ) +
  geom_line(
    data = school_latent_unobs,
    aes(
      x = cohort_min,
      y = latent_cov,
      group = loc_id,
      color = loc_id,
      linetype = "Unobserved Cohorts"
    ),
    linewidth = 0.75, alpha = 0.85
  ) +
  facet_wrap(~county_id) +
  scale_x_continuous(breaks = seq(0, 30, by = 5), minor_breaks = seq(1, 30, by = 1)) +
  coord_cartesian(xlim = c(0, 30), ylim = c(0.6, 1.0)) +
  scale_linetype_manual(
    name = NULL,
    values = c("Observed Cohorts" = "solid", "Unobserved Cohorts" = "dotted")
  ) +
  theme(legend.position = "none") +
  labs(
    x = "Birth Cohort Index",
    y = "Kindergarten Entry Coverage (Dose 2, Age 5)"
  )
```

---

# Summary and Related Vignettes

- For an overview of estimating the underlying process model and predicting coverage, see **[Getting Started with imuGAP](imuGAP.html)**.
- To learn how `imuGAP` configures and estimates parameters across arbitrary spatial depths, see **[Flexible Location Layers in imuGAP](user_specified_layers.html)**.
- For detailed MCMC convergence checks, trace plots, and parameter inference, see **[Fit Inspection and Stan Diagnostics](examining_fits.html)**.
