---
title: "Getting Started with imuGAP"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with imuGAP}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  out.width = "100%",
  dev.args = list(bg = "white")
)
```

```{r setup, echo = FALSE, message = FALSE, warning = FALSE}
if (requireNamespace("thematic", quietly = TRUE)) {
  thematic::thematic_off()
}

# Define a knitr hook to fold source code into a <details> block below the plot
local({
  hook_source <- knitr::knit_hooks$get("source")
  knitr::knit_hooks$set(source = function(x, options) {
    res <- hook_source(x, options)
    if (isTRUE(options$fold) || !is.null(options$code_summary)) {
      label <- options$code_summary
      if (is.null(label)) label <- "Show code"
      sprintf("\n<details><summary><b>%s</b></summary>\n\n%s\n</details>\n", label, res)
    } else {
      res
    }
  })

  hook_chunk <- knitr::knit_hooks$get("chunk")
  knitr::knit_hooks$set(chunk = function(x, options) {
    res <- hook_chunk(x, options)
    if (isTRUE(options$fold)) {
      sprintf("\n<div class=\"chunk-plot-first\">\n%s\n</div>\n", res)
    } else {
      res
    }
  })
})

required_pkgs <- c("imuGAP", "data.table", "ggplot2", "tidyr", "dplyr")
pkgs_available <- all(sapply(required_pkgs, requireNamespace, quietly = TRUE))

if (!pkgs_available) {
  knitr::opts_chunk$set(eval = FALSE)
  message("Some suggested packages are missing. Vignette code will not be executed.")
} else {
  chk <- suppressPackageStartupMessages(
    sapply(required_pkgs, require, character.only = TRUE)
  )

  # Always enforce white background and black text on plots
  ggplot2::theme_set(
    ggplot2::theme_bw() +
      ggplot2::theme(
        plot.background = ggplot2::element_rect(fill = "white", colour = NA),
        panel.background = ggplot2::element_rect(fill = "white", colour = NA),
        legend.background = ggplot2::element_rect(fill = "white", colour = NA),
        legend.box.background = ggplot2::element_blank(),
        text = ggplot2::element_text(colour = "black"),
        axis.text = ggplot2::element_text(colour = "black"),
        axis.title = ggplot2::element_text(colour = "black"),
        plot.title = ggplot2::element_text(colour = "black"),
        plot.subtitle = ggplot2::element_text(colour = "black"),
        strip.text = ggplot2::element_text(colour = "black"),
        strip.background = ggplot2::element_rect(fill = "grey90", colour = "grey20")
      )
  )
}
```

# Introduction

`imuGAP` (Immunity: Geographic & Age-based Projection) estimates and predicts vaccination coverage on arbitrary hierarchical location partitions using Bayesian Stan models.

The package enables researchers and public health analysts to synthesize across multiple heterogeneous surveillance streams (such as national surveys, state school entry records, and local coverage censuses) to:

1. **Estimate current vaccination coverage** across arbitrary sub-populations and geographic levels.
2. **Project lifetime cohort trajectories** (e.g. coverage at each age from birth through adolescence for a given birth cohort).
3. **Impute missing data gaps** across unobserved years, age groups, or sub-regions by borrowing strength hierarchically.

### Core Model Intuition

The core `imuGAP` model represents each target population slice $(i, a)$ (location $i$, birth cohort $a$) as having an underlying non-uptake rate $\phi_{i,a}$, with complementary uptake propensity $1 - \phi_{i,a}$. Given a dose eligibility schedule $\nu(s)$ and a force-of-vaccination rate $\lambda(s)$, cumulative coverage at age $t$ for the first dose is modeled as:

$$
P(\ge\textrm{1 dose}) = \left(1 - \phi_{i, a}\right) \left(1 - \exp\left\{-\int_a^{t} \lambda_{i, a}(s)\nu(s) d\textrm{s}\right\}\right)
$$

This formulation extends sequentially to multiple vaccine doses, conditional on receipt of prior doses.

---

# The Fit & Predict Loop

This walkthrough presents the end-to-end workflow: **Load Data $\rightarrow$ Configure & Fit $\rightarrow$ Define Target Grid $\rightarrow$ Predict $\rightarrow$ Summarize & Plot**.

---

## 1. Preparing the Inputs

`imuGAP` requires three core input tables:
1. `locations`: The nested spatial tree hierarchy (e.g. State $\rightarrow$ County $\rightarrow$ School).
2. `observations`: Empirical survey binomial counts (`positive` and `sample_n`) and censoring indicators.
3. `populations`: Metadata and fractional weights mapping observations to discrete age/cohort/dose slices.

```{r load-data}
library(imuGAP)
library(data.table)
library(ggplot2)

data("locations_sim", package = "imuGAP")
data("observations_sim", package = "imuGAP")
data("populations_sim", package = "imuGAP")

head(locations_sim)
head(observations_sim[, .(obs_id, loc_id, positive, sample_n, censored)])
head(populations_sim)
```

Input datasets are automatically validated and canonicalized during sampling via `canonicalize_locations()`, `canonicalize_observations()`, and `canonicalize_populations()`.

---

## 2. Fitting the Model

Model fitting is executed via `sampling()`. Model options (such as B-spline degrees of freedom and dose eligibility ages) are configured with `imugap_options()`, and MCMC sampler settings (chains, iterations, backend) are specified with `stan_options()`:

```{r sampling-code, eval = FALSE}
# Configure model options
im_opts <- imugap_options(df = 5L, dose_schedule = c(1L, 4L))

# Configure Stan MCMC sampler
st_opts <- stan_options(chains = 4, iter = 2000, seed = 1L)

# Fit the model
fit <- sampling(
  observations = observations_sim,
  populations = populations_sim,
  locations = locations_sim,
  imugap_opts = im_opts,
  stan_opts = st_opts
)
```

For this vignette, we load the pre-computed fit fixture `fit_sim` bundled with the package:

```{r load-fit}
data("fit_sim", package = "imuGAP")
fit_sim
```

---

## 3. Defining a Prediction Target Grid

To generate coverage predictions for specific populations, define a target grid using `create_target()`.

For example, to evaluate a cross-sectional "snapshot" across all locations for ages 1 through 18 and doses 1 and 2:

```{r target}
target_grid <- create_target(
  location = unique(locations_sim$loc_id),
  age = 1:18,
  cohort = max(populations_sim$cohort) - 18,
  dose = c(1, 2),
  mode = "snapshot"
)
head(target_grid)
```

---

## 4. Predicting Coverage

Pass the model fit and target grid to `predict()` to generate posterior draws of coverage:

```{r predict-code, eval = FALSE}
# Generate posterior draws for each target slice
predict_res <- predict(object = fit_sim, target = target_grid, posterior_size = 100)
```

We load the pre-computed prediction fixture `predict_sim`:

```{r load-predictions}
data("predict_sim", package = "imuGAP")
predict_sim
```

---

## 5. Summarizing and Visualizing Predictions

We compute summary statistics (posterior median, mean, and credible intervals) using `summary()`:

```{r summarize-predictions}
summary_predict <- summary(predict_sim)
head(summary_predict)
```

### State-Level Coverage Trajectory

Below, we plot the predicted state-level two-dose coverage across ages 5 to 18 against the true simulation baseline:

```{r state-viz, fold = TRUE, code_summary = "Show plot code"}
data("latent_params_sim", package = "imuGAP")

state_predict <- summary_predict[loc_id == "State" & dose == 2 & age > 4]
state_idx <- predict_sim$target[loc_id == "State" & dose == 2 & age > 4, which = TRUE]
state_predict[, latent := latent_params_sim$coverage[state_idx]]

ggplot(state_predict) +
  aes(x = age) +
  geom_ribbon(aes(ymin = q2_5, ymax = q97_5, fill = "95% Credible Interval"), alpha = 0.25) +
  geom_line(aes(y = q50, color = "Posterior Median"), linewidth = 0.8) +
  geom_line(aes(y = latent, color = "True Latent"), linetype = "dashed", linewidth = 0.8) +
  scale_x_continuous(breaks = 5:18, minor_breaks = NULL) +
  coord_cartesian(ylim = c(0.8, 1.0)) +
  scale_color_manual(
    name = NULL,
    values = c("Posterior Median" = "black", "True Latent" = "firebrick")
  ) +
  scale_fill_manual(name = NULL, values = c("95% Credible Interval" = "grey50")) +
  theme(
    legend.position = "inside",
    legend.position.inside = c(0.05, 0.05),
    legend.justification.inside = c(0, 0)
  ) +
  labs(x = "Age", y = "State-Level Two-Dose Coverage")
```

---

### County-Level Estimates

We can similarly inspect county-level coverage estimates:

```{r county-viz, fold = TRUE, code_summary = "Show plot code"}
summary_predict |>
  subset(loc_id %in% c("Scruggs", "Simone", "Watson") & dose == 2 & age > 4) |>
  transform(loc_id = factor(loc_id, levels = c("Simone", "Watson", "Scruggs"))) |>
  ggplot() +
  aes(x = age) +
  geom_line(aes(y = q50, color = loc_id)) +
  geom_ribbon(aes(ymin = q2_5, ymax = q97_5, fill = loc_id), alpha = 0.2) +
  theme(
    legend.position = "inside",
    legend.position.inside = c(0.12, 0.05),
    legend.justification.inside = c(0, 0)
  ) +
  scale_x_continuous(breaks = 5:18, minor_breaks = NULL) +
  coord_cartesian(ylim = c(0.8, 1.0)) +
  scale_color_discrete(NULL, aesthetics = c("color", "fill")) +
  labs(x = "Age", y = "County-Level Two-Dose Coverage")
```

---

### Extracting Long-Format Draws for Custom Analysis

Use `as.data.frame()` to extract individual MCMC draws in a tidy, long format:

```{r long-draws}
draws_df <- as.data.frame(predict_sim)
head(draws_df)
```

---

# Summary and Guide to Other Vignettes

To explore advanced features and deeper technical details, see the following companion vignettes:

- **[Included Example Datasets and Data Validation](example_data.html)**:
  Detailed documentation of the bundled simulation fixtures, surveillance stream types (ChildVaxView, SchoolVaxView, TeenVaxView), canonicalization rules, and validation error diagnostics.

- **[Flexible Location Layers in imuGAP](user_specified_layers.html)**:
  Demonstrates how `imuGAP` dynamically adapts to 1-layer, 2-layer, and multi-layer spatial hierarchies, showing how partial pooling enables sub-regional estimation.

- **[Fit Inspection and Stan Diagnostics](examining_fits.html)**:
  Comprehensive guide to MCMC convergence assessment, parameter extraction with `extract_imugap()`, and parameter recovery trace plots for $\beta_{\text{bs}}$, $\sigma_{\text{layer}}$, $\delta_{\text{loc}}$, and $\lambda$.
