---
title: "Getting Started with imuGAP"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with imuGAP}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
resource_files:
  - figures/fit_predict_workflow.svg
  - figures/fit_predict_workflow.pdf
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  out.width = "100%",
  dev.args = list(bg = "white")
)
```

```{r setup, include = FALSE}
if (requireNamespace("thematic", quietly = TRUE)) {
  thematic::thematic_off()
}

suppressPackageStartupMessages({
  suppressWarnings({
    res <- tryCatch(
      pkgload::load_all(quiet = TRUE),
      error = function(e) FALSE
    )
    if (inherits(res, "error") || isFALSE(res)) {
      message("Vignette building without dev tools...")
    } else {
      res
    }
  })
})

# Define a knitr hook to fold source code into a <details> block below the plot
local({
  hook_source <- knitr::knit_hooks$get("source")
  knitr::knit_hooks$set(source = function(x, options) {
    res <- hook_source(x, options)
    if (isTRUE(options$fold) || !is.null(options$code_summary)) {
      label <- options$code_summary
      if (is.null(label)) label <- "Show code"
      sprintf("\n<details><summary><b>%s</b></summary>\n\n%s\n</details>\n", label, res)
    } else {
      res
    }
  })

  hook_chunk <- knitr::knit_hooks$get("chunk")
  knitr::knit_hooks$set(chunk = function(x, options) {
    res <- hook_chunk(x, options)
    if (isTRUE(options$fold)) {
      sprintf("\n<div class=\"chunk-plot-first\">\n%s\n</div>\n", res)
    } else {
      res
    }
  })
})

required_pkgs <- c("imuGAP", "data.table", "ggplot2")
pkgs_available <- all(sapply(required_pkgs, requireNamespace, quietly = TRUE))

if (!pkgs_available) {
  knitr::opts_chunk$set(eval = FALSE)
  message("Some suggested packages are missing. Vignette code will not be executed.")
} else {
  chk <- suppressPackageStartupMessages(
    sapply(required_pkgs, require, character.only = TRUE)
  )

  # Always enforce white background and black text on plots
  ggplot2::theme_set(
    ggplot2::theme_bw() +
      ggplot2::theme(
        plot.background = ggplot2::element_rect(fill = "white", colour = NA),
        panel.background = ggplot2::element_rect(fill = "white", colour = NA),
        legend.background = ggplot2::element_rect(fill = "white", colour = NA),
        legend.box.background = ggplot2::element_blank(),
        text = ggplot2::element_text(colour = "black"),
        axis.text = ggplot2::element_text(colour = "black"),
        axis.title = ggplot2::element_text(colour = "black"),
        plot.title = ggplot2::element_text(colour = "black"),
        plot.subtitle = ggplot2::element_text(colour = "black"),
        strip.text = ggplot2::element_text(colour = "black"),
        strip.background = ggplot2::element_rect(fill = "grey90", colour = "grey20")
      )
  )
}
```

# Introduction

`imuGAP` (Immunity: Geographic & Age-based Projection) estimates an underlying process model for vaccination on an arbitrary hierarchical population structure, which can then be used to predict vaccination coverage. Internally, `imuGAP` performs Bayesian inference via the Stan probabilistic programming language.

The package enables researchers and public health analysts to synthesize across multiple heterogeneous surveillance streams (such as national surveys, state school entry records, and local coverage censuses) to:

1. **Estimate current vaccination coverage** across arbitrary sub-populations and geographic levels.
2. **Project cohort trajectories** (e.g. coverage at each age from birth through adolescence for a given birth cohort).
3. **Impute missing data gaps** across unobserved years, age groups, or sub-regions by leveraging inferred relationships via the process model.

### Core Model Intuition

The core `imuGAP` model represents each target population slice $(i, a)$ (location $i$, birth cohort $a$) as having an underlying non-uptake rate $\phi_{i,a}$, with complementary uptake propensity $1 - \phi_{i,a}$. Given a dose eligibility schedule $\nu(s)$ and a force-of-vaccination rate $\lambda(s)$, cumulative coverage at age $t$ for the first dose is modeled as:

$$
P(\ge\textrm{1 dose}) = \left(1 - \phi_{i, a}\right) \left(1 - \exp\left\{-\int_a^{t} \lambda_{i, a}(s)\nu(s) d\textrm{s}\right\}\right)
$$

This formulation extends sequentially to multiple vaccine doses, conditional on receipt of prior doses. Sequential multi-dose regimens (e.g. Dose 1 at $\ge 12$ months, Dose 2 at $\ge 4$ years) are modeled as a continuous-time Markov transition process.

---

# The Fit & Predict Loop

The following diagram shows the end-to-end fit and predict workflow:

```{r workflow-diagram, echo = FALSE, out.width = "100%", fig.align = "center"}
is_html <- knitr::is_html_output()
diag_file <- if (is_html) "figures/fit_predict_workflow.svg" else "figures/fit_predict_workflow.pdf"
if (file.exists(diag_file)) {
  knitr::include_graphics(diag_file)
}
```

---

## 1. Preparing the Inputs

`imuGAP` requires three core input tables:

1. `locations`: The nested population hierarchy (e.g. State $\rightarrow$ County $\rightarrow$ School).
   This may optionally include population sizes, which may be necessary to ensure accurate results
   when entities within the same layer differ substantially in size.
2. `observations`: Empirical survey counts (`positive` and `sample_n`) and censoring indicators.
3. `populations`: Metadata and fractional weights mapping observations to discrete location/age/cohort/dose slices.

```{r load-data}
library(imuGAP)
library(data.table)
library(ggplot2)

data("locations_sim", package = "imuGAP")
data("observations_sim", package = "imuGAP")
data("populations_sim", package = "imuGAP")

head(locations_sim)
head(observations_sim[, .(obs_id, loc_id, positive, sample_n, censored)])
head(populations_sim)
```

Input datasets are automatically validated and canonicalized during sampling via
`canonicalize_locations()`, `canonicalize_observations()`, and `canonicalize_populations()`.

For an in-depth walkthrough of flexible surveillance stream structures—including right-censored
observations (such as when surveillance reports only complete coverage for a bundle of vaccines,
not specifically the target vaccine) and composite observations that mix multiple age groups or
populations—see **[Included Example Datasets and Data Validation](example_data.html)**.

---

## 2. Fitting the Model

You can fit the model to your data by calling `sampling()`. Configure model options (such as B-spline degrees of freedom and dose eligibility ages) with `imugap_options()`, and specify MCMC sampler settings (chains, iterations, backend) with `stan_options()`:

```{r sampling-code, eval = FALSE}
# Configure model options
im_opts <- imugap_options(df = 5L, dose_schedule = c(1L, 4L))

# Configure Stan MCMC sampler
st_opts <- stan_options(chains = 4, iter = 2000, seed = 1L)

# Fit the model
fit <- sampling(
  observations = observations_sim,
  populations = populations_sim,
  locations = locations_sim,
  imugap_opts = im_opts,
  stan_opts = st_opts
)
```

For this vignette, you can load the pre-computed fit fixture `fit_sim` bundled with the package:

```{r load-fit}
data("fit_sim", package = "imuGAP")
fit_sim
```

---

## 3. Defining a Prediction Target Grid

To generate coverage predictions for specific populations, define a target grid using `create_target()`.

For example, to evaluate a cross-sectional "snapshot" across all locations for ages 1 through 18 and doses 1 and 2:

```{r target}
target_grid <- create_target(
  location = unique(locations_sim$loc_id),
  age = 1:18,
  cohort = max(populations_sim$cohort) - 18,
  dose = c(1, 2),
  mode = "snapshot"
)
head(target_grid)
```

---

## 4. Predicting Coverage

Pass the model fit and target grid to `predict()` to generate posterior draws of coverage on the natural (probability) scale:

```{r predict-code, eval = FALSE}
# Generate posterior draws for each target slice
predict_res <- predict(object = fit_sim, target = target_grid, posterior_size = 100)
```

You can load the pre-computed prediction fixture `predict_sim`:

```{r load-predictions}
data("predict_sim", package = "imuGAP")
predict_sim
```

---

## 5. Summarizing and Visualizing Predictions

You can compute vaccination coverage estimate summary statistics (posterior mean, median, and credible intervals) on the natural (probability) scale across your prediction targets using `summary()`:

```{r summarize-predictions}
summary_predict <- summary(predict_sim)

# Filter coverage estimate statistics to a specific target slice (e.g. age 4, dose 2)
summary_predict[age == 4 & dose == 2]
```

### State-Level Coverage Trajectory

The following plot shows predicted state-level two-dose coverage across ages 5 to 18 against the true simulation baseline:

```{r state-viz, fold = TRUE, code_summary = "Show plot code"}
data("latent_params_sim", package = "imuGAP")

state_predict <- summary_predict[loc_id == "State" & dose == 2 & age > 4]
state_idx <- predict_sim$target[loc_id == "State" & dose == 2 & age > 4, which = TRUE]
state_predict[, latent := latent_params_sim$coverage[state_idx]]

ggplot(state_predict) +
  aes(x = age) +
  geom_ribbon(aes(ymin = q2_5, ymax = q97_5, fill = "95% Credible Interval"), alpha = 0.25) +
  geom_line(aes(y = q50, color = "Posterior Median"), linewidth = 0.8) +
  geom_line(aes(y = latent, color = "True Latent"), linetype = "dashed", linewidth = 0.8) +
  scale_x_continuous(breaks = 5:18, minor_breaks = NULL) +
  coord_cartesian(ylim = c(0.8, 1.0)) +
  scale_color_manual(
    name = NULL,
    values = c("Posterior Median" = "black", "True Latent" = "firebrick")
  ) +
  scale_fill_manual(name = NULL, values = c("95% Credible Interval" = "grey50")) +
  theme(
    legend.position = "inside",
    legend.position.inside = c(0.05, 0.05),
    legend.justification.inside = c(0, 0)
  ) +
  labs(x = "Age", y = "State-Level Two-Dose Coverage")
```

---

### County-Level Estimates

The following plot compares county-level coverage estimates across counties:

```{r county-viz, fold = TRUE, code_summary = "Show plot code"}
summary_predict |>
  subset(loc_id %in% c("Scruggs", "Simone", "Watson") & dose == 2 & age > 4) |>
  transform(loc_id = factor(loc_id, levels = c("Simone", "Watson", "Scruggs"))) |>
  ggplot() +
  aes(x = age) +
  geom_line(aes(y = q50, color = loc_id)) +
  geom_ribbon(aes(ymin = q2_5, ymax = q97_5, fill = loc_id), alpha = 0.2) +
  theme(
    legend.position = "inside",
    legend.position.inside = c(0.12, 0.05),
    legend.justification.inside = c(0, 0)
  ) +
  scale_x_continuous(breaks = 5:18, minor_breaks = NULL) +
  coord_cartesian(ylim = c(0.8, 1.0)) +
  scale_color_discrete(NULL, aesthetics = c("color", "fill")) +
  labs(x = "Age", y = "County-Level Two-Dose Coverage")
```

---

### Extracting Long-Format Draws for Custom Analysis

Use `as.data.frame()` to extract individual MCMC draws in a tidy, long format:

```{r long-draws}
draws_df <- as.data.frame(predict_sim)
head(draws_df)
```

---

# Summary and Guide to Other Vignettes

To explore advanced features and deeper technical details, see the following companion vignettes:

- **[Included Example Datasets and Data Validation](example_data.html)**:
  Detailed documentation of the bundled simulation fixtures, surveillance stream types (ChildVaxView, SchoolVaxView, TeenVaxView), canonicalization rules, and validation error diagnostics.

- **[Flexible Location Layers in imuGAP](user_specified_layers.html)**:
  Demonstrates how `imuGAP` estimates the process model across 1-layer, 2-layer, and multi-layer
  hierarchical population structures, showing how inferred relationships enable sub-regional
  estimation and cohort projection.

- **[Fit Inspection and Stan Diagnostics](examining_fits.html)**:
  Comprehensive guide to MCMC convergence assessment, parameter extraction with `extract_imugap()`,
  and parameter recovery trace plots for $\beta_{\text{bs}}$, $\sigma_{\text{layer}}$,
  $\delta_{\text{loc}}$, and $\lambda$.
