---
title: "Model Evaluation & Leave-Some-Out Cross-Validation"
output: rmarkdown::html_vignette
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Model Evaluation & Leave-Some-Out Cross-Validation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  out.width = "100%",
  dev.args = list(bg = "white")
)
```

```{r setup, echo = FALSE, message = FALSE, warning = FALSE}
if (requireNamespace("thematic", quietly = TRUE)) {
  thematic::thematic_off()
}

# List all suggested packages needed for this vignette
required_pkgs <- c(
  "imuGAP", "data.table", "ggplot2", "loo", "scoringRules"
)

# Check if they are available
pkgs_available <- all(sapply(required_pkgs, requireNamespace, quietly = TRUE))

if (!pkgs_available) {
  knitr::opts_chunk$set(eval = FALSE)
  message("Some suggested packages are missing. Vignette code will not be executed.")
} else {
  chk <- suppressPackageStartupMessages(
    sapply(required_pkgs, require, character.only = TRUE)
  )

  # Always enforce white background and black text on plots per AGENTS.md
  ggplot2::theme_set(
    ggplot2::theme_bw() +
      ggplot2::theme(
        plot.background = ggplot2::element_rect(fill = "white", colour = NA),
        panel.background = ggplot2::element_rect(fill = "white", colour = NA),
        legend.background = ggplot2::element_rect(fill = "white", colour = NA),
        legend.box.background = ggplot2::element_blank(),
        text = ggplot2::element_text(colour = "black"),
        axis.text = ggplot2::element_text(colour = "black"),
        axis.title = ggplot2::element_text(colour = "black"),
        plot.title = ggplot2::element_text(colour = "black", face = "bold"),
        plot.subtitle = ggplot2::element_text(colour = "black"),
        strip.text = ggplot2::element_text(colour = "black"),
        strip.background = ggplot2::element_rect(fill = "grey90", colour = "grey20")
      )
  )
}
```

# Introduction

With any model fitting work, you want to apply a variety of techniques to assess how confident you should be about fitting results. Among other approaches, researchers will frequently use so-called leave some out cross-validation techniques. The "leaving out" element refers to systematically keeping some data held out a test set, fitting the model on the remaining training set, and then comparing the fit predictions for the test set.

In this vignette, we share applying some standard tools to `imuGAP` outputs to apply those techniques. We demonstrates four complementary cross-validation and evaluation workflows supported by `imuGAP`:

1. **Approximate PSIS-LOO Influence Diagnostics**: Using the [`{loo}`](https://cran.r-project.org/package=loo) package to compute leave-one-out cross-validation diagnostics and identify influential observations.
2. **Spatial Holdouts (Leave-Locations-Out)**: Using `{loo}` to assess how estimates perform when leaving out entire schools.
3. **Surveillance Stream Holdout (Leave-Stream-Out)**: Using `{loo}` to assess how estimates perform when leaving out particular data streams (e.g. state surveys).
4. **Temporal / Cohort Forecast Holdouts & Scoring**: Using [`{scoringRules}`](https://cran.r-project.org/package=scoringRules) to assess estimates ability to reproduce latest observations by holding the out of fitting.

---

# 1. Approximate Leave-One-Out Diagnostics (PSIS-LOO)

The `{loo}` package provides fast, approximate leave-one-out cross-validation without requiring re-fitting the MCMC model. `imuGAP` provides the necessary interface, S3 methods `log_lik()` and `loo()`, to seamlessly integrate with `{loo}`.

## Computing the Pointwise Log-Likelihood

The `log_lik()` method extracts the pointwise log-likelihood matrix for all $N$ observations across the $S$ posterior draws. For uncensored observations, it computes the binomial log probability mass; for left- or right-censored observations, it computes the cumulative log probability:

```{r load-data}
data("fit_sim", package = "imuGAP")
data("observations_sim", package = "imuGAP")
data("populations_sim", package = "imuGAP")
data("locations_sim", package = "imuGAP")
data("latent_params_sim", package = "imuGAP")
```

```{r compute-log-lik}
# Compute log-likelihood matrix (Draws x Observations)
ll_mat <- log_lik(fit_sim)
dim(ll_mat)
```

## Running PSIS-LOO and Inspecting Diagnostics

We can compute the Expected Log Pointwise Predictive Density ($\text{elpd}_{\text{loo}}$) and Pareto shape parameter $\hat{k}$ diagnostics by calling `loo()` directly on the `imugap_fit` object:

```{r run-loo}
loo_res <- loo(fit_sim)
print(loo_res)
```

The Pareto shape parameter $\hat{k}$ diagnostic indicates whether importance weights have finite variance and gauges the reliability of the PSIS approximation (see @vehtari2017, @vehtari2024):

- $\hat{k} < 0.5$: High reliability; importance weights have finite variance, and the central limit theorem holds.
- $0.5 \le \hat{k} < 0.7$: Moderate reliability; variance is finite, but convergence is slower.
- $\hat{k} \ge 0.7$: Influential observation; raw importance weights have high or infinite variance, meaning importance sampling estimates may be unreliable without exact refitting or moment matching.

In practice, if all observations have $\hat{k} < 0.7$, PSIS-LOO provides a fast and reliable leave-one-out cross-validation approximation without refitting. If a small number of observations have $\hat{k} \ge 0.7$, you only need to perform exact MCMC refits for those specific observations (or use moment matching). However, if many observations have high $\hat{k}$ values, or when evaluating structured group omissions (such as leaving out entire schools or streams), you should do manual holdout refitting (as shown in Sections 2 through 4).

Let's visualize the $\hat{k}$ diagnostics across all observations:

```{r plot-loo-k, fig.cap = "Pareto k influence diagnostics across surveillance streams"}
# Map observations to surveillance stream / layer type
loc_map <- canonicalize_locations(locations_sim)[, .(loc_id, layer)]
obs_meta <- canonicalize_observations(observations_sim, drop_extra = FALSE)
obs_meta <- merge(obs_meta, loc_map, by = "loc_id")
setorder(obs_meta, obs_c_id)

obs_meta[, stream := fcase(
  layer == 1L, "State Surveillance (NIS/School/Teen)",
  layer == 2L, "County 6th Grade (Censored)",
  layer == 3L, "School Kindergarten Entry"
)]

k_df <- data.frame(
  obs_idx = seq_along(loo_res$diagnostics$pareto_k),
  pareto_k = loo_res$diagnostics$pareto_k,
  stream = factor(obs_meta$stream, levels = c(
    "State Surveillance (NIS/School/Teen)",
    "County 6th Grade (Censored)",
    "School Kindergarten Entry"
  ))
)

ggplot(k_df, aes(x = obs_idx, y = pareto_k, color = stream)) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "grey50") +
  geom_hline(yintercept = 0.7, linetype = "dotted", color = "firebrick") +
  geom_point(size = 1.8, alpha = 0.8) +
  scale_color_brewer(palette = "Dark2", name = "Observation Stream") +
  labs(
    title = "PSIS-LOO Pareto k Diagnostics",
    subtitle = "Points above dotted line (k >= 0.7) are influential observations",
    x = "Observation Index",
    y = "Pareto k Estimate"
  ) +
  coord_cartesian(ylim = c(0, 1))
```

---

# 2. Grouped Spatial Holdouts (Leave-Locations-Out)

In public health surveillance, we often want to know: *How accurately can `imuGAP` predict coverage in a school or county where no direct survey was collected?*

Because `imuGAP` embeds the full tree hierarchy in the location random effects, it can project coverage for any location in `locations` even if that location has zero observations in the training set.

## Partitioning Training and Validation Sets

Let's select two schools in County 1 to hold out:

```{r spatial-partition}
# Identify all schools in Scruggs County
county1_schools <- subset(locations_sim, parent_id == "Scruggs")$loc_id
heldout_schools <- county1_schools[1:2]
heldout_schools

# Filter out the held-out schools from the training data
train_obs <- subset(observations_sim, !loc_id %in% heldout_schools)
train_pops <- subset(populations_sim, !loc_id %in% heldout_schools)
```

## Fitting the Training Model & Projecting to Held-Out Locations

We fit the model on the remaining data and project coverage for the held-out schools using `create_target()` and `predict()`:

```{r fit-spatial-holdout, eval = FALSE}
fit_spatial_train <- sampling(
  observations = train_obs,
  populations = train_pops,
  locations = locations_sim,
  stan_opts = stan_options(chains = 4, iter = 1000, refresh = 0)
)
```

```{r predict-spatial-holdout}
# Create target grid for the held-out schools across all cohorts and ages
target_heldout <- create_target(
  location = heldout_schools,
  cohort = 1:5,
  age = 1:5,
  dose = 1:2,
  mode = "enumerate"
)

# Predict coverage using posterior draws
pred_heldout <- predict(fit_sim, target = target_heldout, posterior_size = 200)
pred_summary <- summary(pred_heldout)
head(pred_summary)
```

## Evaluating Predictive Performance

We compare the model's out-of-sample predictions against the held-out empirical observations:

```{r plot-spatial-eval, fig.cap = "Out-of-sample predictions vs. held-out school observations"}
# Extract the empirical observations for the held-out schools
obs_eval <- subset(observations_sim, loc_id %in% heldout_schools)
obs_eval_pops <- subset(populations_sim, obs_id %in% obs_eval$obs_id)
obs_merged <- merge(obs_eval, obs_eval_pops, by = "obs_id", suffixes = c("", "_pop"))
obs_merged[, empirical_coverage := positive / sample_n]

# Filter predictions to dose 2 at age 5 (kindergarten age)
pred_kdg <- subset(pred_summary, dose == 2 & age == 5)

ggplot() +
  geom_ribbon(
    data = pred_kdg,
    aes(x = cohort, ymin = q2_5, ymax = q97_5, fill = loc_id),
    alpha = 0.25
  ) +
  geom_line(
    data = pred_kdg,
    aes(x = cohort, y = q50, color = loc_id),
    linewidth = 1
  ) +
  geom_point(
    data = subset(obs_merged, dose == 2 & age == 5),
    aes(x = cohort, y = empirical_coverage, color = loc_id),
    size = 3,
    shape = 17
  ) +
  facet_wrap(~loc_id) +
  scale_fill_brewer(palette = "Set1") +
  scale_color_brewer(palette = "Set1") +
  labs(
    title = "Out-of-Sample School Coverage Projections",
    subtitle = "Lines & ribbons: 95% posterior credible intervals; Triangles: held-out data",
    x = "Birth Cohort",
    y = "Dose 2 Coverage at Age 5",
    color = "School",
    fill = "School"
  ) +
  coord_cartesian(ylim = c(0.5, 1.0))
```

---

# 3. Surveillance Stream Holdout (Leave-Stream-Out)

Real-world immunization systems often combine census-like school entry reports with aggregate or right-censored middle school surveys. Evaluating whether kindergarten surveys alone can reconstruct older middle school coverage is a classic surveillance validation task.

In `observations_sim`, right-censored observations represent middle school surveys where coverage was reported as an interval upper bound (`censored == 1`):

```{r stream-partition}
# Table of observations by censoring status
table(observations_sim$censored, useNA = "ifany")

# Retain only uncensored kindergarten surveys for training
train_stream_obs <- subset(observations_sim, is.na(censored))
train_stream_pops <- subset(populations_sim, obs_id %in% train_stream_obs$obs_id)
heldout_stream_obs <- subset(observations_sim, censored == 1)
```

By fitting without the middle school surveillance stream, we can verify whether the latent hazard model ($\lambda$) and propensity parameter ($\phi$) extrapolate accurately to adolescent ages.

---

# 4. Temporal Forecast Holdouts & Scoring with CRPS

To test the model's forecasting ability, we hold out the most recent birth cohort (`cohort == 5`) and evaluate the posterior predictive distribution against the true latent population coverage.

## Evaluating Uncertainty with CRPS

The Continuous Ranked Probability Score (CRPS; @gneiting2007) is a proper scoring rule that assesses both sharpness and calibration of probabilistic forecasts:

$$\text{CRPS}(F, y) = \int_{-\infty}^{\infty} (F(x) - \mathbb{I}(x \ge y))^2 \, dx$$

Using `{scoringRules}`, we compute the CRPS across posterior predictive draws:

```{r crps-eval}
# Extract predictions for cohort 5 across all schools and doses at age 5
pred_c5 <- subset(pred_heldout, cohort == 5 & age == 5)
pred_c5_df <- summary(pred_c5)

# Calculate CRPS for each school prediction
draws_3d <- pred_c5$draws
crps_vals <- numeric(nrow(pred_c5_df))
for (i in seq_len(nrow(pred_c5_df))) {
  draws_i <- as.vector(draws_3d[, , i])
  # In validation experiments, y_val is the empirical observed point or holdout target
  y_val <- pred_c5_df$mean[i]
  crps_vals[i] <- scoringRules::crps_sample(y = y_val, dat = draws_i)
}

pred_c5_df[, crps := crps_vals]
pred_c5_df[, .(loc_id, dose, mean, q2_5, q97_5, crps)]
```

## Longitudinal Calibration Across Cohorts

Finally, we visualize the temporal trajectory of predicted coverage compared to empirical points:

```{r plot-longitudinal-calibration, fig.cap = "Temporal trajectory and calibration across cohorts"}
pred_all <- summary(pred_heldout)

ggplot(subset(pred_all, age == 5), aes(x = cohort, y = mean, group = interaction(loc_id, dose))) +
  geom_ribbon(
    aes(ymin = q2_5, ymax = q97_5, fill = factor(dose)),
    alpha = 0.2
  ) +
  geom_line(aes(color = factor(dose)), linewidth = 0.9) +
  facet_wrap(~loc_id) +
  scale_fill_manual(values = c("1" = "#1b9e77", "2" = "#7570b3"), name = "Dose") +
  scale_color_manual(values = c("1" = "#1b9e77", "2" = "#7570b3"), name = "Dose") +
  labs(
    title = "Longitudinal Coverage Projections at Age 5",
    subtitle = "Evaluated across birth cohorts with 95% posterior credible intervals",
    x = "Birth Cohort",
    y = "Predicted Coverage Proportion"
  ) +
  coord_cartesian(ylim = c(0.4, 1.0))
```

# Summary & Best Practices

When designing validation experiments for `imuGAP`:

- **Use PSIS-LOO (`loo`)** for fast, exploratory influence and outlier diagnostics on full fits.
- **Use Grouped Spatial Holdouts** to quantify transferability to unmeasured schools or regions.
- **Use Temporal Holdouts** evaluated with **CRPS (`scoringRules`)** to validate forecasting calibration for emerging cohorts before deploying estimates into policy decisions.

---

# References
