---
title: "Model Evaluation & Leave-Some-Out Cross-Validation"
output: rmarkdown::html_vignette
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Model Evaluation & Leave-Some-Out Cross-Validation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  out.width = "100%",
  dev.args = list(bg = "white")
)
```

```{r setup, echo = FALSE, message = FALSE, warning = FALSE}
if (requireNamespace("thematic", quietly = TRUE)) {
  thematic::thematic_off()
}

# Define a knitr hook to fold source code into a <details> block below the plot
local({
  hook_source <- knitr::knit_hooks$get("source")
  knitr::knit_hooks$set(source = function(x, options) {
    res <- hook_source(x, options)
    if (isTRUE(options$fold) || !is.null(options$code_summary)) {
      label <- options$code_summary
      if (is.null(label)) label <- "Show code"
      sprintf("\n<details><summary><b>%s</b></summary>\n\n%s\n</details>\n", label, res)
    } else {
      res
    }
  })

  hook_chunk <- knitr::knit_hooks$get("chunk")
  knitr::knit_hooks$set(chunk = function(x, options) {
    res <- hook_chunk(x, options)
    if (isTRUE(options$fold)) {
      sprintf("\n<div class=\"chunk-plot-first\">\n%s\n</div>\n", res)
    } else {
      res
    }
  })
})

# List all suggested packages needed for this vignette
required_pkgs <- c(
  "imuGAP", "data.table", "ggplot2", "loo", "scoringRules"
)

# Check if they are available
pkgs_available <- all(sapply(required_pkgs, requireNamespace, quietly = TRUE))

if (!pkgs_available) {
  knitr::opts_chunk$set(eval = FALSE)
  message("Some suggested packages are missing. Vignette code will not be executed.")
} else {
  chk <- suppressPackageStartupMessages(
    sapply(required_pkgs, require, character.only = TRUE)
  )

  # Always enforce white background and black text on plots per AGENTS.md
  ggplot2::theme_set(
    ggplot2::theme_bw() +
      ggplot2::theme(
        plot.background = ggplot2::element_rect(fill = "white", colour = NA),
        panel.background = ggplot2::element_rect(fill = "white", colour = NA),
        legend.background = ggplot2::element_rect(fill = "white", colour = NA),
        legend.box.background = ggplot2::element_blank(),
        text = ggplot2::element_text(colour = "black"),
        axis.text = ggplot2::element_text(colour = "black"),
        axis.title = ggplot2::element_text(colour = "black"),
        plot.title = ggplot2::element_text(colour = "black", face = "bold"),
        plot.subtitle = ggplot2::element_text(colour = "black"),
        strip.text = ggplot2::element_text(colour = "black"),
        strip.background = ggplot2::element_rect(fill = "grey90", colour = "grey20")
      )
  )
}
```

# Introduction

With any model fitting work, you want to apply a variety of techniques to assess how confident you should be about fitting results. Among other approaches, researchers will frequently use leave some out cross-validation techniques. The "leaving out" element refers to systematically keeping some data held out a test set, fitting the model on the remaining training set, comparing the fit predictions for the test set, and then moving on to a new slice of hold out data.

In this vignette, we share using standard tools with `imuGAP` outputs to apply those techniques. We demonstrate four complementary cross-validation and evaluation workflows:

1. **Approximate PSIS-LOO Influence Diagnostics**: Using the [`{loo}`](https://cran.r-project.org/package=loo) package to compute leave-one-out cross-validation diagnostics and identify influential observations.
2. **Spatial Holdouts (Leave-Each-School-Out)**: Systematically leaving out individual schools to evaluate out-of-sample prediction accuracy.
3. **Surveillance Stream Holdout (Leave-Stream-Out)**: Assessing how estimates perform when leaving out particular data streams (e.g. state surveys).
4. **Temporal / Cohort Forecast Holdouts & Scoring**: Using [`{scoringRules}`](https://cran.r-project.org/package=scoringRules) to assess estimates ability to reproduce latest observations by holding them out of fitting.

---

# 1. Approximate Leave-One-Out Diagnostics (PSIS-LOO)

The `{loo}` package provides fast, approximate leave-one-out cross-validation without requiring re-fitting the MCMC model. `imuGAP` provides the necessary interface, S3 methods `log_lik()` and `loo()`, to seamlessly integrate with `{loo}`.

## Computing the Pointwise Log-Likelihood

The `log_lik()` method extracts the pointwise log-likelihood matrix for all $N$ observations across the $S$ posterior draws. For uncensored observations, it computes the binomial log probability mass; for left- or right-censored observations, it computes the cumulative log probability:

```{r load-data}
data("fit_sim", package = "imuGAP")
data("predict_sim", package = "imuGAP")
data("observations_sim", package = "imuGAP")
data("populations_sim", package = "imuGAP")
data("locations_sim", package = "imuGAP")
data("latent_params_sim", package = "imuGAP")
data("leave_school_out_scruggs", package = "imuGAP")
data("leave_stream_out", package = "imuGAP")
data("leave_recent_out", package = "imuGAP")
```

```{r compute-log-lik}
# Compute log-likelihood matrix (Draws x Observations)
ll_mat <- log_lik(fit_sim)
dim(ll_mat)
```

## Running PSIS-LOO and Inspecting Diagnostics

We can compute the Expected Log Pointwise Predictive Density ($\text{elpd}_{\text{loo}}$) and Pareto shape parameter $\hat{k}$ diagnostics by calling `loo()` directly on the `imugap_fit` object:

```{r run-loo}
loo_res <- loo(fit_sim)
print(loo_res)
```

The Pareto shape parameter $\hat{k}$ diagnostic indicates whether importance weights have finite variance and gauges the reliability of the PSIS approximation (see @vehtari2017, @vehtari2024):

- $\hat{k} < 0.5$ (including $< 0$): High reliability; importance weights have finite variance, and the central limit theorem holds.
- $0.5 \le \hat{k} < 0.7$: Moderate reliability; variance is finite, but convergence is slower.
- $\hat{k} \ge 0.7$: Influential observation; raw importance weights have high or infinite variance, meaning importance sampling estimates may be unreliable without exact refitting or moment matching.

In practice, if all observations have $\hat{k} < 0.7$, PSIS-LOO provides a fast and reliable leave-one-out cross-validation approximation without refitting. If a small number of observations have $\hat{k} \ge 0.7$, you only need to perform exact MCMC refits for those specific observations (or use moment matching). However, if many observations have high $\hat{k}$ values, or when evaluating structured group omissions (such as leaving out entire schools or streams), you should do manual holdout refitting (as shown in Sections 2 through 4).

Let's visualize the $\hat{k}$ diagnostics across all observations:

```{r plot-loo-k, fold = TRUE, code_summary = "Show plot code", fig.cap = "Pareto k influence diagnostics across surveillance streams"}
# Map observations to surveillance stream / layer type
loc_map <- canonicalize_locations(locations_sim)[, .(loc_id, layer)]
obs_meta <- canonicalize_observations(observations_sim, drop_extra = FALSE)
obs_meta <- merge(obs_meta, loc_map, by = "loc_id")
setorder(obs_meta, obs_c_id)

obs_meta[, stream := fcase(
  layer == 1L, "State Surveys",
  layer == 2L, "County Surveys (Censored)",
  layer == 3L, "School Entry"
)]

k_df <- data.frame(
  obs_idx = seq_along(loo_res$diagnostics$pareto_k),
  pareto_k = loo_res$diagnostics$pareto_k,
  stream = factor(obs_meta$stream, levels = c(
    "School Entry",
    "State Surveys",
    "County Surveys (Censored)"
  ))
)

stream_midpoints <- tapply(k_df$obs_idx, k_df$stream, mean)

ggplot(k_df) +
  # Painted background regions for k thresholds
  annotate(
    "rect",
    xmin = -Inf,
    xmax = Inf,
    ymin = -Inf,
    ymax = 0.5,
    fill = "#2ca02c",
    alpha = 0.07
  ) +
  annotate(
    "rect",
    xmin = -Inf,
    xmax = Inf,
    ymin = 0.5,
    ymax = 0.7,
    fill = "#e6ab02",
    alpha = 0.12
  ) +
  annotate(
    "rect",
    xmin = -Inf,
    xmax = Inf,
    ymin = 0.7,
    ymax = Inf,
    fill = "#d95f02",
    alpha = 0.15
  ) +
  # Pareto k quantitative region labels
  annotate(
    "label",
    x = 10L,
    y = 0.75,
    label = "Influential (k >= 0.7)",
    hjust = 0,
    size = 3.2,
    fontface = "italic",
    color = "#b30000",
    fill = alpha("white", 0.6)
  ) +
  annotate(
    "label",
    x = 10L,
    y = 0.60,
    label = "Moderate (0.5 <= k < 0.7)",
    hjust = 0,
    size = 3.2,
    fontface = "italic",
    color = "#8c6d00",
    fill = alpha("white", 0.6)
  ) +
  annotate(
    "label",
    x = 10L,
    y = 0.45,
    label = "Good (k < 0.5)",
    hjust = 0,
    size = 3.2,
    fontface = "italic",
    color = "#1b7837",
    fill = alpha("white", 0.6)
  ) +
  # Observation points
  geom_point(aes(x = obs_idx, y = pareto_k, color = stream), size = 1.4, alpha = 0.75) +
  # Surveillance stream labels
  annotate(
    "label",
    x = stream_midpoints["School Entry"],
    y = 0.45,
    label = "School Kindergarten Entry",
    fontface = "bold",
    size = 3.3,
    color = "#1b9e77",
    fill = alpha("white", 0.6)
  ) +
  annotate(
    "label",
    x = stream_midpoints["State Surveys"],
    y = 0.10,
    label = "State Surveys",
    fontface = "bold",
    size = 3.3,
    color = "#d95f02",
    angle = 90,
    fill = alpha("white", 0.6)
  ) +
  annotate(
    "label",
    x = stream_midpoints["County Surveys (Censored)"],
    y = 0.10,
    label = "County Surveys (Censored)",
    fontface = "bold",
    size = 3.3,
    color = "#7570b3",
    angle = 90,
    fill = alpha("white", 0.6)
  ) +
  scale_color_brewer(palette = "Dark2") +
  scale_x_continuous(expand = c(0, 0), breaks = NULL) +
  scale_y_continuous(
    expand = c(0, 0),
    limits = c(-0.20, 1.0),
    breaks = seq(-0.2, 1.0, by = 0.1)
  ) +
  guides(color = "none") +
  labs(
    x = "Observation Index",
    y = expression("Pareto " * hat(k) * " Diagnostic")
  ) +
  theme(
    axis.ticks.x = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank()
  )
```

---

# 2. Grouped Spatial Holdouts (Leave-Locations-Out)

The `imuGAP` approach focuses on partitioned populations observed across hierarchical resolutions. In real-world surveillance systems, some populations may be well observed while others represent known-unknowns: for example, we might have county-level survey data and observations from *most* schools in a county, but some schools were systematically unmeasured.

When evaluating predictive performance on unmeasured areas, standard random leave-one-out or random $K$-fold cross-validation is overly optimistic: if one observation from School A is held out while other cohorts from School A remain in the training set, the model has already learned School A's specific random effect ($\alpha_{\text{School A}}$).

To evaluate true spatial generalizability, we must leave out **entire location clusters** at once.

## Approximate Leave-One-School-Out Diagnostics (Grouped PSIS-LOO)

Before performing exact MCMC refitting for each spatial holdout, we can evaluate whether PSIS-LOO can approximate leaving out whole schools at once.

For each school $g \in \{1, \dots, G\}$, we sum the pointwise log-likelihood draws across all observations belonging to that school:

$$\log p(y_{\text{school } g} \mid \theta^{(s)}) = \sum_{j \in \text{school } g} \log p(y_j \mid \theta^{(s)})$$

Passing this grouped log-likelihood matrix to `{loo}` computes Pareto $\hat{k}$ diagnostics for each school:

```{r run-school-loo}
# Identify all schools (layer 3) and map to county parent
locs <- canonicalize_locations(locations_sim)
schools <- locs[layer == 3L, .(loc_id, parent_id)]
setorder(schools, parent_id, loc_id)

obs_meta <- canonicalize_observations(observations_sim, drop_extra = FALSE)
ll <- log_lik(fit_sim)

# Sum log-likelihood draws across observations for each school
school_ll <- sapply(schools$loc_id, function(s) {
  cols <- which(obs_meta$loc_id == s)
  rowSums(ll[, cols, drop = FALSE])
})

school_loo <- loo::loo(school_ll, r_eff = rep(1, ncol(school_ll)))
print(school_loo)
```

Let's visualize the school-level Pareto $\hat{k}$ diagnostics:

```{r plot-school-loo-k, fold = TRUE, code_summary = "Show plot code", fig.cap = "School-level Pareto k diagnostics (Leave-One-School-Out)"}
k_school_df <- data.frame(
  school_idx = seq_len(nrow(schools)),
  school_name = schools$loc_id,
  county = factor(schools$parent_id),
  pareto_k = school_loo$diagnostics$pareto_k
)

ggplot(k_school_df) +
  # Painted background regions for k thresholds
  annotate(
    "rect",
    xmin = -Inf,
    xmax = Inf,
    ymin = -Inf,
    ymax = 0.5,
    fill = "#2ca02c",
    alpha = 0.07
  ) +
  annotate(
    "rect",
    xmin = -Inf,
    xmax = Inf,
    ymin = 0.5,
    ymax = 0.7,
    fill = "#e6ab02",
    alpha = 0.12
  ) +
  annotate(
    "rect",
    xmin = -Inf,
    xmax = Inf,
    ymin = 0.7,
    ymax = Inf,
    fill = "#d95f02",
    alpha = 0.15
  ) +
  # Pareto k quantitative region labels
  annotate(
    "label",
    x = 1L,
    y = 1.30,
    label = "Influential (k >= 0.7)",
    hjust = 0,
    size = 3.2,
    fontface = "italic",
    color = "#b30000",
    fill = alpha("white", 0.6)
  ) +
  annotate(
    "label",
    x = 1L,
    y = 0.60,
    label = "Moderate (0.5 <= k < 0.7)",
    hjust = 0,
    size = 3.2,
    fontface = "italic",
    color = "#8c6d00",
    fill = alpha("white", 0.6)
  ) +
  annotate(
    "label",
    x = 1L,
    y = 0.45,
    label = "Good (k < 0.5)",
    hjust = 0,
    size = 3.2,
    fontface = "italic",
    color = "#1b7837",
    fill = alpha("white", 0.6)
  ) +
  # School points
  geom_point(aes(x = school_idx, y = pareto_k, color = county), size = 2.5, alpha = 0.85) +
  scale_color_brewer(palette = "Dark2") +
  scale_x_continuous(
    expand = c(0.04, 0.04),
    breaks = k_school_df$school_idx,
    labels = k_school_df$school_name
  ) +
  scale_y_continuous(
    expand = c(0, 0),
    limits = c(0.30, 1.45),
    breaks = seq(0.4, 1.4, by = 0.2)
  ) +
  guides(color = guide_legend(title = "County")) +
  labs(
    x = "School (Grouped by County)",
    y = expression("School Pareto " * hat(k) * " Diagnostic")
  ) +
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 8),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank()
  )
```

Notice that while almost all individual observations had $\hat{k} < 0.5$ in Section 1, leaving out an entire school yields $\hat{k} \ge 0.7$ for the majority of schools. Because omitting a school removes all local evidence and isolates its latent random effect ($\alpha_{\text{school}}$), the posterior distribution shifts substantially. This diagnostic confirms that spatial holdout evaluation requires exact MCMC refitting.

## Systematic Leave-Each-School-Out Workflow

Because `imuGAP` associates observations with explicit location identifiers (`loc_id`) in `populations_sim`, spatial holdouts can be constructed directly by filtering out one school at a time. This leave-each-school-out cross-validation mimics real-world scenarios where an entire school or community has no direct survey measurements.

To hold out a single school (for example, `"School_01"`), identify its corresponding observation IDs and omit them from the training inputs:

```{r school-holdout-split}
target_school <- "School_01"

# Identify and exclude observation IDs belonging to the held-out school
heldout_obs_ids <- subset(populations_sim, loc_id == target_school)$obs_id
train_obs <- subset(observations_sim, !obs_id %in% heldout_obs_ids)
train_pops <- subset(populations_sim, !obs_id %in% heldout_obs_ids)
```

## Fitting the Training Model & Projecting to Held-Out Locations

Because `imuGAP` embeds the entire tree hierarchy in the location random effects, it can project coverage for any location in `locations_sim` even when that location has zero training observations:

```{r fit-spatial-holdout, eval = FALSE}
fit_school_train <- sampling(
  observations = train_obs,
  populations = train_pops,
  locations = locations_sim,
  stan_opts = stan_options(chains = 4, iter = 1000, refresh = 0)
)
```

You can then create a target projection grid for the held-out school and generate posterior predictive draws using `predict()`:

```{r predict-spatial-holdout-demo, eval = FALSE}
target_school_grid <- create_target(
  location = target_school,
  cohort = 1:5,
  age = 1:5,
  dose = 1:2,
  mode = "enumerate"
)

pred_school <- predict(fit_school_train, target = target_school_grid, posterior_size = 200)
```

## Evaluating Predictive Performance Across All Schools

Repeating this workflow for each of the 10 schools in Scruggs County yields independent out-of-sample projections. The package dataset `leave_school_out_scruggs` contains consolidated results from all 10 folds:

```{r inspect-school-metrics}
data.frame(
  Metric = c(
    "Mean Absolute Error (MAE)",
    "Root Mean Squared Error (RMSE)",
    "95% CI Coverage Rate",
    "Evaluation Points"
  ),
  Value = c(
    sprintf("%.4f", leave_school_out_scruggs$metrics$mae),
    sprintf("%.4f", leave_school_out_scruggs$metrics$rmse),
    sprintf("%.1f%%", 100 * leave_school_out_scruggs$metrics$coverage_95),
    as.character(leave_school_out_scruggs$metrics$n_eval_points)
  )
)
```

We compare the out-of-sample posterior predictive trajectories against the held-out empirical observations across all 10 schools:

```{r plot-spatial-eval, fold = TRUE, code_summary = "Show plot code", fig.cap = "Out-of-sample projections vs. held-out school observations (Dose 2 at Age 5)"}
pred_kdg <- subset(leave_school_out_scruggs$predictions, dose == 2 & age == 5)
eval_kdg <- subset(leave_school_out_scruggs$eval_comparison, dose == 2 & age == 5)

ggplot() +
  geom_ribbon(
    data = pred_kdg,
    aes(x = cohort, ymin = q2_5, ymax = q97_5),
    fill = "#386cb0",
    alpha = 0.25
  ) +
  geom_line(
    data = pred_kdg,
    aes(x = cohort, y = q50),
    color = "#386cb0",
    linewidth = 0.9
  ) +
  geom_point(
    data = eval_kdg,
    aes(x = cohort, y = empirical_coverage),
    color = "#d95f02",
    size = 2.5,
    shape = 17
  ) +
  facet_wrap(~loc_id, ncol = 5) +
  labs(
    x = "Birth Cohort",
    y = "Dose 2 Coverage at Age 5"
  ) +
  coord_cartesian(ylim = c(0.4, 1.0))
```

## Spatial Shrinkage & Hierarchical Pooling

When a school has zero observations, its projection shrinks toward the parent county mean (`Scruggs`) and the broader state-level trend. Comparing out-of-sample predictions against in-sample fitted means illustrates this Bayesian hierarchical shrinkage:

```{r plot-school-shrinkage, fold = TRUE, code_summary = "Show plot code", fig.cap = "In-sample fitted vs. out-of-sample projected school coverage (Dose 2 at Age 5)"}
shrinkage_kdg <- subset(
  leave_school_out_scruggs$shrinkage_comparison,
  dose == 2 & age == 5
)

ggplot(shrinkage_kdg, aes(x = insample_mean, y = oos_mean)) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "grey50") +
  geom_point(aes(color = loc_id), size = 2.8, alpha = 0.9) +
  scale_color_brewer(palette = "Paired") +
  labs(
    x = "In-Sample Fitted Posterior Mean",
    y = "Out-of-Sample Projected Posterior Mean",
    color = "School"
  ) +
  coord_cartesian(xlim = c(0.70, 0.95), ylim = c(0.70, 0.95))
```

## Expected Log Predictive Density: PSIS-LOO vs. Exact Spatial Holdouts

In Section 2.1, the grouped Pareto $\hat{k}$ diagnostics flagged that leaving out entire schools shifts the posterior distribution too substantially for importance sampling to reliably approximate out-of-sample predictive density.

With exact MCMC holdout refits for all 10 schools in Scruggs County, we can evaluate the exact out-of-sample Expected Log Predictive Density under the binomial observation process:

$$\text{elpd}_s^{\text{Exact}} = \log \left( \frac{1}{S} \sum_{t=1}^S \prod_{j \in \text{school } s} \binom{N_j}{y_j} (p_j^{(t)})^{y_j} (1 - p_j^{(t)})^{N_j - y_j} \right)$$

We compare $\widehat{\text{elpd}}_s^{\text{PSIS}}$ (from the full-data model) against $\text{elpd}_s^{\text{Exact}}$ (from the holdout refits):

```{r compute-elpd-comparison}
# Extract PSIS-LOO ELPD and Pareto k for Scruggs County schools
psis_elpd_vec <- school_loo$pointwise[, "elpd_loo"]
pareto_k_vec <- school_loo$diagnostics$pareto_k
names(pareto_k_vec) <- schools$loc_id

scruggs_schools <- leave_school_out_scruggs$schools

# Exact out-of-sample ELPD helper (numerically stable log-sum-exp)
log_sum_exp <- function(x) {
  m <- max(x)
  m + log(sum(exp(x - m)))
}

exact_elpd_vec <- numeric(length(scruggs_schools))
names(exact_elpd_vec) <- scruggs_schools

for (s in scruggs_schools) {
  s_eval <- subset(leave_school_out_scruggs$eval_comparison, loc_id == s)
  s_preds <- subset(leave_school_out_scruggs$predictions, loc_id == s)
  draws_s <- leave_school_out_scruggs$draws[[s]]
  n_draws <- dim(draws_s)[1] * dim(draws_s)[2]

  log_lik_draws <- matrix(0, nrow = n_draws, ncol = nrow(s_eval))
  for (j in seq_len(nrow(s_eval))) {
    c_j <- s_eval$cohort[j]
    a_j <- s_eval$age[j]
    d_j <- s_eval$dose[j]
    y_j <- s_eval$positive[j]
    n_j <- s_eval$sample_n[j]

    t_idx <- which(s_preds$cohort == c_j & s_preds$age == a_j & s_preds$dose == d_j)
    p_draws <- as.vector(draws_s[, , t_idx])
    log_lik_draws[, j] <- dbinom(y_j, size = n_j, prob = p_draws, log = TRUE)
  }

  school_ll_draws <- rowSums(log_lik_draws)
  exact_elpd_vec[s] <- log_sum_exp(school_ll_draws) - log(n_draws)
}

elpd_comparison <- data.frame(
  loc_id = scruggs_schools,
  pareto_k = round(pareto_k_vec[scruggs_schools], 2),
  elpd_psis = round(psis_elpd_vec[scruggs_schools], 2),
  elpd_exact = round(exact_elpd_vec[scruggs_schools], 2),
  delta_elpd = round(psis_elpd_vec[scruggs_schools] - exact_elpd_vec[scruggs_schools], 2)
)
elpd_comparison
```

We visualize the approximation discrepancy $\Delta\text{elpd} = \widehat{\text{elpd}}^{\text{PSIS}} - \text{elpd}^{\text{Exact}}$ as a function of the school Pareto $\hat{k}$ diagnostic:

```{r plot-elpd-comparison, fold = TRUE, code_summary = "Show plot code", fig.cap = "PSIS-LOO ELPD approximation discrepancy vs. school Pareto k diagnostic"}
ggplot(elpd_comparison, aes(x = pareto_k, y = delta_elpd)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  geom_point(aes(color = loc_id), size = 3) +
  scale_color_brewer(palette = "Paired") +
  labs(
    x = expression("School Pareto " * hat(k)),
    y = expression(Delta * "ELPD (PSIS " * hat(elpd) * " - Exact " * elpd * ")"),
    color = "School"
  )
```

Because $\Delta\text{elpd} > 0$ across the schools, importance sampling on the full fit is consistently overly optimistic relative to exact out-of-sample predictive density, confirming why high $\hat{k}$ diagnostics necessitate exact MCMC holdouts.

---

# 3. Surveillance Stream Holdout (Leave-Stream-Out)

Real-world immunization surveillance systems frequently integrate multiple heterogeneous datastreams. In our simulated surveillance environment, observations are organized across four distinct surveillance streams:

1. **ChildVaxView (State)**: State-level early childhood surveys capturing Dose 1 uptake across ages 2 and 3.
2. **SchoolVaxView (State)**: State-level aggregate kindergarten entry surveys capturing Dose 2 uptake at age 5.
3. **TeenVaxView (State)**: State-level adolescent surveys capturing Doses 1 and 2 across ages 14 through 18.
4. **Grade 6 Surveys (County)**: County-level middle school surveys capturing Dose 2 at age 11, reported with right-censoring (`censored == 1`).

*(Individual school-level kindergarten entry surveys across the 24 local schools are evaluated under the grouped spatial cross-validation in Section 2).*

## Approximate Leave-Stream-Out Diagnostics (Grouped PSIS-LOO)

Before performing exact MCMC refits without each stream, we evaluate whether PSIS-LOO can approximate holding out an entire surveillance stream. For each stream $k \in \{1, 2, 3, 4\}$, we sum the pointwise log-likelihood draws across all observations belonging to that stream:

$$\log p(y_{\text{stream } k} \mid \theta^{(s)}) = \sum_{j \in \text{stream } k} \log p(y_j \mid \theta^{(s)})$$

Passing this grouped log-likelihood matrix to `{loo}` computes Pareto $\hat{k}$ diagnostics for each surveillance stream:

```{r run-stream-loo}
# Group pointwise log-likelihood draws across the 4 surveillance streams
stream_cols_list <- list(
  childvaxview = which(obs_meta$loc_id == "State" & obs_meta$age_min <= 3L),
  schoolvaxview = which(obs_meta$loc_id == "State" & obs_meta$age_min == 5L),
  teenvaxview = which(obs_meta$loc_id == "State" & obs_meta$age_min >= 14L),
  grade6 = which(obs_meta$layer == 2L)
)

stream_ll <- do.call(cbind, lapply(stream_cols_list, function(cols) {
  rowSums(ll_mat[, cols, drop = FALSE])
}))

stream_loo <- loo::loo(stream_ll, r_eff = rep(1, ncol(stream_ll)))
print(stream_loo)
```

Surveillance streams that establish foundational baseline propensity (such as `ChildVaxView` and `TeenVaxView`) exhibit high Pareto $\hat{k}$ values ($\hat{k} > 0.7$), indicating that their omission substantially shifts posterior distributions and requires exact MCMC refitting.

## Systematic Leave-Stream-Out Cross-Validation

To evaluate true out-of-sample generalizability across surveillance streams, we systematically hold out each stream independently, refit the model on the remaining streams, and project coverage on the held-out stream coordinates.

For example, to hold out the county-level middle school survey (`grade6`), identify its observation IDs and exclude them from training:

```{r stream-holdout-split}
# Identify and exclude observation IDs belonging to the county Grade 6 stream
counties <- c("Scruggs", "Simone", "Watson")
heldout_stream_obs_ids <- subset(populations_sim, loc_id %in% counties)$obs_id
train_stream_obs <- subset(observations_sim, !obs_id %in% heldout_stream_obs_ids)
train_stream_pops <- subset(populations_sim, !obs_id %in% heldout_stream_obs_ids)
```

The precomputed package dataset `leave_stream_out` consolidates results from all 4 stream holdout fits:

```{r inspect-stream-metrics}
stream_metric_rows <- lapply(names(leave_stream_out$metrics$by_stream), function(st) {
  m <- leave_stream_out$metrics$by_stream[[st]]
  data.frame(
    Stream = st,
    MAE = sprintf("%.4f", m$mae),
    RMSE = sprintf("%.4f", m$rmse),
    Coverage_95 = sprintf("%.1f%%", 100 * m$coverage_95),
    Eval_Points = as.character(m$n_eval_points)
  )
})
do.call(rbind, stream_metric_rows)
```

## Out-of-Sample Predictions vs. Held-Out Stream Observations

We compare the out-of-sample projected coverage against the empirical measurements for each held-out stream:

```{r plot-stream-eval, fold = TRUE, code_summary = "Show plot code", fig.cap = "Out-of-sample projections vs. empirical observations across held-out surveillance streams"}
eval_dt <- copy(leave_stream_out$eval_comparison)
eval_dt[, stream_label := fcase(
  heldout_stream == "childvaxview", "ChildVaxView (Age 2-3, Dose 1)",
  heldout_stream == "schoolvaxview", "SchoolVaxView (Age 5, Dose 2)",
  heldout_stream == "teenvaxview", "TeenVaxView (Age 14-18)",
  heldout_stream == "grade6", "Grade 6 Surveys (Age 11, Dose 2)"
)]

ggplot(eval_dt) +
  geom_point(aes(x = cohort, y = empirical_coverage), color = "#d95f02", size = 2, alpha = 0.8) +
  geom_line(aes(x = cohort, y = q50), color = "#386cb0", linewidth = 0.8) +
  geom_ribbon(aes(x = cohort, ymin = q2_5, ymax = q97_5), fill = "#386cb0", alpha = 0.2) +
  facet_wrap(~stream_label, scales = "free_x") +
  labs(
    x = "Birth Cohort",
    y = "Coverage Proportion"
  ) +
  coord_cartesian(ylim = c(0.4, 1.0))
```

## Expected Log Predictive Density: PSIS-LOO vs. Exact Stream Holdouts

We compare the approximate Expected Log Predictive Density ($\widehat{\text{elpd}}^{\text{PSIS}}$) against the exact out-of-sample log predictive density ($\text{elpd}^{\text{Exact}}$) obtained from exact MCMC stream refits:

```{r compute-stream-elpd-comparison}
leave_stream_out$elpd_comparison
```

The discrepancy $\Delta\text{elpd} = \widehat{\text{elpd}}^{\text{PSIS}} - \text{elpd}^{\text{Exact}} > 0$ confirms that importance sampling without refitting is overly optimistic for major surveillance stream holdouts.

---

# 4. Temporal Forecast Holdouts & Scoring with CRPS

In operational surveillance systems, validating forecasting performance for emerging birth cohorts is critical. The most recent observations across all surveillance streams correspond strictly to the maximum observation calendar year:

$$\text{Observation Year} = \text{cohort} + \text{age} = \max(\text{cohort} + \text{age}) = 33$$

This frontier represents the latest available surveillance wave: cohort 30 for `ChildVaxView` (age 3), cohort 28 for kindergarten entry (age 5), and cohort 22 for middle school surveys (age 11).

## Grouped PSIS-LOO for Recent Observations

We evaluate the grouped PSIS-LOO diagnostic for holding out the entire most recent observation frontier:

```{r run-recent-loo}
max_year <- max(populations_sim$cohort + populations_sim$age)
recent_obs_ids <- subset(populations_sim, cohort + age == max_year)$obs_id
recent_cols <- which(obs_meta$obs_id %in% recent_obs_ids)

recent_ll <- matrix(rowSums(ll_mat[, recent_cols, drop = FALSE]), ncol = 1)
colnames(recent_ll) <- "Frontier_Year_33"

recent_loo <- loo::loo(recent_ll, r_eff = 1)
print(recent_loo)
```

The Pareto $\hat{k} \approx 0.46 < 0.5$ confirms that importance weights for this single-year temporal frontier retain finite variance, though exact MCMC refitting provides full predictive distributions for proper scoring.

## Out-of-Sample Forecasting & CRPS Scoring

The Continuous Ranked Probability Score (CRPS; @gneiting2007) is a strictly proper scoring rule assessing both sharpness and calibration of probabilistic forecasts:

$$\text{CRPS}(F, y) = \int_{-\infty}^{\infty} (F(x) - \mathbb{I}(x \ge y))^2 \, dx$$

We inspect out-of-sample forecast accuracy and CRPS scores from `leave_recent_out`:

```{r inspect-recent-metrics}
data.frame(
  Metric = c(
    "Mean Absolute Error (MAE)",
    "Root Mean Squared Error (RMSE)",
    "95% CI Coverage Rate",
    "Mean CRPS",
    "Evaluation Points"
  ),
  Value = c(
    sprintf("%.4f", leave_recent_out$metrics$mae),
    sprintf("%.4f", leave_recent_out$metrics$rmse),
    sprintf("%.1f%%", 100 * leave_recent_out$metrics$coverage_95),
    sprintf("%.4f", leave_recent_out$metrics$mean_crps),
    as.character(leave_recent_out$metrics$n_eval_points)
  )
)
```

## Longitudinal Calibration Across Recent Cohorts

We visualize the out-of-sample forecast trajectories and prediction intervals against the held-out empirical observations:

```{r plot-recent-calibration, fold = TRUE, code_summary = "Show plot code", fig.cap = "Out-of-sample forecast predictions vs. held-out empirical observations at the temporal frontier"}
eval_recent_dt <- copy(leave_recent_out$eval_comparison)

ggplot(eval_recent_dt, aes(x = cohort)) +
  geom_point(aes(y = empirical_coverage), color = "#d95f02", size = 2.5, shape = 17) +
  geom_point(aes(y = q50), color = "#386cb0", size = 2) +
  geom_errorbar(aes(ymin = q2_5, ymax = q97_5), color = "#386cb0", width = 0.3) +
  facet_wrap(~loc_id, scales = "free_x") +
  labs(
    x = "Birth Cohort",
    y = "Coverage Proportion"
  ) +
  coord_cartesian(ylim = c(0.4, 1.0))
```

## ELPD Forecast Comparison

Comparing the PSIS-LOO approximation with exact temporal refits:

```{r compute-recent-elpd-comparison}
leave_recent_out$elpd_comparison
```

---

# Summary & Best Practices

When designing validation experiments for `imuGAP`:

- **Use PSIS-LOO (`loo`)** for fast, exploratory influence and outlier diagnostics on full fits.
- **Use Leave-Each-School-Out Cross-Validation** to evaluate spatial shrinkage and generalizability for unmeasured schools or communities.
- **Use Leave-Stream-Out Cross-Validation** to verify whether subsets of surveillance streams can accurately identify latent hazards ($\lambda$) and propensity parameters ($\phi$).
- **Use Temporal Horizon Holdouts** scored with **CRPS (`scoringRules`)** to validate prospective forecasting calibration before deploying estimates into policy decisions.

---

# References
