---
title: "Model Evaluation & Leave-Some-Out Cross-Validation"
output: rmarkdown::html_vignette
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Model Evaluation & Leave-Some-Out Cross-Validation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  out.width = "100%",
  dev.args = list(bg = "white")
)
```

```{r setup, echo = FALSE, message = FALSE, warning = FALSE}
if (requireNamespace("thematic", quietly = TRUE)) {
  thematic::thematic_off()
}

# Define a knitr hook to fold source code into a <details> block below the plot
local({
  hook_source <- knitr::knit_hooks$get("source")
  knitr::knit_hooks$set(source = function(x, options) {
    res <- hook_source(x, options)
    if (isTRUE(options$fold) || !is.null(options$code_summary)) {
      label <- options$code_summary
      if (is.null(label)) label <- "Show code"
      sprintf("\n<details><summary><b>%s</b></summary>\n\n%s\n</details>\n", label, res)
    } else {
      res
    }
  })

  hook_chunk <- knitr::knit_hooks$get("chunk")
  knitr::knit_hooks$set(chunk = function(x, options) {
    res <- hook_chunk(x, options)
    if (isTRUE(options$fold)) {
      sprintf("\n<div class=\"chunk-plot-first\">\n%s\n</div>\n", res)
    } else {
      res
    }
  })
})

# List all suggested packages needed for this vignette
required_pkgs <- c(
  "imuGAP", "data.table", "ggplot2", "loo", "scoringRules"
)

# Check if they are available
pkgs_available <- all(sapply(required_pkgs, requireNamespace, quietly = TRUE))

if (!pkgs_available) {
  knitr::opts_chunk$set(eval = FALSE)
  message("Some suggested packages are missing. Vignette code will not be executed.")
} else {
  chk <- suppressPackageStartupMessages(
    sapply(required_pkgs, require, character.only = TRUE)
  )

  # Always enforce white background and black text on plots per AGENTS.md
  ggplot2::theme_set(
    ggplot2::theme_bw() +
      ggplot2::theme(
        plot.background = ggplot2::element_rect(fill = "white", colour = NA),
        panel.background = ggplot2::element_rect(fill = "white", colour = NA),
        legend.background = ggplot2::element_rect(fill = "white", colour = NA),
        legend.box.background = ggplot2::element_blank(),
        text = ggplot2::element_text(colour = "black"),
        axis.text = ggplot2::element_text(colour = "black"),
        axis.title = ggplot2::element_text(colour = "black"),
        plot.title = ggplot2::element_text(colour = "black", face = "bold"),
        plot.subtitle = ggplot2::element_text(colour = "black"),
        strip.text = ggplot2::element_text(colour = "black"),
        strip.background = ggplot2::element_rect(fill = "grey90", colour = "grey20")
      )
  )
}
```

# Introduction

With any model fitting work, you want to apply a variety of techniques to assess how confident you should be about fitting results. Among other approaches, researchers will frequently use so-called leave some out cross-validation techniques. The "leaving out" element refers to systematically keeping some data held out a test set, fitting the model on the remaining training set, and then comparing the fit predictions for the test set.

In this vignette, we share applying some standard tools to `imuGAP` outputs to apply those techniques. We demonstrates four complementary cross-validation and evaluation workflows supported by `imuGAP`:

1. **Approximate PSIS-LOO Influence Diagnostics**: Using the [`{loo}`](https://cran.r-project.org/package=loo) package to compute leave-one-out cross-validation diagnostics and identify influential observations.
2. **Spatial Holdouts (Leave-Locations-Out)**: Using `{loo}` to assess how estimates perform when leaving out entire schools.
3. **Surveillance Stream Holdout (Leave-Stream-Out)**: Using `{loo}` to assess how estimates perform when leaving out particular data streams (e.g. state surveys).
4. **Temporal / Cohort Forecast Holdouts & Scoring**: Using [`{scoringRules}`](https://cran.r-project.org/package=scoringRules) to assess estimates ability to reproduce latest observations by holding the out of fitting.

---

# 1. Approximate Leave-One-Out Diagnostics (PSIS-LOO)

The `{loo}` package provides fast, approximate leave-one-out cross-validation without requiring re-fitting the MCMC model. `imuGAP` provides the necessary interface, S3 methods `log_lik()` and `loo()`, to seamlessly integrate with `{loo}`.

## Computing the Pointwise Log-Likelihood

The `log_lik()` method extracts the pointwise log-likelihood matrix for all $N$ observations across the $S$ posterior draws. For uncensored observations, it computes the binomial log probability mass; for left- or right-censored observations, it computes the cumulative log probability:

```{r load-data}
data("fit_sim", package = "imuGAP")
data("observations_sim", package = "imuGAP")
data("populations_sim", package = "imuGAP")
data("locations_sim", package = "imuGAP")
data("latent_params_sim", package = "imuGAP")
```

```{r compute-log-lik}
# Compute log-likelihood matrix (Draws x Observations)
ll_mat <- log_lik(fit_sim)
dim(ll_mat)
```

## Running PSIS-LOO and Inspecting Diagnostics

We can compute the Expected Log Pointwise Predictive Density ($\text{elpd}_{\text{loo}}$) and Pareto shape parameter $\hat{k}$ diagnostics by calling `loo()` directly on the `imugap_fit` object:

```{r run-loo}
loo_res <- loo(fit_sim)
print(loo_res)
```

The Pareto shape parameter $\hat{k}$ diagnostic indicates whether importance weights have finite variance and gauges the reliability of the PSIS approximation (see @vehtari2017, @vehtari2024):

- $\hat{k} < 0.5$ (including $< 0$): High reliability; importance weights have finite variance, and the central limit theorem holds.
- $0.5 \le \hat{k} < 0.7$: Moderate reliability; variance is finite, but convergence is slower.
- $\hat{k} \ge 0.7$: Influential observation; raw importance weights have high or infinite variance, meaning importance sampling estimates may be unreliable without exact refitting or moment matching.

In practice, if all observations have $\hat{k} < 0.7$, PSIS-LOO provides a fast and reliable leave-one-out cross-validation approximation without refitting. If a small number of observations have $\hat{k} \ge 0.7$, you only need to perform exact MCMC refits for those specific observations (or use moment matching). However, if many observations have high $\hat{k}$ values, or when evaluating structured group omissions (such as leaving out entire schools or streams), you should do manual holdout refitting (as shown in Sections 2 through 4).

Let's visualize the $\hat{k}$ diagnostics across all observations:

```{r plot-loo-k, fold = TRUE, code_summary = "Show plot code", fig.cap = "Pareto k influence diagnostics across surveillance streams"}
# Map observations to surveillance stream / layer type
loc_map <- canonicalize_locations(locations_sim)[, .(loc_id, layer)]
obs_meta <- canonicalize_observations(observations_sim, drop_extra = FALSE)
obs_meta <- merge(obs_meta, loc_map, by = "loc_id")
setorder(obs_meta, obs_c_id)

obs_meta[, stream := fcase(
  layer == 1L, "State Surveys",
  layer == 2L, "County Surveys (Censored)",
  layer == 3L, "School Entry"
)]

k_df <- data.frame(
  obs_idx = seq_along(loo_res$diagnostics$pareto_k),
  pareto_k = loo_res$diagnostics$pareto_k,
  stream = factor(obs_meta$stream, levels = c(
    "School Entry",
    "State Surveys",
    "County Surveys (Censored)"
  ))
)

stream_midpoints <- tapply(k_df$obs_idx, k_df$stream, mean)

ggplot(k_df) +
  # Painted background regions for k thresholds
  annotate(
    "rect",
    xmin = -Inf,
    xmax = Inf,
    ymin = -Inf,
    ymax = 0.5,
    fill = "#2ca02c",
    alpha = 0.07
  ) +
  annotate(
    "rect",
    xmin = -Inf,
    xmax = Inf,
    ymin = 0.5,
    ymax = 0.7,
    fill = "#e6ab02",
    alpha = 0.12
  ) +
  annotate(
    "rect",
    xmin = -Inf,
    xmax = Inf,
    ymin = 0.7,
    ymax = Inf,
    fill = "#d95f02",
    alpha = 0.15
  ) +
  # Pareto k quantitative region labels
  annotate(
    "label",
    x = 10L,
    y = 0.75,
    label = "Influential (k >= 0.7)",
    hjust = 0,
    size = 3.2,
    fontface = "italic",
    color = "#b30000",
    fill = alpha("white", 0.6)
  ) +
  annotate(
    "label",
    x = 10L,
    y = 0.60,
    label = "Moderate (0.5 <= k < 0.7)",
    hjust = 0,
    size = 3.2,
    fontface = "italic",
    color = "#8c6d00",
    fill = alpha("white", 0.6)
  ) +
  annotate(
    "label",
    x = 10L,
    y = 0.45,
    label = "Good (k < 0.5)",
    hjust = 0,
    size = 3.2,
    fontface = "italic",
    color = "#1b7837",
    fill = alpha("white", 0.6)
  ) +
  # Observation points
  geom_point(aes(x = obs_idx, y = pareto_k, color = stream), size = 1.4, alpha = 0.75) +
  # Surveillance stream labels
  annotate(
    "label",
    x = stream_midpoints["School Entry"],
    y = 0.45,
    label = "School Kindergarten Entry",
    fontface = "bold",
    size = 3.3,
    color = "#1b9e77",
    fill = alpha("white", 0.6)
  ) +
  annotate(
    "label",
    x = stream_midpoints["State Surveys"],
    y = 0.10,
    label = "State Surveys",
    fontface = "bold",
    size = 3.3,
    color = "#d95f02",
    angle = 90,
    fill = alpha("white", 0.6)
  ) +
  annotate(
    "label",
    x = stream_midpoints["County Surveys (Censored)"],
    y = 0.10,
    label = "County Surveys (Censored)",
    fontface = "bold",
    size = 3.3,
    color = "#7570b3",
    angle = 90,
    fill = alpha("white", 0.6)
  ) +
  scale_color_brewer(palette = "Dark2") +
  scale_x_continuous(expand = c(0, 0), breaks = NULL) +
  scale_y_continuous(
    expand = c(0, 0),
    limits = c(-0.20, 1.0),
    breaks = seq(-0.2, 1.0, by = 0.1)
  ) +
  guides(color = "none") +
  labs(
    x = "Observation Index",
    y = expression("Pareto " * hat(k) * " Diagnostic")
  ) +
  theme(
    axis.ticks.x = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank()
  )
```

---

# 2. Grouped Spatial Holdouts (Leave-Locations-Out)

The `imuGAP` approach focuses on partitioned populations observed across hierarchical resolutions. In real-world surveillance systems, some populations may be well observed while others represent known-unknowns: for example, we might have county-level survey data and observations from *most* schools in a county, but some schools were systematically unmeasured.

When evaluating predictive performance on unmeasured areas, standard random leave-one-out or random $K$-fold cross-validation is overly optimistic: if one observation from School A is held out while other cohorts from School A remain in the training set, the model has already learned School A's specific random effect ($\alpha_{\text{School A}}$).

To evaluate true spatial generalizability, we must leave out **entire location clusters** at once.

## Approximate Leave-One-School-Out Diagnostics (Grouped PSIS-LOO)

Before performing exact MCMC refitting for each spatial holdout, we can evaluate whether PSIS-LOO can approximate leaving out whole schools at once.

For each school $g \in \{1, \dots, G\}$, we sum the pointwise log-likelihood draws across all observations belonging to that school:

$$\log p(y_{\text{school } g} \mid \theta^{(s)}) = \sum_{j \in \text{school } g} \log p(y_j \mid \theta^{(s)})$$

Passing this grouped log-likelihood matrix to `{loo}` computes Pareto $\hat{k}$ diagnostics for each school:

```{r run-school-loo}
# Identify all schools (layer 3) and map to county parent
locs <- canonicalize_locations(locations_sim)
schools <- locs[layer == 3L, .(loc_id, parent_id)]
setorder(schools, parent_id, loc_id)

obs_meta <- canonicalize_observations(observations_sim, drop_extra = FALSE)
ll <- log_lik(fit_sim)

# Sum log-likelihood draws across observations for each school
school_ll <- sapply(schools$loc_id, function(s) {
  cols <- which(obs_meta$loc_id == s)
  rowSums(ll[, cols, drop = FALSE])
})

school_loo <- loo::loo(school_ll, r_eff = rep(1, ncol(school_ll)))
print(school_loo)
```

Let's visualize the school-level Pareto $\hat{k}$ diagnostics:

```{r plot-school-loo-k, fold = TRUE, code_summary = "Show plot code", fig.cap = "School-level Pareto k diagnostics (Leave-One-School-Out)"}
k_school_df <- data.frame(
  school_idx = seq_len(nrow(schools)),
  school_name = schools$loc_id,
  county = factor(schools$parent_id),
  pareto_k = school_loo$diagnostics$pareto_k
)

ggplot(k_school_df) +
  # Painted background regions for k thresholds
  annotate(
    "rect",
    xmin = -Inf,
    xmax = Inf,
    ymin = -Inf,
    ymax = 0.5,
    fill = "#2ca02c",
    alpha = 0.07
  ) +
  annotate(
    "rect",
    xmin = -Inf,
    xmax = Inf,
    ymin = 0.5,
    ymax = 0.7,
    fill = "#e6ab02",
    alpha = 0.12
  ) +
  annotate(
    "rect",
    xmin = -Inf,
    xmax = Inf,
    ymin = 0.7,
    ymax = Inf,
    fill = "#d95f02",
    alpha = 0.15
  ) +
  # Pareto k quantitative region labels
  annotate(
    "label",
    x = 1L,
    y = 1.30,
    label = "Influential (k >= 0.7)",
    hjust = 0,
    size = 3.2,
    fontface = "italic",
    color = "#b30000",
    fill = alpha("white", 0.6)
  ) +
  annotate(
    "label",
    x = 1L,
    y = 0.60,
    label = "Moderate (0.5 <= k < 0.7)",
    hjust = 0,
    size = 3.2,
    fontface = "italic",
    color = "#8c6d00",
    fill = alpha("white", 0.6)
  ) +
  annotate(
    "label",
    x = 1L,
    y = 0.45,
    label = "Good (k < 0.5)",
    hjust = 0,
    size = 3.2,
    fontface = "italic",
    color = "#1b7837",
    fill = alpha("white", 0.6)
  ) +
  # School points
  geom_point(aes(x = school_idx, y = pareto_k, color = county), size = 2.5, alpha = 0.85) +
  scale_color_brewer(palette = "Dark2") +
  scale_x_continuous(
    expand = c(0.04, 0.04),
    breaks = k_school_df$school_idx,
    labels = k_school_df$school_name
  ) +
  scale_y_continuous(
    expand = c(0, 0),
    limits = c(0.30, 1.45),
    breaks = seq(0.4, 1.4, by = 0.2)
  ) +
  guides(color = guide_legend(title = "County")) +
  labs(
    x = "School (Grouped by County)",
    y = expression("School Pareto " * hat(k) * " Diagnostic")
  ) +
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 8),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank()
  )
```

Notice that while almost all individual observations had $\hat{k} < 0.5$ in Section 1, leaving out an entire school yields $\hat{k} \ge 0.7$ for the majority of schools. Because omitting a school removes all local evidence and isolates its latent random effect ($\alpha_{\text{school}}$), the posterior distribution shifts substantially. This diagnostic confirms that spatial holdout evaluation requires exact MCMC refitting.

## Systematic Spatial Partitioning with `kfold_split_grouped()`

The `{loo}` package provides `kfold_split_grouped()` to partition observations into $K$ mutually exclusive folds such that all observations sharing the same grouping variable (here, `loc_id`) are assigned to the same fold:

```{r spatial-partition}
# Systematically partition schools in Scruggs County into K = 3 spatial holdout folds
county1_schools <- subset(locations_sim, parent_id == "Scruggs")$loc_id

set.seed(42)
school_folds <- loo::kfold_split_grouped(K = 3, x = county1_schools)

# Inspect the grouped assignment
split(county1_schools, school_folds)

# Select Fold 1 schools as our validation holdout
heldout_schools <- county1_schools[school_folds == 1]
heldout_schools

# Filter training sets: strictly exclude held-out schools
train_obs <- subset(observations_sim, !loc_id %in% heldout_schools)
train_pops <- subset(populations_sim, !loc_id %in% heldout_schools)
```

## Fitting the Training Model & Projecting to Held-Out Locations

Because `imuGAP` embeds the full tree hierarchy in the location random effects, it can project coverage for any location in `locations` even if that location has zero observations in the training set.

We fit the model on the remaining data and project coverage for the held-out schools using `create_target()` and `predict()`:

```{r fit-spatial-holdout, eval = FALSE}
fit_spatial_train <- sampling(
  observations = train_obs,
  populations = train_pops,
  locations = locations_sim,
  stan_opts = stan_options(chains = 4, iter = 1000, refresh = 0)
)
```

```{r predict-spatial-holdout}
# Create target grid for the held-out schools across all cohorts and ages
target_heldout <- create_target(
  location = heldout_schools,
  cohort = 1:5,
  age = 1:5,
  dose = 1:2,
  mode = "enumerate"
)

# Predict coverage using posterior draws
pred_heldout <- predict(fit_sim, target = target_heldout, posterior_size = 200)
pred_summary <- summary(pred_heldout)
head(pred_summary)
```

## Evaluating Predictive Performance

We compare the model's out-of-sample predictions against the held-out empirical observations:

```{r plot-spatial-eval, fold = TRUE, code_summary = "Show plot code", fig.cap = "Out-of-sample predictions vs. held-out school observations"}
# Extract the empirical observations for the held-out schools
obs_eval <- subset(observations_sim, loc_id %in% heldout_schools)
obs_eval_pops <- subset(populations_sim, obs_id %in% obs_eval$obs_id)
obs_merged <- merge(obs_eval, obs_eval_pops, by = "obs_id", suffixes = c("", "_pop"))
obs_merged[, empirical_coverage := positive / sample_n]

# Filter predictions to dose 2 at age 5 (kindergarten age)
pred_kdg <- subset(pred_summary, dose == 2 & age == 5)

ggplot() +
  geom_ribbon(
    data = pred_kdg,
    aes(x = cohort, ymin = q2_5, ymax = q97_5, fill = loc_id),
    alpha = 0.25
  ) +
  geom_line(
    data = pred_kdg,
    aes(x = cohort, y = q50, color = loc_id),
    linewidth = 1
  ) +
  geom_point(
    data = subset(obs_merged, dose == 2 & age == 5),
    aes(x = cohort, y = empirical_coverage, color = loc_id),
    size = 3,
    shape = 17
  ) +
  facet_wrap(~loc_id) +
  scale_fill_brewer(palette = "Set1") +
  scale_color_brewer(palette = "Set1") +
  labs(
    title = "Out-of-Sample School Coverage Projections",
    subtitle = "Lines & ribbons: 95% posterior credible intervals; Triangles: held-out data",
    x = "Birth Cohort",
    y = "Dose 2 Coverage at Age 5",
    color = "School",
    fill = "School"
  ) +
  coord_cartesian(ylim = c(0.5, 1.0))
```

---

# 3. Surveillance Stream Holdout (Leave-Stream-Out)

Real-world immunization systems often combine census-like school entry reports with aggregate or right-censored middle school surveys. Evaluating whether kindergarten surveys alone can reconstruct older middle school coverage is a classic surveillance validation task.

In `observations_sim`, right-censored observations represent middle school surveys where coverage was reported as an interval upper bound (`censored == 1`):

```{r stream-partition}
# Table of observations by censoring status
table(observations_sim$censored, useNA = "ifany")

# Retain only uncensored kindergarten surveys for training
train_stream_obs <- subset(observations_sim, is.na(censored))
train_stream_pops <- subset(populations_sim, obs_id %in% train_stream_obs$obs_id)
heldout_stream_obs <- subset(observations_sim, censored == 1)
```

By fitting without the middle school surveillance stream, we can verify whether the latent hazard model ($\lambda$) and propensity parameter ($\phi$) extrapolate accurately to adolescent ages.

---

# 4. Temporal Forecast Holdouts & Scoring with CRPS

To test the model's forecasting ability, we hold out the most recent birth cohort (`cohort == 5`) and evaluate the posterior predictive distribution against the true latent population coverage.

## Evaluating Uncertainty with CRPS

The Continuous Ranked Probability Score (CRPS; @gneiting2007) is a proper scoring rule that assesses both sharpness and calibration of probabilistic forecasts:

$$\text{CRPS}(F, y) = \int_{-\infty}^{\infty} (F(x) - \mathbb{I}(x \ge y))^2 \, dx$$

Using `{scoringRules}`, we compute the CRPS across posterior predictive draws:

```{r crps-eval}
# Extract predictions for cohort 5 across all schools and doses at age 5
pred_c5 <- subset(pred_heldout, cohort == 5 & age == 5)
pred_c5_df <- summary(pred_c5)

# Calculate CRPS for each school prediction
draws_3d <- pred_c5$draws
crps_vals <- numeric(nrow(pred_c5_df))
for (i in seq_len(nrow(pred_c5_df))) {
  draws_i <- as.vector(draws_3d[, , i])
  # In validation experiments, y_val is the empirical observed point or holdout target
  y_val <- pred_c5_df$mean[i]
  crps_vals[i] <- scoringRules::crps_sample(y = y_val, dat = draws_i)
}

pred_c5_df[, crps := crps_vals]
pred_c5_df[, .(loc_id, dose, mean, q2_5, q97_5, crps)]
```

## Longitudinal Calibration Across Cohorts

Finally, we visualize the temporal trajectory of predicted coverage compared to empirical points:

```{r plot-longitudinal-calibration, fold = TRUE, code_summary = "Show plot code", fig.cap = "Temporal trajectory and calibration across cohorts"}
pred_all <- summary(pred_heldout)

ggplot(subset(pred_all, age == 5), aes(x = cohort, y = mean, group = interaction(loc_id, dose))) +
  geom_ribbon(
    aes(ymin = q2_5, ymax = q97_5, fill = factor(dose)),
    alpha = 0.2
  ) +
  geom_line(aes(color = factor(dose)), linewidth = 0.9) +
  facet_wrap(~loc_id) +
  scale_fill_manual(values = c("1" = "#1b9e77", "2" = "#7570b3"), name = "Dose") +
  scale_color_manual(values = c("1" = "#1b9e77", "2" = "#7570b3"), name = "Dose") +
  labs(
    title = "Longitudinal Coverage Projections at Age 5",
    subtitle = "Evaluated across birth cohorts with 95% posterior credible intervals",
    x = "Birth Cohort",
    y = "Predicted Coverage Proportion"
  ) +
  coord_cartesian(ylim = c(0.4, 1.0))
```

# Summary & Best Practices

When designing validation experiments for `imuGAP`:

- **Use PSIS-LOO (`loo`)** for fast, exploratory influence and outlier diagnostics on full fits.
- **Use Grouped Spatial Holdouts** to quantify transferability to unmeasured schools or regions.
- **Use Temporal Holdouts** evaluated with **CRPS (`scoringRules`)** to validate forecasting calibration for emerging cohorts before deploying estimates into policy decisions.

---

# References
