---
title: "Model Evaluation & PSIS-LOO Cross-Validation"
output: rmarkdown::html_vignette
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Model Evaluation & PSIS-LOO Cross-Validation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  out.width = "100%",
  dev.args = list(bg = "white")
)
```

```{r setup, echo = FALSE, message = FALSE, warning = FALSE}
if (requireNamespace("thematic", quietly = TRUE)) {
  thematic::thematic_off()
}

# Define a knitr hook to fold source code into a <details> block below the plot
local({
  hook_source <- knitr::knit_hooks$get("source")
  knitr::knit_hooks$set(source = function(x, options) {
    res <- hook_source(x, options)
    if (isTRUE(options$fold) || !is.null(options$code_summary)) {
      label <- options$code_summary
      if (is.null(label)) label <- "Show code"
      sprintf("\n<details><summary><b>%s</b></summary>\n\n%s\n</details>\n", label, res)
    } else {
      res
    }
  })

  hook_chunk <- knitr::knit_hooks$get("chunk")
  knitr::knit_hooks$set(chunk = function(x, options) {
    res <- hook_chunk(x, options)
    if (isTRUE(options$fold)) {
      sprintf("\n<div class=\"chunk-plot-first\">\n%s\n</div>\n", res)
    } else {
      res
    }
  })
})

# List all suggested packages needed for this vignette
required_pkgs <- c(
  "imuGAP", "data.table", "ggplot2", "loo"
)

# Check if they are available
pkgs_available <- all(sapply(required_pkgs, requireNamespace, quietly = TRUE))

if (!pkgs_available) {
  knitr::opts_chunk$set(eval = FALSE)
  message("Some suggested packages are missing. Vignette code will not be executed.")
} else {
  chk <- suppressPackageStartupMessages(
    sapply(required_pkgs, require, character.only = TRUE)
  )

  # Always enforce white background and black text on plots per AGENTS.md
  ggplot2::theme_set(
    ggplot2::theme_bw() +
      ggplot2::theme(
        plot.background = ggplot2::element_rect(fill = "white", colour = NA),
        panel.background = ggplot2::element_rect(fill = "white", colour = NA),
        legend.background = ggplot2::element_rect(fill = "white", colour = NA),
        legend.box.background = ggplot2::element_blank(),
        text = ggplot2::element_text(colour = "black"),
        axis.text = ggplot2::element_text(colour = "black"),
        axis.title = ggplot2::element_text(colour = "black"),
        plot.title = ggplot2::element_text(colour = "black", face = "bold"),
        plot.subtitle = ggplot2::element_text(colour = "black"),
        strip.text = ggplot2::element_text(colour = "black"),
        strip.background = ggplot2::element_rect(fill = "grey90", colour = "grey20")
      )
  )
}
```

# Introduction

When evaluating model fit, you can use leave-one-out cross-validation (LOO-CV) to estimate out-of-sample predictive performance and identify highly influential observations without having to refit the MCMC model for every observation.

`imuGAP` provides S3 methods for `log_lik()` and `loo()`, enabling integration with the [`{loo}`](https://cran.r-project.org/package=loo) package [@vehtari2017; @vehtari2024] to compute Pareto Smoothed Importance Sampling Leave-One-Out (PSIS-LOO) diagnostics.

---

# Computing the Pointwise Log-Likelihood

The `log_lik()` method extracts the pointwise log-likelihood matrix for all $N$ observations across the $S$ posterior draws. For uncensored observations, it computes the binomial log probability mass; for left- or right-censored observations, it computes the corresponding cumulative log probability:

```{r load-data}
data("fit_sim", package = "imuGAP")
data("observations_sim", package = "imuGAP")
data("locations_sim", package = "imuGAP")
```

```{r compute-log-lik}
# Compute log-likelihood matrix (Draws x Observations)
ll_mat <- log_lik(fit_sim)
dim(ll_mat)
```

---

# Running PSIS-LOO and Inspecting Diagnostics

You can compute the Expected Log Pointwise Predictive Density ($\text{elpd}_{\text{loo}}$) and Pareto shape parameter $\hat{k}$ diagnostics by calling `loo()` directly on an `imugap_fit` object:

```{r run-loo}
loo_res <- loo(fit_sim)
print(loo_res)
```

The Pareto shape parameter $\hat{k}$ diagnostic indicates whether importance weights have finite variance and gauges the reliability of the PSIS approximation [@vehtari2017; @vehtari2024]:

- $\hat{k} < 0.5$ (including $< 0$): High reliability; importance weights have finite variance, and the central limit theorem holds.
- $0.5 \le \hat{k} < 0.7$: Moderate reliability; variance is finite, but convergence is slower.
- $\hat{k} \ge 0.7$: Influential observation; raw importance weights have high or infinite variance, meaning importance sampling estimates may be unreliable without exact refitting or moment matching.

In practice, if all observations have $\hat{k} < 0.7$, PSIS-LOO provides a fast and reliable leave-one-out cross-validation approximation without refitting.

---

# Visualizing Pareto $\hat{k}$ Diagnostics

You can visualize the $\hat{k}$ diagnostics across all observations to check whether any specific surveillance streams or data sources exhibit high influence:

```{r plot-loo-k, fold = TRUE, code_summary = "Show plot code", fig.cap = "Pareto k influence diagnostics across surveillance streams"}
# Map observations to surveillance stream / layer type
loc_map <- canonicalize_locations(locations_sim)[, .(loc_id, layer)]
obs_meta <- canonicalize_observations(observations_sim, drop_extra = FALSE)
obs_meta <- merge(obs_meta, loc_map, by = "loc_id")
data.table::setorder(obs_meta, obs_c_id)

obs_meta[, stream := data.table::fcase(
  layer == 1L, "State Surveys",
  layer == 2L, "County Surveys (Censored)",
  layer == 3L, "School Entry"
)]

k_df <- data.frame(
  obs_idx = seq_along(loo_res$diagnostics$pareto_k),
  pareto_k = loo_res$diagnostics$pareto_k,
  stream = factor(obs_meta$stream, levels = c(
    "School Entry",
    "State Surveys",
    "County Surveys (Censored)"
  ))
)

ggplot(k_df, aes(x = obs_idx, y = pareto_k, color = stream)) +
  # Painted background regions for k thresholds
  annotate(
    "rect",
    xmin = -Inf,
    xmax = Inf,
    ymin = -Inf,
    ymax = 0.5,
    fill = "#2ca02c",
    alpha = 0.07
  ) +
  annotate(
    "rect",
    xmin = -Inf,
    xmax = Inf,
    ymin = 0.5,
    ymax = 0.7,
    fill = "#e6ab02",
    alpha = 0.12
  ) +
  annotate(
    "rect",
    xmin = -Inf,
    xmax = Inf,
    ymin = 0.7,
    ymax = Inf,
    fill = "#d95f02",
    alpha = 0.15
  ) +
  # Pareto k quantitative region labels
  annotate(
    "label",
    x = 10L,
    y = 0.75,
    label = "Influential (k >= 0.7)",
    hjust = 0,
    size = 3.2,
    fontface = "italic",
    color = "#b30000",
    fill = ggplot2::alpha("white", 0.6)
  ) +
  annotate(
    "label",
    x = 10L,
    y = 0.60,
    label = "Moderate (0.5 <= k < 0.7)",
    hjust = 0,
    size = 3.2,
    fontface = "italic",
    color = "#b37400",
    fill = ggplot2::alpha("white", 0.6)
  ) +
  annotate(
    "label",
    x = 10L,
    y = 0.40,
    label = "Reliable (k < 0.5)",
    hjust = 0,
    size = 3.2,
    fontface = "italic",
    color = "#006600",
    fill = ggplot2::alpha("white", 0.6)
  ) +
  geom_hline(yintercept = c(0.5, 0.7), linetype = "dashed", color = "grey40") +
  geom_point(alpha = 0.8, size = 2) +
  coord_cartesian(ylim = c(min(k_df$pareto_k, 0) - 0.05, max(k_df$pareto_k, 0.8) + 0.05)) +
  scale_color_brewer(palette = "Set2") +
  labs(
    title = "PSIS-LOO Pareto k Diagnostics",
    subtitle = "Observation influence diagnostic by surveillance stream",
    x = "Observation Index (Canonical ID)",
    y = "Pareto Shape Parameter (k)",
    color = "Surveillance Stream"
  ) +
  theme(legend.position = "bottom")
```

---

# References
