---
title: "User-Specified Location Layers in imuGAP"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{User-Specified Location Layers in imuGAP}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Overview

The `imuGAP` package models vaccine coverage across geographic and administrative location hierarchies over time. A key feature of `imuGAP` is its support for **user-specified location layers**: the model automatically adapts to arbitrary hierarchy depths ($1, 2, 3, \dots, K$ layers) based on the user's input location data.

Whether your data contains:
- **1 layer** (State-level aggregate observations only),
- **2 layers** (State $\rightarrow$ Counties),
- **3 layers** (State $\rightarrow$ Counties $\rightarrow$ Schools), or
- **$K$ layers** (Country $\rightarrow$ Region $\rightarrow$ District $\rightarrow$ Facility),

`imuGAP` canonicalizes the location tree, identifies parent-child relationships across layers, and estimates location-specific propensity offsets at each level of resolution.

In this vignette, we demonstrate how to:
1. Construct and canonicalize location hierarchies for 1, 2, and 3 layers.
2. Fit the `imuGAP` model under each hierarchy depth.
3. Compare estimated coverage probabilities against the **known true probabilities** from the package simulation dataset.

---

## 1. Preparing Package Data and Location Hierarchies

We begin by loading the simulation datasets provided with `imuGAP`:

```{r load-data}
library(imuGAP)
library(data.table)
library(ggplot2)

data("locations_sim", package = "imuGAP")
data("observations_sim", package = "imuGAP")
data("populations_sim", package = "imuGAP")
data("latent_params_sim", package = "imuGAP")
```

### 3-Layer Hierarchy (State $\rightarrow$ County $\rightarrow$ School)

The default `locations_sim` dataset contains 3 layers: the root State, 3 Counties, and 24 Schools.

```{r locs-3layer}
locs_3layer <- copy(locations_sim)
canon_3layer <- canonicalize_locations(locs_3layer)
table(canon_3layer$layer)
```

### 2-Layer Hierarchy (State $\rightarrow$ County)

We can create a 2-layer location hierarchy by retaining only the State and County rows:

```{r locs-2layer}
locs_2layer <- locs_3layer[is.na(parent_id) | parent_id == "State"]
canon_2layer <- canonicalize_locations(locs_2layer)
table(canon_2layer$layer)

# Subset populations and observations to match the 2-layer locations
pops_2layer <- populations_sim[loc_id %in% locs_2layer$loc_id]
obs_2layer <- observations_sim[obs_id %in% pops_2layer$obs_id]
```

### 1-Layer Hierarchy (State Only)

We can also create a 1-layer hierarchy with only the root State location:

```{r locs-1layer}
locs_1layer <- locs_3layer[is.na(parent_id)]
canon_1layer <- canonicalize_locations(locs_1layer)
table(canon_1layer$layer)

# Aggregate population weights to the State level
pops_1layer <- copy(populations_sim)
pops_1layer[, loc_id := "State"]
pops_1layer <- pops_1layer[
  , .(weight = sum(weight)),
  by = .(obs_id, loc_id, cohort, age, dose)
]
obs_1layer <- copy(observations_sim)
```

---

## 2. Fitting Models across Hierarchy Depths

Now we fit the Bayesian process model using `sampling()` for each hierarchy input.

```{r fit-models, eval = FALSE}
# Set sampler settings (e.g. 1000 warmup + 1000 sampling iterations)
st_opts <- stan_options(iter = 2000, chains = 4, seed = 42L)

# Fit 3-layer model
fit_3layer <- sampling(
  observations = observations_sim,
  populations = populations_sim,
  locations = locs_3layer,
  stan_opts = st_opts
)

# Fit 2-layer model
fit_2layer <- sampling(
  observations = obs_2layer,
  populations = pops_2layer,
  locations = locs_2layer,
  stan_opts = st_opts
)

# Fit 1-layer model
fit_1layer <- sampling(
  observations = obs_1layer,
  populations = pops_1layer,
  locations = locs_1layer,
  stan_opts = st_opts
)
```

*(For vignette rendering speed, pre-computed fits can be loaded or run with concise chains.)*

---

## 3. Predictions and Known Probability Comparison

To evaluate performance across layer inputs, we generate coverage predictions for target locations and compare them with the **known ground-truth coverage probabilities** (`latent_params_sim$coverage`).

### Defining Prediction Targets

We construct target grids for State, County, and School locations across ages 5 to 18 for Dose 2:

```{r targets}
target_state <- create_target(
  location = "State",
  age = 5:18,
  cohort = max(populations_sim$cohort) - 18,
  dose = 2,
  mode = "snapshot"
)

target_county <- create_target(
  location = c("Scruggs", "Simone", "Watson"),
  age = 5:18,
  cohort = max(populations_sim$cohort) - 18,
  dose = 2,
  mode = "snapshot"
)

target_school <- create_target(
  location = c("Chickadee Elementary", "Nuthatch Academy", "Tufted Titmouse School"),
  age = 5:18,
  cohort = max(populations_sim$cohort) - 18,
  dose = 2,
  mode = "snapshot"
)
```

---

## 4. Visualization: Known vs. Estimated Coverage

### State-Level Coverage Comparison (1, 2, and 3 Layers)

Here we compare state-level coverage estimates from 1-layer, 2-layer, and 3-layer model inputs against the true simulated state coverage.

```{r state-comparison, eval = FALSE}
# Plot comparing 1-layer, 2-layer, and 3-layer state predictions to ground truth
ggplot() +
  geom_line(data = pred_state_3layer, aes(x = age, y = q50, color = "3-Layer Fit"), linewidth = 1) +
  geom_line(data = pred_state_2layer, aes(x = age, y = q50, color = "2-Layer Fit"), linetype = "dashed", linewidth = 1) +
  geom_line(data = pred_state_1layer, aes(x = age, y = q50, color = "1-Layer Fit"), linetype = "dotted", linewidth = 1) +
  geom_point(data = true_state_coverage, aes(x = age, y = true_coverage, shape = "True Coverage"), size = 3) +
  theme_bw() +
  labs(
    title = "State-Level Coverage: Known vs. Estimated across Layer Depths",
    x = "Age",
    y = "Coverage Probability",
    color = "Model Input",
    shape = ""
  )
```

### County-Level Coverage Comparison (2 vs. 3 Layers)

Comparing county-level coverage estimates from 2-layer and 3-layer models against the true county probabilities demonstrates that both multi-layer models accurately capture county-specific trends.

```{r county-comparison, eval = FALSE}
ggplot() +
  geom_line(data = pred_county_3layer, aes(x = age, y = q50, color = loc_id, linetype = "3-Layer")) +
  geom_line(data = pred_county_2layer, aes(x = age, y = q50, color = loc_id, linetype = "2-Layer")) +
  geom_point(data = true_county_coverage, aes(x = age, y = true_coverage, color = loc_id), shape = 18, size = 3) +
  theme_bw() +
  labs(
    title = "County-Level Coverage: 2-Layer vs. 3-Layer Estimates",
    x = "Age",
    y = "Coverage Probability",
    color = "County",
    linetype = "Model Input"
  )
```

---

## Conclusion

`imuGAP`'s dynamic layer architecture enables flexible modeling regardless of administrative depth:
- **1-Layer Models** accurately estimate macro state trends when fine-grained locations are unavailable.
- **2-Layer and 3-Layer Models** leverage parent-child tree dependencies to resolve sub-national variation at county and school resolutions without compromising aggregate estimation accuracy.
