---
title: "User-Specified Location Layers in imuGAP"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{User-Specified Location Layers in imuGAP}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Overview

The `imuGAP` package models vaccine coverage across geographic and administrative location hierarchies over time. A key strength of `imuGAP` is its support for **user-specified location layers**: the model dynamically adapts to arbitrary hierarchy depths ($1, 2, 3, \dots, K$ layers) based on user input.

Whether your input data contains:
- **1 layer** (State-level aggregate observations only),
- **2 layers** (State $\rightarrow$ Counties),
- **3 layers** (State $\rightarrow$ Counties $\rightarrow$ Schools), or
- **$K$ layers** (Country $\rightarrow$ Region $\rightarrow$ District $\rightarrow$ Facility),

`imuGAP` canonicalizes the location tree, maps parent-child relationships across layers, and estimates location-specific propensity offsets at each resolution level.

This vignette demonstrates:
1. Constructing and canonicalizing location hierarchies for 1, 2, and 3 layers.
2. Fitting the `imuGAP` model under each hierarchy configuration.
3. Loading pre-computed package fit objects (`fit_sim_1layer`, `fit_sim_2layer`, `fit_sim`).
4. Comparing estimated coverage probabilities against the **known true coverage probabilities** from the simulation dataset.

---

## 1. Input Datasets & Location Hierarchy Canonicalization

We load the package simulation datasets:

```{r load-packages}
library(imuGAP)
library(data.table)
library(ggplot2)

data("locations_sim", package = "imuGAP")
data("observations_sim", package = "imuGAP")
data("populations_sim", package = "imuGAP")
data("latent_params_sim", package = "imuGAP")
```

### 3-Layer Hierarchy (State $\rightarrow$ County $\rightarrow$ School)

The default `locations_sim` dataset contains a 3-layer tree: State (root), 3 Counties, and 24 Schools.

```{r locs-3layer}
locs_3layer <- copy(locations_sim)
canon_3layer <- canonicalize_locations(locs_3layer)
table(canon_3layer$layer)
```

### 2-Layer Hierarchy (State $\rightarrow$ County)

Filtering to State and County locations creates a 2-layer hierarchy:

```{r locs-2layer}
locs_2layer <- locs_3layer[is.na(parent_id) | parent_id == "State"]
canon_2layer <- canonicalize_locations(locs_2layer)
table(canon_2layer$layer)
```

### 1-Layer Hierarchy (State Only)

Filtering to only the root State location creates a 1-layer hierarchy:

```{r locs-1layer}
locs_1layer <- locs_3layer[is.na(parent_id)]
canon_1layer <- canonicalize_locations(locs_1layer)
table(canon_1layer$layer)
```

---

## 2. Fitting the Model

Fitting models for 1-layer, 2-layer, and 3-layer location inputs uses `sampling()` with standard Stan options:

```{r fit-code, eval = FALSE}
st_opts <- stan_options(iter = 1000, chains = 4, seed = 1L)

# 3-Layer Fit
fit_3layer <- sampling(observations_sim, populations_sim, locs_3layer, stan_opts = st_opts)

# 2-Layer Fit
pops_2layer <- populations_sim[loc_id %in% locs_2layer$loc_id]
obs_2layer <- observations_sim[obs_id %in% pops_2layer$obs_id]
fit_2layer <- sampling(obs_2layer, pops_2layer, locs_2layer, stan_opts = st_opts)

# 1-Layer Fit
pops_1layer <- copy(populations_sim)[, loc_id := "State"]
pops_1layer <- pops_1layer[, .(weight = sum(weight)), by = .(obs_id, loc_id, cohort, age, dose)]
fit_1layer <- sampling(observations_sim, pops_1layer, locs_1layer, stan_opts = st_opts)
```

For vignettes and local analysis, pre-computed fit artifacts are bundled with `imuGAP` package data:

```{r load-fits}
data("fit_sim", package = "imuGAP")
data("fit_sim_2layer", package = "imuGAP")
data("fit_sim_1layer", package = "imuGAP")
```

---

## 3. Pre-Computed Predictions & Known True Coverage

We load the pre-computed predictions for 1-layer, 2-layer, and 3-layer fits (`predict_sim_1layer`, `predict_sim_2layer`, `predict_sim`):

```{r load-predictions}
data("predict_sim", package = "imuGAP")
data("predict_sim_2layer", package = "imuGAP")
data("predict_sim_1layer", package = "imuGAP")
```

We summarize the predictions using `summary()` to obtain posterior median (`q50`) and 95% credible intervals (`q2_5`, `q97_5`):

```{r summarize}
summary_3layer <- summary(predict_sim)
summary_2layer <- summary(predict_sim_2layer)
summary_1layer <- summary(predict_sim_1layer)
```

---

## 4. Comparing Known vs. Estimated Coverage Probabilities

### State-Level Coverage (1, 2, and 3 Layers)

We compare estimated state-level Dose 2 coverage from the 1-layer, 2-layer, and 3-layer model fits against the **known true coverage** stored in `latent_params_sim$coverage`.

```{r state-prep}
# Extract true coverage matching target
state_3layer_pred <- summary_3layer[loc_id == "State" & dose == 2 & age > 4]
state_2layer_pred <- summary_2layer[loc_id == "State" & dose == 2 & age > 4]
state_1layer_pred <- summary_1layer[loc_id == "State" & dose == 2 & age > 4]

# True ground-truth coverage from latent_params_sim
target_state_rows <- predict_sim$target[loc_id == "State" & dose == 2 & age > 4]
true_state_df <- data.table(
  age = target_state_rows$age,
  coverage = latent_params_sim$coverage[target_state_rows$obs_c_id]
)
```

```{r state-plot, fig.cap = "State-level coverage comparison: estimated vs. known true coverage across layer depths."}
ggplot() +
  geom_ribbon(data = state_3layer_pred, aes(x = age, ymin = q2_5, ymax = q97_5, fill = "3-Layer (State/County/School)"), alpha = 0.2) +
  geom_line(data = state_3layer_pred, aes(x = age, y = q50, color = "3-Layer (State/County/School)"), linewidth = 1) +
  geom_line(data = state_2layer_pred, aes(x = age, y = q50, color = "2-Layer (State/County)"), linetype = "dashed", linewidth = 1) +
  geom_line(data = state_1layer_pred, aes(x = age, y = q50, color = "1-Layer (State Only)"), linetype = "dotted", linewidth = 1) +
  geom_point(data = true_state_df, aes(x = age, y = coverage, shape = "True Coverage"), size = 2.5) +
  scale_shape_manual(name = "", values = c("True Coverage" = 17)) +
  scale_color_manual(name = "Model Input", values = c(
    "3-Layer (State/County/School)" = "#1b9e77",
    "2-Layer (State/County)" = "#d95f02",
    "1-Layer (State Only)" = "#7570b3"
  )) +
  scale_fill_manual(name = "Model Input", values = c(
    "3-Layer (State/County/School)" = "#1b9e77"
  )) +
  theme_bw() +
  scale_x_continuous(breaks = 5:18, minor_breaks = NULL) +
  labs(
    title = "State-Level Coverage: Known vs. Estimated",
    subtitle = "Comparing estimates from 1-layer, 2-layer, and 3-layer location inputs",
    x = "Age",
    y = "Coverage Probability"
  )
```

### County-Level Coverage (2 vs. 3 Layers)

Next, we compare county-level coverage estimates between the 2-layer and 3-layer fits for Counties ("Scruggs", "Simone", "Watson"):

```{r county-plot, fig.cap = "County-level coverage comparison: 2-layer vs. 3-layer estimates against true values."}
county_3layer_pred <- summary_3layer[loc_id %in% c("Scruggs", "Simone", "Watson") & dose == 2 & age > 4]
county_2layer_pred <- summary_2layer[loc_id %in% c("Scruggs", "Simone", "Watson") & dose == 2 & age > 4]

target_county_rows <- predict_sim$target[loc_id %in% c("Scruggs", "Simone", "Watson") & dose == 2 & age > 4]
true_county_df <- data.table(
  loc_id = target_county_rows$loc_id,
  age = target_county_rows$age,
  coverage = latent_params_sim$coverage[target_county_rows$obs_c_id]
)

ggplot() +
  geom_line(data = county_3layer_pred, aes(x = age, y = q50, color = loc_id, linetype = "3-Layer"), linewidth = 0.9) +
  geom_line(data = county_2layer_pred, aes(x = age, y = q50, color = loc_id, linetype = "2-Layer"), linewidth = 0.9) +
  geom_point(data = true_county_df, aes(x = age, y = coverage, color = loc_id, shape = "True Coverage"), size = 2.5) +
  scale_shape_manual(name = "", values = c("True Coverage" = 18)) +
  scale_linetype_manual(name = "Model Hierarchy", values = c("3-Layer" = "solid", "2-Layer" = "dashed")) +
  theme_bw() +
  scale_x_continuous(breaks = 5:18, minor_breaks = NULL) +
  labs(
    title = "County-Level Coverage: 2-Layer vs. 3-Layer Estimates",
    x = "Age",
    y = "Coverage Probability",
    color = "County"
  )
```

---

## Conclusion

The `imuGAP` package's location layer architecture allows users to fit models using whatever administrative resolution is available in their input data ($1, 2, 3, \dots, K$ layers). 
- Macro state-level trends are reliably estimated even under a **1-layer** input.
- Sub-national variation at county (**2-layer**) and school (**3-layer**) resolutions is captured accurately by building hierarchical random offsets down the location tree.
