---
title: "Flexible Location Layers in imuGAP"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Flexible Location Layers in imuGAP}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
library(imuGAP)
library(data.table)
library(ggplot2)
library(knitr)
library(kableExtra)
```

## Overview

The `imuGAP` package models vaccine coverage on nested population partitions - e.g. a state divided into counties divided into schools. A key strength of `imuGAP` is support for **flexible location layers**: you can change the layer depth based on the location hierarchy you provide. So, for example if you provide only the most aggregated population (e.g. state-wide data), the model will have 1 layer, but if you provide sub-populations (e.g. counties) there will be 2 layers. If you also provide sub-sub-populations (e.g. schools), there will be 3 layers, and so on.

Whatever the data resolution, `imuGAP` canonicalizes the location tree, maps parent-child relationships across layers, and estimates location offset parameters for each resolution level (beyond fully aggregated populations).

This vignette demonstrates:

1. Constructing and canonicalizing location hierarchies for different population structure resolutions, using examples from 1 to 3 layers.
2. Fitting the `imuGAP` model under each hierarchy configuration, using the package example data.
3. Comparing estimations across those different resolutions to the known probabilities from the simulation dataset.

## Location Hierarchy

Let's examine the example location dataset; you can load it from the package or consult this table:

```{r load-locations}
data("locations_sim", package = "imuGAP") # if you want to examine it directly
```

```{r display-hierarchy-table, echo = FALSE}
state_name <- locations_sim[is.na(parent_id), loc_id]
counties <- locations_sim[parent_id == state_name, loc_id]

schools_by_county <- lapply(counties, function(cnt) locations_sim[parent_id == cnt, loc_id])
max_len <- max(lengths(schools_by_county))
pad <- function(x, n) c(x, rep("", n - length(x)))

horiz_df <- as.data.frame(
  lapply(schools_by_county, pad, n = max_len),
  col.names = counties,
  check.names = FALSE
)

header_state <- setNames(length(counties), state_name)

kable(horiz_df, col.names = paste(counties, "County"), booktabs = TRUE) |>
  add_header_above(header_state) |>
  kable_styling(latex_options = "HOLD_position", font_size = 9)
```

Most of our examples focus on using the higher resolution dataset (i.e. including 3 layers, or down to schools). For this demonstration, let's create some alternative `location` data corresponding to different data availability. Here, we're using the convenient layer information created by `canonicalize_locations`, but in real applications you would of course only have the locations you have. These filtered views simulate having high to low resolution data, correspond to all the locations in the table (`locs_3layer`), the state and county locations (`locs_2layer`), or just the state (`locs_1layer`).

```{r locs-nlayer}
locs_3layer <- canonicalize_locations(locations_sim) # all the data
locs_2layer <- locs_3layer[layer <= 2] # state and county only
locs_1layer <- locs_3layer[layer <= 1] # state only
```

Similarly, we filter our observations and associated metadata (i.e. populations) to match each location resolution:

```{r obs-nlayer}
data("observations_sim", package = "imuGAP")
data("populations_sim", package = "imuGAP")
data("latent_params_sim", package = "imuGAP")

# 3-Layer views (State, County, School)
# 1. Filter populations to locations present in locs_3layer
pops_3layer <- populations_sim[loc_id %in% locs_3layer$loc_id]
# 2. Extract relevant observations corresponding to filtered populations
obs_3layer <- observations_sim[obs_id %in% pops_3layer$obs_id]
# 3. Confirm extracted observations don't have unpresent locations in original population data
stopifnot(!populations_sim[obs_id %in% obs_3layer$obs_id, any(!loc_id %in% locs_3layer$loc_id)])

# 2-Layer views (State and County)
# 1. Filter populations to locations present in locs_2layer
pops_2layer <- populations_sim[loc_id %in% locs_2layer$loc_id]
# 2. Extract relevant observations corresponding to filtered populations
obs_2layer <- observations_sim[obs_id %in% pops_2layer$obs_id]
# 3. Confirm extracted observations don't have unpresent locations in original population data
stopifnot(!populations_sim[obs_id %in% obs_2layer$obs_id, any(!loc_id %in% locs_2layer$loc_id)])

# 1-Layer views (State only)
# 1. Aggregate and filter populations to State level (locs_1layer)
pops_1layer <- copy(populations_sim)[, loc_id := "State"]
pops_1layer <- pops_1layer[, .(weight = sum(weight)), by = .(obs_id, loc_id, cohort, age, dose)]
pops_1layer <- pops_1layer[loc_id %in% locs_1layer$loc_id]
# 2. Extract relevant observations corresponding to filtered populations
obs_1layer <- observations_sim[obs_id %in% pops_1layer$obs_id]
# 3. Confirm extracted observations don't have unpresent locations in population data
stopifnot(!pops_1layer[obs_id %in% obs_1layer$obs_id, any(!loc_id %in% locs_1layer$loc_id)])
```

---

## 2. Fitting the Model

To fit the model at different resolutions looks basically the same: just use 1-layer, 2-layer, or 3-layer location inputs. Note that no other setting tweaks are required, but these can take some time to run; if following along, you probably want to load the results from the package data.

```{r fit-code, eval = FALSE}
st_opts <- stan_options(iter = 1000, chains = 4, seed = 1L)

# various layer fits
fit_3layer <- sampling(obs_3layer, pops_3layer, locs_3layer, stan_opts = st_opts)
fit_2layer <- sampling(obs_2layer, pops_2layer, locs_2layer, stan_opts = st_opts)
fit_1layer <- sampling(obs_1layer, pops_1layer, locs_1layer, stan_opts = st_opts)
```

To load precomputed results, get the following items from package data:

```{r load-fits}
data("fit_sim", package = "imuGAP")
data("fit_sim_2layer", package = "imuGAP")
data("fit_sim_1layer", package = "imuGAP")
```

## 3. Using the Fits to Predict Coverage

Predicting coverage from fitted models involves providing a target grid for the desired locations, ages, cohorts, and doses, and then calling `predict()`. We can load the example target dataset bundled with `imuGAP` package data and filter it to match each location resolution:

```{r predict-code, eval = FALSE}
data("target_sim", package = "imuGAP")

# 3-Layer Prediction (State, County, School)
target_3layer <- target_sim[loc_id %in% locs_3layer$loc_id]
predict_3layer <- predict(object = fit_3layer, target = target_3layer, posterior_size = 100)

# 2-Layer Prediction (State and County)
target_2layer <- target_sim[loc_id %in% locs_2layer$loc_id]
predict_2layer <- predict(object = fit_2layer, target = target_2layer, posterior_size = 100)

# 1-Layer Prediction (State only)
target_1layer <- target_sim[loc_id %in% locs_1layer$loc_id]
predict_1layer <- predict(object = fit_1layer, target = target_1layer, posterior_size = 100)
```

If you are working along through this vignette, you may wish to simply load the pre-computed prediction results, as we bundled them with `imuGAP` package data:

```{r load-predictions}
data("predict_sim", package = "imuGAP")
data("predict_sim_2layer", package = "imuGAP")
data("predict_sim_1layer", package = "imuGAP")
```

You can summarize the predictions using `summary()` to obtain posterior median (`q50`) and 95% credible intervals (`q2_5`, `q97_5`):

```{r summarize}
summary_3layer <- summary(predict_sim)
summary_2layer <- summary(predict_sim_2layer)
summary_1layer <- summary(predict_sim_1layer)
```

---

## 4. Comparing Different Resolution Fits

As seen in previous sections, the actual fitting and prediction steps are identical irrespective of data resolution. Let's see how the different resolutions affect parameter and coverage estimation.

First, compare estimated State-level coverage from the 1-layer, 2-layer, and 3-layer model fits against the latent coverage stored in `latent_params_sim$coverage`. Faceting by data resolution (columns) shows that macro State trends are estimated accurately regardless of whether lower-level sub-population data are provided:

```{r state-plot, echo = FALSE, fig.cap = "State-level coverage comparison across 1-layer, 2-layer, and 3-layer model fits against true values.", fig.width = 8, fig.height = 4}
state_name <- locations_sim[is.na(parent_id), loc_id]

state_1layer_pred <- summary_1layer[
  loc_id == state_name & dose == 2 & age > 4
][, hierarchy := "1-Layer (State Only)"]
state_2layer_pred <- summary_2layer[
  loc_id == state_name & dose == 2 & age > 4
][, hierarchy := "2-Layer (State/County)"]
state_3layer_pred <- summary_3layer[
  loc_id == state_name & dose == 2 & age > 4
][, hierarchy := "3-Layer (State/County/School)"]

state_cov_pred <- rbindlist(list(state_1layer_pred, state_2layer_pred, state_3layer_pred))
state_cov_pred[, hierarchy := factor(hierarchy, levels = c(
  "1-Layer (State Only)",
  "2-Layer (State/County)",
  "3-Layer (State/County/School)"
))]

target_state_rows <- predict_sim$target[loc_id == state_name & dose == 2 & age > 4]
true_state_df <- data.table(
  age = target_state_rows$age,
  coverage = latent_params_sim$coverage[target_state_rows$obs_c_id]
)

h_levels_3 <- c(
  "1-Layer (State Only)",
  "2-Layer (State/County)",
  "3-Layer (State/County/School)"
)
true_state_faceted <- rbindlist(lapply(h_levels_3, function(h) {
  df <- copy(true_state_df)
  df[, hierarchy := factor(h, levels = h_levels_3)]
  df
}))

ggplot() +
  geom_ribbon(
    data = state_cov_pred,
    aes(x = age, ymin = q2_5, ymax = q97_5, fill = "Estimated (95% CI)"),
    alpha = 0.2
  ) +
  geom_line(
    data = state_cov_pred,
    aes(x = age, y = q50, color = "Estimated (Median)"),
    linewidth = 1
  ) +
  geom_point(
    data = true_state_faceted,
    aes(x = age, y = coverage, shape = "True Coverage"),
    color = "black",
    size = 2.2
  ) +
  facet_grid(. ~ hierarchy) +
  scale_shape_manual(name = "", values = c("True Coverage" = 17)) +
  scale_color_manual(name = "", values = c("Estimated (Median)" = "#1b9e77")) +
  scale_fill_manual(name = "", values = c("Estimated (95% CI)" = "#1b9e77")) +
  theme_bw() +
  theme(legend.position = "bottom") +
  scale_x_continuous(breaks = seq(5, 18, by = 3), minor_breaks = NULL) +
  labs(
    title = "State-Level Coverage Across Data Resolutions",
    x = "Age",
    y = "Coverage Probability"
  )
```

### County-Level Coverage Across Data Resolutions

Next, we compare County-level coverage estimates between the 2-layer and 3-layer model fits across counties (*Scruggs*, *Simone*, *Watson*):

```{r county-plot, echo = FALSE, fig.cap = "County-level coverage comparison: 2-layer vs. 3-layer model estimates against true values.", fig.width = 8, fig.height = 4}
counties <- locations_sim[parent_id == state_name, loc_id]

county_2layer_pred <- summary_2layer[
  loc_id %in% counties & dose == 2 & age > 4
][, hierarchy := "2-Layer (State/County)"]
county_3layer_pred <- summary_3layer[
  loc_id %in% counties & dose == 2 & age > 4
][, hierarchy := "3-Layer (State/County/School)"]

county_cov_pred <- rbindlist(list(county_2layer_pred, county_3layer_pred))
h_levels_county <- c("2-Layer (State/County)", "3-Layer (State/County/School)")
county_cov_pred[, hierarchy := factor(hierarchy, levels = h_levels_county)]

target_county_rows <- predict_sim$target[loc_id %in% counties & dose == 2 & age > 4]
true_county_df <- data.table(
  loc_id = target_county_rows$loc_id,
  age = target_county_rows$age,
  coverage = latent_params_sim$coverage[target_county_rows$obs_c_id]
)

true_county_faceted <- rbindlist(lapply(h_levels_county, function(h) {
  df <- copy(true_county_df)
  df[, hierarchy := factor(h, levels = h_levels_county)]
  df
}))

ggplot() +
  geom_ribbon(
    data = county_cov_pred,
    aes(x = age, ymin = q2_5, ymax = q97_5, fill = loc_id),
    alpha = 0.15
  ) +
  geom_line(
    data = county_cov_pred,
    aes(x = age, y = q50, color = loc_id),
    linewidth = 0.9
  ) +
  geom_point(
    data = true_county_faceted,
    aes(x = age, y = coverage, color = loc_id, shape = "True Coverage"),
    size = 2
  ) +
  facet_grid(. ~ hierarchy) +
  scale_shape_manual(name = "", values = c("True Coverage" = 18)) +
  theme_bw() +
  theme(legend.position = "bottom") +
  scale_x_continuous(breaks = seq(5, 18, by = 3), minor_breaks = NULL) +
  labs(
    title = "County-Level Coverage Across Data Resolutions",
    x = "Age",
    y = "Coverage Probability",
    color = "County",
    fill = "County"
  )
```

### Force of Vaccination ($\lambda$) Estimation

We also examine the model's ability to estimate the underlying force of vaccination parameters ($\lambda$) across different location data resolutions.

```{r lambda-plot, echo = FALSE, fig.cap = "Force of vaccination (lambda) estimates across data resolutions compared to true values.", fig.width = 7, fig.height = 4}
extract_lambda_summary <- function(fit, label) {
  draws <- rstan::extract(fit$stanfit, pars = "lambda_raw")$lambda_raw
  lambda_draws <- exp(draws)
  doses_factor <- factor(c(
    rep("Dose 1", nrow(lambda_draws)),
    rep("Dose 2", nrow(lambda_draws))
  ))
  data.table(
    hierarchy = label,
    dose = doses_factor,
    lambda = c(lambda_draws[, 1], lambda_draws[, 2])
  )[, .(
    q50 = median(lambda),
    q2_5 = quantile(lambda, 0.025),
    q97_5 = quantile(lambda, 0.975)
  ), by = .(hierarchy, dose)]
}

lambda_est <- rbindlist(list(
  extract_lambda_summary(fit_sim_1layer, "1-Layer"),
  extract_lambda_summary(fit_sim_2layer, "2-Layer"),
  extract_lambda_summary(fit_sim, "3-Layer")
))
lambda_est[, hierarchy := factor(hierarchy, levels = c("1-Layer", "2-Layer", "3-Layer"))]

true_lambda <- data.table(
  dose = factor(c("Dose 1", "Dose 2")),
  true_val = latent_params_sim$lambda
)

ggplot(lambda_est, aes(x = hierarchy, y = q50, color = dose)) +
  geom_hline(
    data = true_lambda,
    aes(yintercept = true_val, color = dose),
    linetype = "dashed",
    linewidth = 0.8
  ) +
  geom_pointrange(
    aes(ymin = q2_5, ymax = q97_5),
    position = position_dodge(width = 0.4),
    size = 0.7
  ) +
  facet_wrap(~dose, scales = "free_y") +
  theme_bw() +
  theme(legend.position = "bottom") +
  labs(
    title = "Force of Vaccination (lambda) Parameter Estimates",
    subtitle = paste(
      "Posterior median & 95% CIs across data resolutions",
      "vs. true parameters (dashed lines)"
    ),
    x = "Data Resolution (Model Hierarchy)",
    y = "Force of Vaccination (lambda)",
    color = "Dose"
  )
```

---

## Conclusion

Support for **flexible location layers** allows `imuGAP` to adapt to whatever location hierarchy you provide. 
- If you provide only state-level aggregated data (**1 layer**), macro trends are estimated accurately.
- If you provide sub-populations like counties (**2 layers**) or sub-sub-populations like schools (**3 layers**), the model automatically builds hierarchical random offsets down the location tree to capture finer resolution variation.
